---
title: "8. go + kegg analysis"
format:
  html:
    embed-resources: true
    toc: true
    toc_float: true
    toc-depth: 6
editor: visual
---

# egg-NOG-mapper

2.1.12

```{bash}
#| eval: false
#| include: true
# gffread -g Locusta_migratoria_v2.genome.fa -y locusta_proteins.fa Locusta_migratoria_v2_fixed.gff3
# 
# gffread -g Harmonia_axyridis.genome.fa -y harmonia_proteins.fa Harmonia_axyridis.gff3
# 
# gffread -g Plutella_xylostella.genome.fa -y plutella_proteins.fa Plutella_xylostella.gff3
# 
# gffread -g Drosophila_melanogaster.genome.fa -y drosophila_proteins.fa Drosophila_melanogaster.gff3
# 
# gffread -g Blattella_germanica.genome.fa -y blattella_proteins.fa Blattella_germanica.gff3
```

| **File** | **Sequences (type)** | **Total length** | **Min. hit e-value** | **Min. hit bit-score** | **Min. % of identity** | **Min. % of query cov.** | **Min. % of subject cov.** | **Tax. scope** | **Orth. restrictions** | **GO evidence** |
|----|----|----|----|----|----|----|----|----|----|----|
| drosophila_proteins.fa | 13955 (proteins) | 7321147 | 0.001 | 60 | 40 | 20 | 20 | auto | all | non-electronic |

| **File** | **Sequences (type)** | **Total length** | **Min. hit e-value** | **Min. hit bit-score** | **Min. % of identity** | **Min. % of query cov.** | **Min. % of subject cov.** | **Tax. scope** | **Orth. restrictions** | **GO evidence** |
|----|----|----|----|----|----|----|----|----|----|----|
| harmonia_proteins.fa | 25685 (proteins) | 9811406 | 0.001 | 60 | 40 | 20 | 20 | auto | all | non-electronic |

| **File** | **Sequences (type)** | **Total length** | **Min. hit e-value** | **Min. hit bit-score** | **Min. % of identity** | **Min. % of query cov.** | **Min. % of subject cov.** | **Tax. scope** | **Orth. restrictions** | **GO evidence** |
|----|----|----|----|----|----|----|----|----|----|----|
| plutella_proteins.fa | 29185 (proteins) | 9739172 | 0.001 | 60 | 40 | 20 | 20 | auto | all | non-electronic |

| **Date created** | **File** | **Sequences (type)** | **Total length** | **Min. hit e-value** | **Min. hit bit-score** | **Min. % of identity** | **Min. % of query cov.** | **Min. % of subject cov.** | **Tax. scope** | **Orth. restrictions** | **GO evidence** |
|----|----|----|----|----|----|----|----|----|----|----|----|
| 09/24/25 | blattella_proteins.fa.gz | 28106 (proteins) | 9442452 | 0.001 | 60 | 40 | 20 | 20 | auto | all | non-electronic |

# libraries

```{r}
library(clusterProfiler)
library(enrichplot)
library(tidyr)
library(DESeq2)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(uwot)
library(readxl)
library(grid)
library(tibble)
library(GSVA)
library(princurve)
library(GO.db)
library(dplyr)
library(stringr)
library(pheatmap)
library(tidytext)
library(topGO)
library(Rgraphviz)
library(UpSetR)
library(forcats)


conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("count", "dplyr")
conflicted::conflict_prefer("unname", "base")
conflicted::conflict_prefer("intersect", "base")
conflicted::conflict_prefer("union", "base")
conflicted::conflict_prefer("setdiff", "base")
conflicted::conflict_prefer("slice", "clusterProfiler")

```

# set-up

```{r}
dds <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/dds.rds")
gene2og <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/gene2og.rds")

background_ogs <- rownames(dds)
length(background_ogs)
head(background_ogs)
length(rownames(gene2og))/5

#without renaming to anything there are errors
gene2og <- gene2og %>%
  dplyr::rename(
    gene_id = "Gene",
    og = "orthogroup"
  )
```

```{r}
stages <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")

all_sig_ogs <- list()

for(i in 1:(length(stages)-1)){
  for(j in (i+1):length(stages)){
    stage1 <- stages[i]
    stage2 <- stages[j]
    
    res <- results(dds, contrast = c("Stage2", stage2, stage1)) 
    
    sig_ogs <- rownames(res[which(res$padj < 0.05 & abs(res$log2FoldChange) > 1), ])
    
    contrast_name <- paste(stage2, "_vs_", stage1, sep="")
    all_sig_ogs[[contrast_name]] <- sig_ogs
  }
}

names(all_sig_ogs)
#lapply(all_sig_ogs, length)
```

# GO & KEGG

## Drosophila

```{r}
eggnog_drosophila <- read.table("/home/cluster_user/Magisterka/drosophila.emapper.annotations.tsv",
                                sep = "\t",
                                header = FALSE,
                                comment.char = "#",
                                quote = "",
                                stringsAsFactors = FALSE)

colnames(eggnog_drosophila) <- c("query", "seed_ortholog", "evalue", "score", "eggNOG_OGs", 
                                 "max_annot_lvl", "COG_category", "Description", "Preferred_name",
                                 "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
                                 "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC",
                                 "CAZy", "BiGG_Reaction", "PFAMs")

#delete .1 from the end of names
eggnog_drosophila$query <- sub("\\.\\d+$", "", eggnog_drosophila$query)

head(eggnog_drosophila)

go_drosophila <- eggnog_drosophila %>%
  select(term = GOs, gene = query) %>%
  filter(!is.na(term) & term != "-") %>%
  separate_rows(term, sep = ",") %>%
  left_join(gene2og, by = c("gene" = "gene_id")) %>%
  filter(!is.na(og)) %>%
  select(term, gene = og) %>%
  distinct()

head(go_drosophila)

kegg_drosophila <- eggnog_drosophila %>%
  select(term = KEGG_Pathway, gene = query) %>%
  filter(!is.na(term) & term != "-") %>%
  separate_rows(term, sep = ",") %>%
  left_join(gene2og, by = c("gene" = "gene_id")) %>%
  filter(!is.na(og)) %>%
  select(term, gene = og) %>%
  distinct()

head(kegg_drosophila)
```

## Harmonia

```{r}
eggnog_harmonia <- read.table("/home/cluster_user/Magisterka/harmonia.emapper.annotations.tsv",
                                sep = "\t",
                                header = FALSE,
                                comment.char = "#",
                                quote = "",
                                stringsAsFactors = FALSE)

colnames(eggnog_harmonia) <- c("query", "seed_ortholog", "evalue", "score", "eggNOG_OGs", 
                                 "max_annot_lvl", "COG_category", "Description", "Preferred_name",
                                 "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
                                 "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC",
                                 "CAZy", "BiGG_Reaction", "PFAMs")

#delete .1 from the end of names
eggnog_harmonia$query <- sub("\\.\\d+$", "", eggnog_harmonia$query)

head(eggnog_harmonia)

go_harmonia <- eggnog_harmonia %>%
  select(term = GOs, gene = query) %>%
  filter(!is.na(term) & term != "-") %>%
  separate_rows(term, sep = ",") %>%
  left_join(gene2og, by = c("gene" = "gene_id")) %>%
  filter(!is.na(og)) %>%
  select(term, gene = og) %>%
  distinct()

head(go_harmonia)

kegg_harmonia <- eggnog_harmonia %>%
  select(term = KEGG_Pathway, gene = query) %>%
  filter(!is.na(term) & term != "-") %>%
  separate_rows(term, sep = ",") %>%
  left_join(gene2og, by = c("gene" = "gene_id")) %>%
  filter(!is.na(og)) %>%
  select(term, gene = og) %>%
  distinct()

head(kegg_harmonia)
```

## Locusta

```{r}
#| eval: false
#| include: true
# head -n 4 /home/cluster_user/Magisterka/locusta1.emapper.annotations.tsv > /home/cluster_user/Magisterka/locusta.emapper.annotations.tsv  
# tail -n +5 /home/cluster_user/Magisterka/locusta1.emapper.annotations.tsv >> /home/cluster_user/Magisterka/locusta.emapper.annotations.tsv
# tail -n +5 /home/cluster_user/Magisterka/locusta2.emapper.annotations.tsv >> /home/cluster_user/Magisterka/locusta.emapper.annotations.tsv
# tail -n +5 /home/cluster_user/Magisterka/locusta3.emapper.annotations.tsv >> /home/cluster_user/Magisterka/locusta.emapper.annotations.tsv
# tail -n +5 /home/cluster_user/Magisterka/locusta4.emapper.annotations.tsv >> /home/cluster_user/Magisterka/locusta.emapper.annotations.tsv
# tail -n +5 /home/cluster_user/Magisterka/locusta5.emapper.annotations.tsv >> /home/cluster_user/Magisterka/locusta.emapper.annotations.tsv
```

```{r}
eggnog_locusta <- read.table("/home/cluster_user/Magisterka/locusta.emapper.annotations.tsv",
                                sep = "\t",
                                header = FALSE,
                                comment.char = "#",
                                quote = "",
                                stringsAsFactors = FALSE)

colnames(eggnog_locusta) <- c("query", "seed_ortholog", "evalue", "score", "eggNOG_OGs", 
                                 "max_annot_lvl", "COG_category", "Description", "Preferred_name",
                                 "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
                                 "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC",
                                 "CAZy", "BiGG_Reaction", "PFAMs")

#delete .t1 .t2 from the end of names
eggnog_locusta$query <- sub("\\.t\\d+$", "", eggnog_locusta$query)

head(eggnog_locusta)

go_locusta <- eggnog_locusta %>%
  select(term = GOs, gene = query) %>%
  filter(!is.na(term) & term != "-") %>%
  separate_rows(term, sep = ",") %>%
  left_join(gene2og, by = c("gene" = "gene_id")) %>%
  filter(!is.na(og)) %>%
  select(term, gene = og) %>%
  distinct()

head(go_locusta)

kegg_locusta <- eggnog_locusta %>%
  select(term = KEGG_Pathway, gene = query) %>%
  filter(!is.na(term) & term != "-") %>%
  separate_rows(term, sep = ",") %>%
  left_join(gene2og, by = c("gene" = "gene_id")) %>%
  filter(!is.na(og)) %>%
  select(term, gene = og) %>%
  distinct()

head(kegg_locusta)
```

## Plutella

```{r}
eggnog_plutella <- read.table("/home/cluster_user/Magisterka/plutella.emapper.annotations.tsv",
                                sep = "\t",
                                header = FALSE,
                                comment.char = "#",
                                quote = "",
                                stringsAsFactors = FALSE)

colnames(eggnog_plutella) <- c("query", "seed_ortholog", "evalue", "score", "eggNOG_OGs", 
                                 "max_annot_lvl", "COG_category", "Description", "Preferred_name",
                                 "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
                                 "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC",
                                 "CAZy", "BiGG_Reaction", "PFAMs")

#delete .1 from the end of names
eggnog_plutella$query <- sub("\\.\\d+$", "", eggnog_plutella$query)

head(eggnog_plutella)

go_plutella <- eggnog_plutella %>%
  select(term = GOs, gene = query) %>%
  filter(!is.na(term) & term != "-") %>%
  separate_rows(term, sep = ",") %>%
  left_join(gene2og, by = c("gene" = "gene_id")) %>%
  filter(!is.na(og)) %>%
  select(term, gene = og) %>%
  distinct()

head(go_plutella)

kegg_plutella <- eggnog_plutella %>%
  select(term = KEGG_Pathway, gene = query) %>%
  filter(!is.na(term) & term != "-") %>%
  separate_rows(term, sep = ",") %>%
  left_join(gene2og, by = c("gene" = "gene_id")) %>%
  filter(!is.na(og)) %>%
  select(term, gene = og) %>%
  distinct()

head(kegg_plutella)
```

## Blattella

```{r}
eggnog_blattella <- read_excel("/home/cluster_user/Magisterka/blattella.emapper.annotations.xlsx", skip =2)


colnames(eggnog_blattella) <- c("query", "seed_ortholog", "evalue", "score", "eggNOG_OGs", 
                                 "max_annot_lvl", "COG_category", "Description", "Preferred_name",
                                 "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
                                 "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC",
                                 "CAZy", "BiGG_Reaction", "PFAMs")

#delete .1 from the end of names
eggnog_blattella$query <- sub("\\.\\d+$", "", eggnog_blattella$query)

head(eggnog_blattella)

go_blattella <- eggnog_blattella %>%
  select(term = GOs, gene = query) %>%
  filter(!is.na(term) & term != "-") %>%
  separate_rows(term, sep = ",") %>%
  left_join(gene2og, by = c("gene" = "gene_id")) %>%
  filter(!is.na(og)) %>%
  select(term, gene = og) %>%
  distinct()

head(go_blattella)

kegg_blattella <- eggnog_blattella %>%
  select(term = KEGG_Pathway, gene = query) %>%
  filter(!is.na(term) & term != "-") %>%
  separate_rows(term, sep = ",") %>%
  left_join(gene2og, by = c("gene" = "gene_id")) %>%
  filter(!is.na(og)) %>%
  select(term, gene = og) %>%
  distinct()

head(kegg_blattella)
```

## t2g

```{r}
###enricher wont work because it needs many GOs with multiple orthogroups (dont have, 821 orthogroups is too little)

# to check: GOslim (from 15k to ...) -> can i get it?
#KEGG pathways (from 700 to over 3k)

#term to gene for all species
t2g_go_common <- bind_rows(go_drosophila, go_locusta, go_harmonia, go_plutella, go_blattella) %>%
   distinct()
# 
 t2g_kegg_common <- bind_rows(kegg_drosophila, kegg_locusta, kegg_harmonia, kegg_plutella, kegg_blattella) %>%
   distinct()

```

### gsea

```{r}
#| eval: false
#| include: false
vst <- varianceStabilizingTransformation(dds, blind = FALSE)
mat <- assay(vst)

#prepare gene sets
go_sets <- split(t2g_go_common$gene, t2g_go_common$term)
kegg_sets <- split(t2g_kegg_common$gene, t2g_kegg_common$term)

length(go_sets)
length(kegg_sets)
sapply(go_sets, length) %>% summary()
sapply(kegg_sets, length) %>% summary()

#check matrixStats version
packageVersion("matrixStats")
options(matrixStats.useNames = FALSE)

#GSVA (Gene Set Variation Analysis) using the ssGSEA method
#based on the vst matrix and orthogroup to term table it counts the activity of genes for each sample separately
#result is GO x samples matrix where the value is the enrichment score based on the ranking
gsva_go <- gsva(
  mat,
  go_sets,
  method = "ssgsea", #single-sample GSEA (gene expression -> rankings-> enrichment scores) -> separate for each sample
  #enrichment scores are normalized between samples by: background value, numbers of genes, enrichment ranges, centering around 0
  #-> so the result is relative to other samples as not an absolute value but a more/less <- from comparing the rankings
  min.sz = 3,
  max.sz = 2000,
  verbose = TRUE
)

gsva_kegg <- gsva(
  mat,
  kegg_sets,
  method = "ssgsea",
  min.sz = 3,
  max.sz = 2000,
  verbose = TRUE
)

dim(gsva_go)
dim(gsva_kegg)
```

## plots - go & kegg

go - pca

```{r}
#| eval: false
#| include: false
#transpose
gsva_go_t <- as.data.frame(t(gsva_go))
gsva_go_t$Sample <- rownames(gsva_go_t)

#metadata
sample_info <- as.data.frame(colData(dds))
sample_info$Sample <- rownames(sample_info)
sample_info <- sample_info %>% select(Sample, Sample2, Stage2, Species)

#merge
gsva_go_t <- left_join(gsva_go_t, sample_info, by = "Sample")

#median (of biological replicates)
gsva_go_median <- gsva_go_t %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

expr_values <- gsva_go_median %>% select(-Sample2, -Stage2, -Species)
exprs_mat <- t(as.matrix(expr_values))
colnames(exprs_mat) <- gsva_go_median$Sample2
rownames(exprs_mat) <- colnames(expr_values)

#pca, variance
pca_go <- prcomp(t(exprs_mat), scale. = TRUE)
percentVar_go <- round(100 * (pca_go$sdev^2 / sum(pca_go$sdev^2)), 1)

#data + metadata
pcaData_go <- as.data.frame(pca_go$x)
pcaData_go$Sample2 <- gsva_go_median$Sample2
pcaData_go$Stage2 <- gsva_go_median$Stage2
pcaData_go$Species <- gsva_go_median$Species

p_go <- ggplot(pcaData_go, aes(PC1, PC2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar_go[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_go[2], "% variance")) +
  ggtitle("PCA of GO enrichment scores") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p_go)

ggsave("/home/cluster_user/Magisterka/plots/PCA_GO_stage_species.png",
       p_go, width = 11, height = 8, dpi = 300)
```

heatmap

```{r}
#| eval: false
#| include: false
#100 most changeable genes
var_idx <- order(apply(gsva_go, 1, var), decreasing = TRUE)[1:100]

species_labels <- c(
  "drosophila" = "D. melanogaster",
  "blattella"  = "B. germanica",
  "plutella"   = "P. xylostella",
  "locusta"    = "L. migratoria",
  "harmonia"   = "H. axyridis"
)

#annotate columns
ann_col <- gsva_go_median %>%
  select(Sample2, Stage2, Species) %>%
  dplyr::rename(Stage = Stage2) %>%
  column_to_rownames("Sample2")

ann_col$Species <- factor(species_labels[ann_col$Species],
                          levels = species_labels)

#colors from ggplot (pca, umap)
stages_order <- c("Adult Female", "Adult Male", "Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa")
stage_colors <- setNames(scales::hue_pal()(length(stages_order)), stages_order)

species_levels <- unique(ann_col$Species)
species_colors <- setNames(RColorBrewer::brewer.pal(n = length(species_levels), "Accent"),
                                 species_levels)

ann_colors <- list(Stage = stage_colors,
                   Species = species_colors)

heat_colors <- colorRampPalette(rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")))(100)
#heatmap
ph <- pheatmap(exprs_mat[var_idx, ],
         annotation_col = ann_col,
         annotation_colors = ann_colors,
         #normalization according to row
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         show_rownames = TRUE,
         show_colnames = TRUE,
         fontsize_col = 8,
         fontsize = 12,
         border_color = NA,
         angle_col = 45,
         fontsize_row = 5,
         main = "Top 100 variable GO terms",
         color = heat_colors
         )

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/heatmap_GO_top100.png", ph, width = 11, height = 8)
```

umap

```{r}
#| eval: false
#| include: false
exprs_mat_t <- t(exprs_mat)

umap_res <- umap(exprs_mat_t, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")

umap_df <- as.data.frame(umap_res)
colnames(umap_df) <- c("UMAP1", "UMAP2")

umap_df$Sample2 <- gsva_go_median$Sample2
umap_df$Stage2 <- gsva_go_median$Stage2
umap_df$Species <- gsva_go_median$Species

p_umap <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 1000) +
  ggtitle("UMAP of GO enrichment scores") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species")

print(p_umap)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_GO_stage_species.png",
       p_umap, width = 11, height = 8, dpi = 300)
```

pseudotime

```{r}
#| eval: false
#| include: false
print(percentVar_go)
print(sum(percentVar_go[0:6]))

#from pca matrix, first 6 main components (74.4% variance explained)
pc_mat_all <- as.matrix(pcaData_go[, paste0("PC", 1:6)])
rownames(pc_mat_all) <- pcaData_go$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData_go$Pseudotime <- lambda_norm_all

stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")

pcaData_go$Stage2 <- factor(pcaData_go$Stage2, levels = stages_order)
pcaData_go$Stage2 <- droplevels(pcaData_go$Stage2)

cor_test <- cor(as.numeric(factor(pcaData_go$Stage2, levels = stages_order)), pcaData_go$Pseudotime, use = "complete.obs")

print(cor_test)
#make sure pseudotime goes from eggs to adults and not from adult to eggs
if (cor_test < 0) {
  pcaData_go$Pseudotime <- 1 - pcaData_go$Pseudotime
}

p1 <- ggplot(pcaData_go, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime across stages (GOs) - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_GO_stage_species.png",
       p1, width = 11, height = 8, dpi = 300)

p2 <- ggplot(pcaData_go, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime distribution (GOs) - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_GO_stage_species.png",
       p2, width = 11, height = 8, dpi = 300)

pseudotime_summary <- pcaData_go %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )

pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory (GOs) - PCA",
    y = "Median pseudotime",
    x = "Developmental stage"
  )

print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_GO_stage_species.png",
       p3, width = 11, height = 8, dpi = 300)
```

pca

```{r}
#| eval: false
#| include: false
curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:6)
#sort point
curve_line <- curve_line[order(pcaData_go$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData_go, aes(x = PC1, y = PC2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar_go[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_go[2], "% variance")) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory on PCA plot (GOs)",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_GO_stage_species.png",
        p_curve, width = 11, height = 8, dpi = 300)
```

umap

```{r}
#| eval: false
#| include: false
umap_mat_all <- as.matrix(umap_df[, c("UMAP1", "UMAP2")])
rownames(umap_mat_all) <- umap_df$Sample2

#match curve on umap
umap_curve_all <- principal_curve(umap_mat_all)

#count pseudotime
lambda_all <- umap_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))
umap_df$Pseudotime <- lambda_norm_all

#check pseudotime direction
stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")
umap_df$Stage2 <- factor(umap_df$Stage2, levels = stages_order)
umap_df$Stage2 <- droplevels(umap_df$Stage2)

cor_test <- cor(as.numeric(factor(umap_df$Stage2, levels = stages_order)),
                umap_df$Pseudotime, use = "complete.obs")
print(cor_test)

#swap pseudotime if needed
if (cor_test < 0) {
  umap_df$Pseudotime <- 1 - umap_df$Pseudotime
}

#plots

#pseudotime vs stage
p1 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime across stages (GOs) - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_umap_GO_stage_species.png", p1, width = 11, height = 8, dpi = 300)

#pseudotime vs stage+species
p2 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime distribution (GOs) - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_umap_GO_stage_species.png", p2, width = 11, height = 8, dpi = 300)

#pseudotime summary line plot
pseudotime_summary <- umap_df %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )
pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory (GOs) - UMAP",
    y = "Median pseudotime",
    x = "Developmental stage"
  )
print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_umap_GO_stage_species.png", p3, width = 11, height = 8, dpi = 300)

#trajectory curve
curve_line <- as.data.frame(umap_curve_all$s)
colnames(curve_line) <- c("UMAP1", "UMAP2")
curve_line <- curve_line[order(umap_df$Pseudotime), ]

#umap plot
p_umap_curve <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(
    data = curve_line,
    aes(x = UMAP1, y = UMAP2),
    inherit.aes = FALSE,
    color = "black",
    size = 1.2,
    alpha = 0.7
  ) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory on UMAP (GOs)",
    x = "UMAP1",
    y = "UMAP2",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_umap_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_GO_stage_species.png",
       p_umap_curve, width = 11, height = 8, dpi = 300)
```

kegg pathways - pca

```{r}
#| eval: false
#| include: false
#transpose
gsva_kegg_t <- as.data.frame(t(gsva_kegg))
gsva_kegg_t$Sample <- rownames(gsva_kegg_t)

#metadata
sample_info <- as.data.frame(colData(dds))
sample_info$Sample <- rownames(sample_info)
sample_info <- sample_info %>% select(Sample, Sample2, Stage2, Species)

#merge
gsva_kegg_t <- left_join(gsva_kegg_t, sample_info, by = "Sample")

#median
gsva_kegg_median <- gsva_kegg_t %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

expr_values <- gsva_kegg_median %>% select(-Sample2, -Stage2, -Species)
exprs_mat <- t(as.matrix(expr_values))
colnames(exprs_mat) <- gsva_kegg_median$Sample2
rownames(exprs_mat) <- colnames(expr_values)

#pca, variance
pca_kegg <- prcomp(t(exprs_mat), scale. = TRUE)
percentVar_kegg <- round(100 * (pca_kegg$sdev^2 / sum(pca_kegg$sdev^2)), 1)

#data + metadata
pcaData_kegg <- as.data.frame(pca_kegg$x)
pcaData_kegg$Sample2 <- gsva_kegg_median$Sample2
pcaData_kegg$Stage2 <- gsva_kegg_median$Stage2
pcaData_kegg$Species <- gsva_kegg_median$Species

p_kegg <- ggplot(pcaData_kegg, aes(PC1, PC2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar_kegg[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_kegg[2], "% variance")) +
  ggtitle("PCA of KEGG enrichment scores") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p_kegg)

ggsave("/home/cluster_user/Magisterka/plots/PCA_KEGG_stage_species.png",
       p_kegg, width = 11, height = 8, dpi = 300)
```

heatmap

```{r}
#| eval: false
#| include: false
#100 most changeable genes
var_idx <- order(apply(gsva_kegg, 1, var), decreasing = TRUE)[1:100]

#annotate columns
ann_col <- gsva_kegg_median %>%
  select(Sample2, Stage2, Species) %>%
  dplyr::rename(Stage = Stage2) %>%
  column_to_rownames("Sample2")

ann_col$Species <- factor(species_labels[ann_col$Species],
                          levels = species_labels)

#colors from ggplot (pca, umap)
stages_order <- c("Adult Female", "Adult Male", "Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa")
stage_colors <- setNames(scales::hue_pal()(length(stages_order)), stages_order)


species_levels <- unique(ann_col$Species)
species_colors <- setNames(RColorBrewer::brewer.pal(n = length(species_levels), "Accent"),
                                 species_levels)

ann_colors <- list(Stage = stage_colors,
                   Species = species_colors)
#heatmap
ph <- pheatmap(exprs_mat[var_idx, ],
         annotation_col = ann_col,
         annotation_colors = ann_colors,
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         show_rownames = TRUE,
         show_colnames = TRUE,
         fontsize_col = 8,
         fontsize = 12,
         border_color = NA,
         angle_col = 45,
         fontsize_row = 5,
         main = "Top 100 variable KEGG terms",
         color = heat_colors
         )

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/heatmap_kegg_top100.png", ph, width = 11, height = 8)
```

umap

```{r}
#| eval: false
#| include: false
exprs_mat_t <- t(exprs_mat)

umap_res <- umap(exprs_mat_t, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")

umap_df <- as.data.frame(umap_res)
colnames(umap_df) <- c("UMAP1", "UMAP2")

umap_df$Sample2 <- gsva_kegg_median$Sample2
umap_df$Stage2 <- gsva_kegg_median$Stage2
umap_df$Species <- gsva_kegg_median$Species

p_umap <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 1000) +
  ggtitle("UMAP of KEGG enrichment scores") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species")

print(p_umap)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_kegg_stage_species.png",
       p_umap, width = 11, height = 8, dpi = 300)
```

pseudotime

```{r}
#| eval: false
#| include: false
print(percentVar_kegg)
print(sum(percentVar_kegg[0:6]))

#from pca matrix, first 6 main components (73.1% variance explained)
pc_mat_all <- as.matrix(pcaData_kegg[, paste0("PC", 1:6)])
rownames(pc_mat_all) <- pcaData_kegg$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData_kegg$Pseudotime <- lambda_norm_all

stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")

pcaData_kegg$Stage2 <- factor(pcaData_kegg$Stage2, levels = stages_order)
pcaData_kegg$Stage2 <- droplevels(pcaData_kegg$Stage2)

cor_test <- cor(as.numeric(factor(pcaData_kegg$Stage2, levels = stages_order)), pcaData_kegg$Pseudotime, use = "complete.obs")

print(cor_test)
#make sure pseudotime goes from eggs to adults and not from adult to eggs
if (cor_test < 0) {
  pcaData_kegg$Pseudotime <- 1 - pcaData_kegg$Pseudotime
}

p1 <- ggplot(pcaData_kegg, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime across stages (KEGG pathways) - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_kegg_stage_species.png",
       p1, width = 11, height = 8, dpi = 300)

p2 <- ggplot(pcaData_kegg, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime distribution (KEGG pathways) - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_kegg_stage_species.png",
       p2, width = 11, height = 8, dpi = 300)

pseudotime_summary <- pcaData_kegg %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )

pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory (KEGG pathways) - PCA",
    y = "Median pseudotime",
    x = "Developmental stage"
  )

print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_kegg_stage_species.png",
       p3, width = 11, height = 8, dpi = 300)
```

pca

```{r}
#| eval: false
#| include: false
curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:6)
#sort point
curve_line <- curve_line[order(pcaData_kegg$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData_kegg, aes(x = PC1, y = PC2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar_kegg[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_kegg[2], "% variance")) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory on PCA plot (KEGG pathways)",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_kegg_stage_species.png",
        p_curve, width = 11, height = 8, dpi = 300)
```

umap

```{r}
#| eval: false
#| include: false
umap_mat_all <- as.matrix(umap_df[, c("UMAP1", "UMAP2")])
rownames(umap_mat_all) <- umap_df$Sample2

#match curve on umap
umap_curve_all <- principal_curve(umap_mat_all)

#count pseudotime
lambda_all <- umap_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))
umap_df$Pseudotime <- lambda_norm_all

#check pseudotime direction
stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")
umap_df$Stage2 <- factor(umap_df$Stage2, levels = stages_order)
umap_df$Stage2 <- droplevels(umap_df$Stage2)

cor_test <- cor(as.numeric(factor(umap_df$Stage2, levels = stages_order)),
                umap_df$Pseudotime, use = "complete.obs")
print(cor_test)

#swap pseudotime if needed
if (cor_test < 0) {
  umap_df$Pseudotime <- 1 - umap_df$Pseudotime
}

#plots

#pseudotime vs stage
p1 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime across stages (KEGG pathways) - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_umap_KEGG_stage_species.png", p1, width = 11, height = 8, dpi = 300)

#pseudotime vs stage+species
p2 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime distribution (KEGG pathways) - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_umap_KEGG_stage_species.png", p2, width = 11, height = 8, dpi = 300)

#pseudotime summary line plot
pseudotime_summary <- umap_df %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )
pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory (KEGG pathways) - UMAP",
    y = "Median pseudotime",
    x = "Developmental stage"
  )
print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_umap_KEGG_stage_species.png", p3, width = 11, height = 8, dpi = 300)

#trajectory curve
curve_line <- as.data.frame(umap_curve_all$s)
colnames(curve_line) <- c("UMAP1", "UMAP2")
curve_line <- curve_line[order(umap_df$Pseudotime), ]

#umap plot
p_umap_curve <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(
    data = curve_line,
    aes(x = UMAP1, y = UMAP2),
    inherit.aes = FALSE,
    color = "black",
    size = 1.2,
    alpha = 0.7
  ) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory on UMAP (KEGG pathways)",
    x = "UMAP1",
    y = "UMAP2",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_umap_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_KEGG_stage_species.png",
       p_umap_curve, width = 11, height = 8, dpi = 300)
```

## no species effect

go - pca

```{r}
#| eval: false
#| include: false
species <- colData(dds)[colnames(gsva_go), "Species"]

#delete species effect by linear function
residuals_go <- apply(gsva_go, 1, function(x) {
  lm(x ~ species)$residuals
})
residuals_go <- t(residuals_go)
colnames(residuals_go) <- colnames(gsva_go)
rownames(residuals_go) <- rownames(gsva_go)

#median
resid_df <- as.data.frame(t(residuals_go))
resid_df$Sample <- rownames(resid_df)

sample_info <- data.frame(colData(dds))
sample_info$Sample <- rownames(sample_info)
sample_info <- sample_info %>% select(Sample, Sample2, Stage2, Species)

resid_df <- dplyr::left_join(resid_df, sample_info, by = "Sample")

resid_median <- resid_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

expr_median <- resid_median %>% select(-Stage2, -Species)
samples <- resid_median$Sample2
expr_values <- resid_median %>% select(-Sample2, -Stage2, -Species)

exprs_mat <- t(as.matrix(expr_values))
colnames(exprs_mat) <- resid_median$Sample2
rownames(exprs_mat) <- colnames(expr_values)

#pca, variance
pca_go <- prcomp(t(exprs_mat), scale. = TRUE)
percentVar_go <- round(100 * (pca_go$sdev^2 / sum(pca_go$sdev^2)), 1)

#data + metadata
pcaData_go <- as.data.frame(pca_go$x)
pcaData_go$Sample2 <- resid_median$Sample2
pcaData_go$Stage2 <- resid_median$Stage2
pcaData_go$Species <- resid_median$Species

p_go <- ggplot(pcaData_go, aes(PC1, PC2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar_go[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_go[2], "% variance")) +
  ggtitle("PCA of GO enrichment scores (removed species effect)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p_go)

ggsave("/home/cluster_user/Magisterka/plots/PCA_GO_stage.png",
       p_go, width = 11, height = 8, dpi = 300)
```

heatmap

```{r}
#| eval: false
#| include: false
#100 most changeable genes
var_idx <- order(apply(exprs_mat, 1, var), decreasing = TRUE)[1:100]

#annotate columns
ann_col <- data.frame(
  Stage = resid_median$Stage2,
  Species = resid_median$Species
)
rownames(ann_col) <- colnames(exprs_mat)

ann_col$Species <- factor(species_labels[ann_col$Species],
                          levels = species_labels)

#colors from ggplot (pca, umap)
stages_order <- c("Adult Female", "Adult Male", "Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa")
stage_colors <- setNames(scales::hue_pal()(length(stages_order)), stages_order)


species_levels <- unique(ann_col$Species)
species_colors_named <- setNames(RColorBrewer::brewer.pal(n = length(species_levels), "Accent"),
                                 species_levels)
ann_colors <- list(Stage = stage_colors,
                   Species = species_colors_named)

#heatmap
ph <- pheatmap(exprs_mat[var_idx, ],
         annotation_col = ann_col,
         annotation_colors = ann_colors,
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         show_rownames = TRUE,
         show_colnames = TRUE,
         fontsize_col = 8,
         fontsize = 12,
         border_color = NA,
         angle_col = 45,
         fontsize_row = 5,
         main = "Top 100 variable GO terms(removed species effect)",
         color = heat_colors
         )

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/heatmap_GO_top100_stage.png", ph, width = 11, height = 8)
```

umap

```{r}
#| eval: false
#| include: false
exprs_mat_t <- t(exprs_mat)

umap_res <- umap(exprs_mat_t, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")

umap_df <- as.data.frame(umap_res)
colnames(umap_df) <- c("UMAP1", "UMAP2")

umap_df$Sample2 <- resid_median$Sample2
umap_df$Stage2 <- resid_median$Stage2
umap_df$Species <- resid_median$Species

p_umap <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 1000) +
  ggtitle("UMAP of GO enrichment scores (removed species effect)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species")

print(p_umap)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_GO_stage.png",
       p_umap, width = 11, height = 8, dpi = 300)
```

pseudotime

```{r}
#| eval: false
#| include: false
print(percentVar_go)
print(sum(percentVar_go[0:6]))

#from pca matrix, first 6 main components (71.9% variance explained)
pc_mat_all <- as.matrix(pcaData_go[, paste0("PC", 1:6)])
rownames(pc_mat_all) <- pcaData_go$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData_go$Pseudotime <- lambda_norm_all

stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")

pcaData_go$Stage2 <- factor(pcaData_go$Stage2, levels = stages_order)
pcaData_go$Stage2 <- droplevels(pcaData_go$Stage2)

cor_test <- cor(as.numeric(factor(pcaData_go$Stage2, levels = stages_order)), pcaData_go$Pseudotime, use = "complete.obs")

print(cor_test)
#make sure pseudotime goes from eggs to adults and not from adult to eggs
if (cor_test < 0) {
  pcaData_go$Pseudotime <- 1 - pcaData_go$Pseudotime
}

p1 <- ggplot(pcaData_go, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime across stages (GOs, removed species effect) - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_GO_stage.png",
       p1, width = 11, height = 8, dpi = 300)

p2 <- ggplot(pcaData_go, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime distribution (GOs, removed species effect) - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_GO_stage.png",
       p2, width = 11, height = 8, dpi = 300)

pseudotime_summary <- pcaData_go %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )

pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory (GOs, removed species effect) - PCA",
    y = "Median pseudotime",
    x = "Developmental stage"
  )

print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_GO_stage.png",
       p3, width = 11, height = 8, dpi = 300)
```

pca

```{r}
#| eval: false
#| include: false
curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:6)
#sort point
curve_line <- curve_line[order(pcaData_go$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData_go, aes(x = PC1, y = PC2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar_go[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_go[2], "% variance")) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory on PCA plot (GOs, removed species effect)",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_GO_stage.png",
        p_curve, width = 11, height = 8, dpi = 300)
```

umap

```{r}
#| eval: false
#| include: false
umap_mat_all <- as.matrix(umap_df[, c("UMAP1", "UMAP2")])
rownames(umap_mat_all) <- umap_df$Sample2

#match curve on umap
umap_curve_all <- principal_curve(umap_mat_all)

#count pseudotime
lambda_all <- umap_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))
umap_df$Pseudotime <- lambda_norm_all

#check pseudotime direction
stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")
umap_df$Stage2 <- factor(umap_df$Stage2, levels = stages_order)
umap_df$Stage2 <- droplevels(umap_df$Stage2)

cor_test <- cor(as.numeric(factor(umap_df$Stage2, levels = stages_order)),
                umap_df$Pseudotime, use = "complete.obs")
print(cor_test)

#swap pseudotime if needed
if (cor_test < 0) {
  umap_df$Pseudotime <- 1 - umap_df$Pseudotime
}

#plots

#pseudotime vs stage
p1 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime across stages (GOs, removed species effect) - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_umap_GO_stage.png", p1, width = 11, height = 8, dpi = 300)

#pseudotime vs stage+species
p2 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime distribution (GOs, removed species effect) - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_umap_GO_stage.png", p2, width = 11, height = 8, dpi = 300)

#pseudotime summary line plot
pseudotime_summary <- umap_df %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )
pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory (GOs, removed species effect) - UMAP",
    y = "Median pseudotime",
    x = "Developmental stage"
  )
print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_umap_GO_stage.png", p3, width = 11, height = 8, dpi = 300)

#trajectory curve
curve_line <- as.data.frame(umap_curve_all$s)
colnames(curve_line) <- c("UMAP1", "UMAP2")
curve_line <- curve_line[order(umap_df$Pseudotime), ]

#umap plot
p_umap_curve <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(
    data = curve_line,
    aes(x = UMAP1, y = UMAP2),
    inherit.aes = FALSE,
    color = "black",
    size = 1.2,
    alpha = 0.7
  ) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory on UMAP (GOs, removed species effect)",
    x = "UMAP1",
    y = "UMAP2",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_umap_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_GO_stage.png",
       p_umap_curve, width = 11, height = 8, dpi = 300)
```

kegg pathway - pca

```{r}
#| eval: false
#| include: false
species <- colData(dds)[colnames(gsva_kegg), "Species"]

#delete species effect by linear function
residuals_KEGG <- apply(gsva_kegg, 1, function(x) {
  lm(x ~ species)$residuals
})
residuals_KEGG <- t(residuals_KEGG)
colnames(residuals_KEGG) <- colnames(gsva_kegg)
rownames(residuals_KEGG) <- rownames(gsva_kegg)

#median
resid_df <- as.data.frame(t(residuals_KEGG))
resid_df$Sample <- rownames(resid_df)

sample_info <- data.frame(colData(dds))
sample_info$Sample <- rownames(sample_info)
sample_info <- sample_info %>% select(Sample, Sample2, Stage2, Species)

resid_df <- dplyr::left_join(resid_df, sample_info, by = "Sample")

resid_median <- resid_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

expr_median <- resid_median %>% select(-Stage2, -Species)
samples <- resid_median$Sample2
expr_values <- resid_median %>% select(-Sample2, -Stage2, -Species)

exprs_mat <- t(as.matrix(expr_values))
colnames(exprs_mat) <- resid_median$Sample2
rownames(exprs_mat) <- colnames(expr_values)

#pca, variance
pca_KEGG <- prcomp(t(exprs_mat), scale. = TRUE)
percentVar_KEGG <- round(100 * (pca_KEGG$sdev^2 / sum(pca_KEGG$sdev^2)), 1)

#data + metadata
pcaData_KEGG <- as.data.frame(pca_KEGG$x)
pcaData_KEGG$Sample2 <- resid_median$Sample2
pcaData_KEGG$Stage2 <- resid_median$Stage2
pcaData_KEGG$Species <- resid_median$Species

p_KEGG <- ggplot(pcaData_KEGG, aes(PC1, PC2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar_KEGG[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_KEGG[2], "% variance")) +
  ggtitle("PCA of KEGG enrichment scores (removed species effect)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p_KEGG)

ggsave("/home/cluster_user/Magisterka/plots/PCA_KEGG_stage.png",
       p_KEGG, width = 11, height = 8, dpi = 300)
```

heatmap

```{r}
#| eval: false
#| include: false
#100 most changeable genes
var_idx <- order(apply(exprs_mat, 1, var), decreasing = TRUE)[1:100]

#annotate columns
ann_col <- data.frame(
  Stage = resid_median$Stage2,
  Species = resid_median$Species
)
rownames(ann_col) <- colnames(exprs_mat)

ann_col$Species <- factor(species_labels[ann_col$Species],
                          levels = species_labels)

#colors from ggplot (pca, umap)
stages_order <- c("Adult Female", "Adult Male", "Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa")
stage_colors <- setNames(scales::hue_pal()(length(stages_order)), stages_order)


species_levels <- unique(ann_col$Species)
species_colors_named <- setNames(RColorBrewer::brewer.pal(n = length(species_levels), "Accent"),
                                 species_levels)
ann_colors <- list(Stage = stage_colors,
                   Species = species_colors_named)

#heatmap
ph <- pheatmap(exprs_mat[var_idx, ],
         annotation_col = ann_col,
         annotation_colors = ann_colors,
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         show_rownames = TRUE,
         show_colnames = TRUE,
         fontsize_col = 8,
         fontsize = 12,
         border_color = NA,
         angle_col = 45,
         fontsize_row = 5,
         main = "  Top 100 variable KEGG terms (removed species effect)",
         color = heat_colors
         )

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/heatmap_KEGG_top100_stage.png", ph, width = 11, height = 8)
```

umap

```{r}
#| eval: false
#| include: false
exprs_mat_t <- t(exprs_mat)

umap_res <- umap(exprs_mat_t, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")

umap_df <- as.data.frame(umap_res)
colnames(umap_df) <- c("UMAP1", "UMAP2")

umap_df$Sample2 <- resid_median$Sample2
umap_df$Stage2 <- resid_median$Stage2
umap_df$Species <- resid_median$Species

p_umap <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 1000) +
  ggtitle("UMAP of KEGG enrichment scores (removed species effect)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species")

print(p_umap)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_KEGG_stage.png",
       p_umap, width = 11, height = 8, dpi = 300)
```

pseudotime

```{r}
#| eval: false
#| include: false
print(percentVar_KEGG)
print(sum(percentVar_KEGG[0:6]))

#from pca matrix, first 6 main components (70.9% variance explained)
pc_mat_all <- as.matrix(pcaData_KEGG[, paste0("PC", 1:6)])
rownames(pc_mat_all) <- pcaData_KEGG$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData_KEGG$Pseudotime <- lambda_norm_all

stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")

pcaData_KEGG$Stage2 <- factor(pcaData_KEGG$Stage2, levels = stages_order)
pcaData_KEGG$Stage2 <- droplevels(pcaData_KEGG$Stage2)

cor_test <- cor(as.numeric(factor(pcaData_KEGG$Stage2, levels = stages_order)), pcaData_KEGG$Pseudotime, use = "complete.obs")

print(cor_test)
#make sure pseudotime goes from eggs to adults and not from adult to eggs
if (cor_test < 0) {
  pcaData_KEGG$Pseudotime <- 1 - pcaData_KEGG$Pseudotime
}

p1 <- ggplot(pcaData_KEGG, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime across stages (KEGG pathways, removed species effect) - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_kegg_stage.png",
       p1, width = 11, height = 8, dpi = 300)

p2 <- ggplot(pcaData_KEGG, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime distribution (KEGG pathways, removed species effect) - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_kegg_stage.png",
       p2, width = 11, height = 8, dpi = 300)

pseudotime_summary <- pcaData_KEGG %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )

pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory (KEGG pathways, removed species effect) - PCA",
    y = "Median pseudotime",
    x = "Developmental stage"
  )

print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_kegg_stage.png",
       p3, width = 11, height = 8, dpi = 300)
```

pca

```{r}
#| eval: false
#| include: false
curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:6)
#sort point
curve_line <- curve_line[order(pcaData_KEGG$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData_KEGG, aes(x = PC1, y = PC2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar_KEGG[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_KEGG[2], "% variance")) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory on PCA plot (KEGG pathways, removed species effect)",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_kegg_stage.png",
        p_curve, width = 11, height = 8, dpi = 300)
```

umap

```{r}
#| eval: false
#| include: false
umap_mat_all <- as.matrix(umap_df[, c("UMAP1", "UMAP2")])
rownames(umap_mat_all) <- umap_df$Sample2

#match curve on umap
umap_curve_all <- principal_curve(umap_mat_all)

#count pseudotime
lambda_all <- umap_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))
umap_df$Pseudotime <- lambda_norm_all

#check pseudotime direction
stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")
umap_df$Stage2 <- factor(umap_df$Stage2, levels = stages_order)
umap_df$Stage2 <- droplevels(umap_df$Stage2)

cor_test <- cor(as.numeric(factor(umap_df$Stage2, levels = stages_order)),
                umap_df$Pseudotime, use = "complete.obs")
print(cor_test)

#swap pseudotime if needed
if (cor_test < 0) {
  umap_df$Pseudotime <- 1 - umap_df$Pseudotime
}

#plots

#pseudotime vs stage
p1 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime across stages (KEGG pathways, removed species effect) - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_umap_KEGG_stage.png", p1, width = 11, height = 8, dpi = 300)

#pseudotime vs stage+species
p2 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime distribution (KEGG pathways, removed species effect) - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_umap_KEGG_stage.png", p2, width = 11, height = 8, dpi = 300)

#pseudotime summary line plot
pseudotime_summary <- umap_df %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )
pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Accent"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory (KEGG pathways, removed species effect) - UMAP",
    y = "Median pseudotime",
    x = "Developmental stage"
  )
print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_umap_KEGG_stage.png", p3, width = 11, height = 8, dpi = 300)

#trajectory curve
curve_line <- as.data.frame(umap_curve_all$s)
colnames(curve_line) <- c("UMAP1", "UMAP2")
curve_line <- curve_line[order(umap_df$Pseudotime), ]

#umap plot
p_umap_curve <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(
    data = curve_line,
    aes(x = UMAP1, y = UMAP2),
    inherit.aes = FALSE,
    color = "black",
    size = 1.2,
    alpha = 0.7
  ) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pesudotime trajectory on UMAP (KEGG pathways, removed species effect)",
    x = "UMAP1",
    y = "UMAP2",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_umap_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_KEGG_stage.png",
       p_umap_curve, width = 11, height = 8, dpi = 300)
```

# correlation to stages - impactful genes/GOs

```{r}
dds <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/dds_batch.rds")

vst <- varianceStabilizingTransformation(dds, blind = FALSE)
```

## functions - correlation and enrichement and plots

```{r}
run_stage_correlation <- function(expr_mat,
                                   stage_vector,
                                   species_vector,
                                   method = "spearman",
                                   aggregate = "mean") {

  # checks
  stopifnot(
    ncol(expr_mat) == length(stage_vector),
    length(stage_vector) == length(species_vector)
  )

  # per-gene correlation within species
  cor_list <- apply(expr_mat, 1, function(gene_expr) {

    df <- data.frame(
      expr   = gene_expr,
      stage  = stage_vector,
      species = species_vector
    )

    # compute correlation per species
    per_species <- lapply(split(df, df$species), function(d) {

      if (length(unique(d$stage)) < 2) return(c(rho = NA, p = NA))

      ct <- suppressWarnings(
        cor.test(d$expr, d$stage, method = method)
      )

      c(rho = unname(ct$estimate), p = ct$p.value)
    })

    per_species <- do.call(rbind, per_species)

    # aggregate across species
    rho_vals <- per_species[, "rho"]

    rho_agg <- switch(
      aggregate,
      mean = mean(rho_vals, na.rm = TRUE),
      median = median(rho_vals, na.rm = TRUE)
    )

    # combine p-values (Fisher method)
    p_vals <- per_species[, "p"]

    p_vals <- p_vals[!is.na(p_vals)]

    if (length(p_vals) > 0) {
      chi <- -2 * sum(log(p_vals))
      p_agg <- pchisq(chi, df = 2 * length(p_vals), lower.tail = FALSE)
    } else {
      p_agg <- NA
    }

    c(rho = rho_agg, p = p_agg)
  })

  # format output
  cor_df <- as.data.frame(t(cor_list))
  cor_df$gene_id <- rownames(cor_df)

  cor_df$padj <- p.adjust(cor_df$p, method = "BH")

  return(cor_df)
}

classify_trend_genes <- function(cor_df, rho_cutoff = 0.5, padj_cutoff = 0.05) {
  #classify genes showing strong early or late developmental trends
  
  cor_df %>%
    mutate(
      trend = case_when(
        rho > rho_cutoff & padj < padj_cutoff ~ "Late",
        rho < -rho_cutoff & padj < padj_cutoff ~ "Early",
        TRUE ~ "Other"
      )
    )
}

run_gsea <- function(cor_df, term2gene,
                                minGSSize = 10,
                                maxGSSize = 500,
                                pvalueCutoff = 0.05) {
  #Run gene set enrichment analysis using ranked gene–stage correlations
  
  #create ranked gene list using correlation coefficient
  gene_list <- cor_df %>%
    arrange(desc(rho)) %>%
    select(gene_id, rho)
  
  #convert to named numeric vector required by GSEA
  gene_list_vec <- setNames(gene_list$rho, gene_list$gene_id)
  gene_list_vec <- sort(gene_list_vec, decreasing = TRUE)
  
  #run GSEA
  gsea_res <- GSEA(
    geneList = gene_list_vec,
    TERM2GENE = term2gene,
    minGSSize = minGSSize,
    maxGSSize = maxGSSize,
    pvalueCutoff = pvalueCutoff,
    verbose = TRUE
  )
  
  #convert result to dataframe and classify enrichment direction
  as.data.frame(gsea_res) %>%
    dplyr::mutate(direction = ifelse(NES > 0, "Late-associated", "Early-associated"))
}

annotate_go_terms <- function(gsea_df) {
  #Add GO term names and ontology categories to GSEA results
  
  #helper function returning GO term name
  get_term <- function(go_id) {
    term <- GO.db::GOTERM[[go_id]]
    if (is.null(term)) NA_character_ else Term(term)
  }
  
  #helper function returning GO ontology (BP/MF/CC)
  get_ontology <- function(go_id) {
    term <- GO.db::GOTERM[[go_id]]
    if (is.null(term)) NA_character_ else Ontology(term)
  }
  
  #annotate GO IDs with term name and ontology
  gsea_df %>%
    mutate(
      Term = vapply(ID, get_term, character(1)),
      Ontology = vapply(ID, get_ontology, character(1))
    )
}

plot_gene_trends <- function(expr_matrix,
                             gene_cor,
                             sample_info,
                             dataset_label = "nymph",
                             rho_cutoff = 0.5,
                             bins = 50) {
  # 1. Ensure trend factor
  gene_cor <- gene_cor %>%
    mutate(trend = case_when(
      rho >  rho_cutoff ~ "Late",
      rho < -rho_cutoff ~ "Early",
      TRUE ~ "Other"
    )) %>%
    mutate(trend = factor(trend, levels = c("Other", "Early", "Late")))

  # 2. Correlation distribution plot
  p_hist <- ggplot(gene_cor, aes(rho, fill = trend)) +
    geom_histogram(bins = bins, position = "stack", color = "white") +
    geom_vline(
      xintercept = c(-rho_cutoff, 0, rho_cutoff),
      linetype = c("dotted", "dashed", "dotted"),
      color = "grey40"
    ) +
    scale_fill_manual(values = c(
      Early = "#2C7BB6",
      Late = "#D7191C",
      Other = "grey85"
    )) +
    theme_classic() +
    labs(
      x = "Spearman correlation (ρ)",
      y = "Gene count",
      fill = "Trend",
      title = paste0(
        "Distribution of gene–development correlations\nin ", dataset_label
      )
    )

  # 3. Long format expression data
  expr_long <- expr_matrix %>%
    as.data.frame() %>%
    rownames_to_column("gene_id") %>%
    pivot_longer(
      -gene_id,
      names_to = "Sample",
      values_to = "expression"
    ) %>%
    left_join(gene_cor %>% select(gene_id, rho), by = "gene_id") %>%
    left_join(sample_info %>% select(Sample, instar_scaled), by = "Sample") %>%
    mutate(
      trend = case_when(
        rho >  rho_cutoff ~ "Late",
        rho < -rho_cutoff ~ "Early",
        TRUE ~ "Other"
      )
    )

  # 4. Trend dynamics plot
  p_trend <- expr_long %>%
    filter(trend != "Other") %>%
    ggplot(aes(instar_scaled, expression, color = trend, fill = trend)) +
    geom_smooth(
      method = "loess",
      se = FALSE,
      linewidth = 1.5
    ) +
    stat_summary(
      fun.data = function(x) {
        data.frame(
          y = median(x),
          ymin = quantile(x, 0.25),
          ymax = quantile(x, 0.75)
        )
      },
      geom = "ribbon",
      alpha = 0.2,
      color = NA
    ) +
    theme_classic(base_size = 14) +
    scale_color_manual(values = c(
      "Early" = "#4575B4",
      "Late"  = "#D73027"
    )) +
    scale_fill_manual(values = c(
      "Early" = "#4575B4",
      "Late"  = "#D73027"
    )) +
    labs(
      x = "Developmental progression (scaled)",
      y = "Expression (VST)",
      title = paste0(
        "Opposing expression dynamics of early and late genes\nin ",
        dataset_label
      )
    )

  # Return both plots
  return(list(
    correlation_plot = p_hist,
    trend_plot = p_trend
  ))
}

plot_gene_trends_holo <- function(expr_matrix,
                                  gene_cor,
                                  sample_info,
                                  dataset_label = "holometabolous",
                                  rho_cutoff = 0.5,
                                  bins = 50) {
  # 1. Build trend classification
  gene_cor <- gene_cor %>%
    mutate(trend = case_when(
      rho >  rho_cutoff ~ "Late",
      rho < -rho_cutoff ~ "Early",
      TRUE ~ "Other"
    )) %>%
    mutate(trend = factor(trend, levels = c("Other", "Early", "Late")))

  # 2. Correlation histogram
  p_hist <- ggplot(gene_cor, aes(rho, fill = trend)) +
    geom_histogram(bins = bins, position = "stack", color = "white") +
    geom_vline(
      xintercept = c(-rho_cutoff, 0, rho_cutoff),
      linetype = c("dotted", "dashed", "dotted"),
      color = "grey40"
    ) +
    scale_fill_manual(values = c(
      Early = "#2C7BB6",
      Late = "#D7191C",
      Other = "grey85"
    )) +
    theme_classic() +
    labs(
      x = "Spearman correlation (ρ)",
      y = "Gene count",
      fill = "Trend",
      title = paste0(
        "Distribution of gene–development correlations\nin ", dataset_label
      )
    )

  # 3. Long expression table
  expr_long <- expr_matrix %>%
    as.data.frame() %>%
    rownames_to_column("gene_id") %>%
    pivot_longer(
      -gene_id,
      names_to = "Sample",
      values_to = "expression"
    ) %>%
    left_join(gene_cor %>% select(gene_id, rho), by = "gene_id") %>%
    left_join(
      sample_info %>%
        select(Sample, dev_scaled),   # <<< FIXED HERE
      by = "Sample"
    ) %>%
    mutate(
      trend = case_when(
        rho >  rho_cutoff ~ "Late",
        rho < -rho_cutoff ~ "Early",
        TRUE ~ "Other"
      )
    )

  # 4. Expression dynamics plot
  p_trend <- expr_long %>%
    filter(trend != "Other") %>%
    ggplot(aes(dev_scaled, expression, color = trend, fill = trend)) +
    geom_smooth(
      method = "loess",
      se = FALSE,
      linewidth = 1.5
    ) +
    stat_summary(
      fun.data = function(x) {
        data.frame(
          y = median(x),
          ymin = quantile(x, 0.25),
          ymax = quantile(x, 0.75)
        )
      },
      geom = "ribbon",
      alpha = 0.2,
      color = NA
    ) +
    theme_classic(base_size = 14) +
    scale_color_manual(values = c(
      "Early" = "#4575B4",
      "Late"  = "#D73027"
    )) +
    scale_fill_manual(values = c(
      "Early" = "#4575B4",
      "Late"  = "#D73027"
    )) +
    labs(
      x = "Developmental progression (scaled)",
      y = "Expression (VST)",
      title = paste0(
        "Opposing expression dynamics of early and late genes\nin ",
        dataset_label
      )
    )

  return(list(
    correlation_plot = p_hist,
    trend_plot = p_trend
  ))
}
```

## nymph

```{r}
# Expression matrix
expr <- assay(vst)

# Metadata
sample_info <- as.data.frame(colData(vst))
sample_info$Sample <- rownames(sample_info)

sample_info <- sample_info %>%
  mutate(
    instar_number = case_when(
      Stage2 == "Nymph" & grepl("\\d+", Stage) ~
        as.numeric(gsub(".*?(\\d+).*", "\\1", Stage)),
      TRUE ~ NA_real_
    )
  )

#Nymph samples with species-specific ordering + normalization
nymph_samples <- sample_info %>%
  filter(Stage2 == "Nymph", !is.na(instar_number)) %>%
  group_by(Species) %>%
  mutate(
    instar_rank   = dense_rank(instar_number),
    instar_scaled = instar_rank / max(instar_rank)
  ) %>%
  ungroup()

expr_nymph <- expr[, nymph_samples$Sample]
instar     <- nymph_samples$instar_scaled


#Correlation
gene_cor <- run_stage_correlation(expr_nymph, instar, nymph_samples$Species)
gene_cor <- classify_trend_genes(gene_cor)

plots <- plot_gene_trends(
  expr_matrix = expr_nymph,
  gene_cor = gene_cor,
  sample_info = nymph_samples,
  dataset_label = "hemimetabolous nymphs"
)

plots$correlation_plot
plots$trend_plot

```

### gsea

```{r}
#GSEA
gsea_go_df <- run_gsea(gene_cor, t2g_go_common)
```

```{r}
gsea_go_annotated <- annotate_go_terms(gsea_go_df)

gsea_plot_df <- gsea_go_annotated %>%
  filter(
    p.adjust < 0.05,
    !is.na(Term),
    !is.na(NES)
  ) %>%
  group_by(direction) %>%
  slice_max(abs(NES), n = 20) %>%
  ungroup() %>%
  mutate(
    Term = stringr::str_wrap(Term, width = 40),
    direction = factor(direction,
      levels = c("Early-associated", "Late-associated")
    )
  )

p <- ggplot(
  gsea_plot_df,
  aes(
    x = NES,
    y = tidytext::reorder_within(Term, NES, direction),
    size = -log10(p.adjust)
  )
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
  geom_point(alpha = 0.8, color = "black") +
  tidytext::scale_y_reordered() +
  facet_wrap(~direction, scales = "free") +
  scale_size(range = c(2, 6)) +
  theme_classic(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey95", color = NA),
    axis.text.y = element_text(size = 8),
    panel.spacing = unit(1, "cm")
  ) +
  labs(
    x = "Normalized Enrichment Score (NES)",
    y = "GO term",
    size = "-log10(FDR)",
    title = "Functional shifts during development\nnymphs"
  )

print(p)

ggsave(
  "/home/cluster_user/Magisterka/plots/nymph_gsea.png",
  p,
  width = 10,
  height = 10,
  dpi = 300
)

gsea_plot_df_bp <- gsea_go_annotated %>%
  filter(
    Ontology == "BP",
    p.adjust < 0.05,
    !is.na(Term),
    !is.na(NES)
  ) %>%
  group_by(direction) %>%
  slice_max(abs(NES), n = 20) %>%
  ungroup() %>%
  mutate(
    Term = stringr::str_wrap(Term, width = 40),
    direction = factor(direction,
      levels = c("Early-associated", "Late-associated")
    )
  )

p_bp <- ggplot(
  gsea_plot_df_bp,
  aes(
    x = NES,
    y = reorder_within(Term, NES, direction),
    size = -log10(p.adjust)
  )
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
  geom_point(alpha = 0.8, color = "black") +
  tidytext::scale_y_reordered() +
  facet_wrap(~direction, scales = "free") +
  scale_size(range = c(2, 6)) +
  theme_classic(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey95", color = NA),
    axis.text.y = element_text(size = 8),
    panel.spacing = unit(1, "cm")
  ) +
  labs(
    x = "Normalized Enrichment Score (NES)",
    y = "GO biological process",
    size = "-log10(FDR)",
    title = "Development-associated biological processes"
  )

print(p_bp)

ggsave(
  "/home/cluster_user/Magisterka/plots/nymph_gsea_bp.png",
  p_bp,
  width = 10,
  height = 10,
  dpi = 300
)
```

```{r}
# Extract core enrichment genes per GO
gsea_go_genes <- gsea_go_annotated %>%
  select(ID, core_enrichment) %>%
  mutate(
    gene_list = strsplit(core_enrichment, "/")
  )

# Compute pairwise Jaccard similarity
jaccard_matrix <- outer(
  gsea_go_genes$gene_list,
  gsea_go_genes$gene_list,
  Vectorize(function(x, y) length(intersect(x, y)) / length(union(x, y)))
)

# Hierarchical clustering
hc <- hclust(as.dist(1 - jaccard_matrix))
plot(hc)

# Cut into clusters (e.g., 0.7 similarity threshold)
clusters <- cutree(hc, h = 0.3)

gsea_go_annotated$cluster <- clusters

# Pick one representative per cluster (highest NES)
gsea_collapsed <- gsea_go_annotated %>%
  group_by(cluster) %>%
  slice_max(NES, n = 1) %>%
  ungroup()
```

## holometabolous

```{r}
# Expression matrix
expr <- assay(vst)

# Metadata
sample_info <- as.data.frame(colData(vst))
sample_info$Sample <- rownames(sample_info)

sample_info <- sample_info %>% mutate(
  larva_inst = case_when(
    Stage2 == "Larva" & grepl("\\d+", Stage) ~
      as.numeric(gsub(".*?(\\d+).*", "\\1", Stage)),
    TRUE ~ NA_real_
  ),
  
  prepupa_rank = case_when(
    Stage2 == "Prepupa" & grepl("\\d+", Sample2) ~
      as.numeric(gsub(".*?(\\d+).*", "\\1", Sample2)),
    TRUE ~ NA_real_
  ),
  
  pupa_rank = case_when(
    Stage2 == "Pupa" & grepl("\\d+", Stage) ~
      as.numeric(gsub(".*?(\\d+).*", "\\1", Stage)),
    TRUE ~ NA_real_
  )
)

sample_info <- sample_info %>%
  group_by(Species) %>%
  mutate(
    max_larva   = max(larva_inst, na.rm = TRUE),
    max_prepupa = max(prepupa_rank, na.rm = TRUE),
    
    dev_order = case_when(
      Stage2 == "Larva"   ~ larva_inst,
      Stage2 == "Prepupa" ~ prepupa_rank + max_larva,
      Stage2 == "Pupa"    ~ pupa_rank + max_larva + max_prepupa,
      TRUE ~ NA_real_
    )
  ) %>%
  ungroup()

#normalize within species
sample_info <- sample_info %>%
  group_by(Species) %>%
  mutate(dev_scaled = dev_order / max(dev_order, na.rm = TRUE)) %>%
  ungroup()


# Keep only holometabolic samples
holo_samples <- sample_info %>%
  filter(Stage2 %in% c("Larva", "Prepupa", "Pupa"),
         !is.na(dev_scaled))

expr_holo <- expr[, holo_samples$Sample]
dev_order <- holo_samples$dev_scaled


#Correlation
gene_cor_holo <- run_stage_correlation(expr_holo, dev_order, holo_samples$Species)
gene_cor_holo <- classify_trend_genes(gene_cor_holo)


plots <- plot_gene_trends_holo(
  expr_matrix = expr_holo,
  gene_cor = gene_cor,
  sample_info = holo_samples,
  dataset_label = "holometabolous"
)

plots$correlation_plot
plots$trend_plot
```

### gsea

```{r}
#GSEA
gsea_go_holo_df <- run_gsea(gene_cor_holo, t2g_go_common)
```

```{r}
# Annotate GSEA results
gsea_go_holo_annotated <- annotate_go_terms(gsea_go_holo_df)

# Full GO plot (all ontologies)
gsea_plot_df <- gsea_go_holo_annotated %>%
  filter(
    p.adjust < 0.05,
    !is.na(Term),
    !is.na(NES)
  ) %>%
  group_by(direction) %>%
  slice_max(abs(NES), n = 20) %>%
  ungroup() %>%
  mutate(
    #Term = stringr::str_wrap(Term, width = 40),
    direction = factor(direction,
      levels = c("Early-associated", "Late-associated")
    )
  )


p_holo <- ggplot(gsea_plot_df, aes(x = NES, y = reorder_within(Term, NES, direction),
      size = -log10(p.adjust)
    )) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
  geom_point(alpha = 0.8, color = "black") +
  tidytext::scale_y_reordered() +
  facet_wrap(~direction, scales = "free") +
  scale_size(range = c(2, 6)) +
  theme_classic(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey95", color = NA),
    axis.text.y = element_text(size = 8),
    panel.spacing = unit(1, "cm")
  ) +
  labs(
    x = "Normalized Enrichment Score (NES)",
    y = "GO term",
    size = "-log10(FDR)",
    title = "Functional context of genes driving holometabolic developmental trends"
  )

print(p_holo)

ggsave(
  "/home/cluster_user/Magisterka/plots/holo_gsea.png",
  p_holo,
  width = 12,
  height = 10,
  dpi = 300
)

# Biological Processes only
gsea_plot_holo_bp <- gsea_go_holo_annotated %>%
  filter(
    Ontology == "BP",
    p.adjust < 0.05,
    !is.na(Term),
    !is.na(NES)
  ) %>%
  group_by(direction) %>%
  slice_max(abs(NES), n = 20) %>%
  ungroup() %>%
  mutate(
    Term = stringr::str_wrap(Term, width = 40),
    direction = factor(direction,
      levels = c("Early-associated", "Late-associated")
    )
  )

p_holo_bp <- ggplot(
  gsea_plot_holo_bp,
  aes(
    x = NES,
    y = reorder_within(Term, NES, direction),
    size = -log10(p.adjust)
  )
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
  geom_point(alpha = 0.8, color = "black") +
  tidytext::scale_y_reordered() +
  facet_wrap(~direction, scales = "free") +
  scale_size(range = c(2, 6)) +
  theme_classic(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey95", color = NA),
    axis.text.y = element_text(size = 8),
    panel.spacing = unit(1, "cm")
  ) +  
  labs(
    x = "Normalized Enrichment Score (NES)",
    y = "GO biological process",
    size = "-log10(FDR)",
    title = "Biological processes associated with gene expression trends\nacross holometabolic development"
  )

print(p_holo_bp)

ggsave(
  "/home/cluster_user/Magisterka/plots/holo_gsea_bp.png",
  p_holo_bp,
  width = 12,
  height = 10,
  dpi = 300
)
```

### common GOs

```{r}
sig_nymph <- gsea_go_annotated %>%
  select(ID, Term, direction, NES)

sig_holo <- gsea_go_holo_annotated %>%
  select(ID, Term, direction, NES)

common_go <- intersect(sig_nymph$ID, sig_holo$ID)
length(common_go)

common_df <- sig_nymph %>%
  filter(ID %in% common_go) %>%
  left_join(sig_holo %>% select(ID, direction, NES), by = "ID", suffix = c("_nymph", "_holo"))

head(common_df)

```

# DEG - separate for species

based on Stage2 (no locusta)

## deseq for stage vs all other stages

```{r}
run_stage_vs_all_deseq <- function(counts,
                                   metadata,
                                   stage_column = "Stage2",
                                   min_count = 10,
                                   padj_cutoff = 0.05,
                                   lfc_cutoff = 0) {
  #creating DESeqDataSet
  dds <- DESeqDataSetFromMatrix(
    countData = counts,
    colData = metadata,
    design = as.formula(paste0("~ ", stage_column))
  )
  
  #low reads filtering
  keep <- rowSums(counts(dds)) >= min_count
  dds <- dds[keep, ]
  
  #run DESeq
  dds <- DESeq(dds)
  
  stages <- levels(colData(dds)[[stage_column]])
  deg_list <- list()
  
  #check contrast for each stage vs all other stages
  for (st in stages) {
    
    dds_tmp <- dds
    
    dds_tmp$Stage_vs_All <- ifelse(
      colData(dds_tmp)[[stage_column]] == st,
      st,
      "other"
    )
    
    dds_tmp$Stage_vs_All <- factor(dds_tmp$Stage_vs_All)
    design(dds_tmp) <- ~ Stage_vs_All
    
    dds_tmp <- DESeq(dds_tmp, quiet = TRUE)
    
    res <- results(dds_tmp,
                   contrast = c("Stage_vs_All", st, "other"))
    
    deg_list[[paste0(st, "_vs_all")]] <-
      subset(res,
             padj < padj_cutoff &
             log2FoldChange > lfc_cutoff)
  }
  
  return(list(
    dds = dds,
    deg_stage2 = deg_list
  ))
}
```

## enrichement - GO

```{r}
run_stage_go_enrichment <- function(deg_list, eggnog_table) {
  # build GO - gene mapping
  go_species <- eggnog_table %>%
    select(term = GOs, gene = query) %>%
    filter(!is.na(term) & term != "-") %>%
    separate_rows(term, sep = ",") %>%
    distinct()

  gene2GO <- split(go_species$term, go_species$gene)

  background_genes <- unique(go_species$gene)

  # FINAL OUTPUT: store full topGO objects
  stage_topGO_objects <- list()

  for (st in names(deg_list)) {

    deg_genes <- rownames(deg_list[[st]])
    deg_genes <- deg_genes[deg_genes %in% background_genes]

    cat(st, ":", length(deg_genes), "genes used for enrichment\n")

    # binary gene list
    geneList <- factor(as.integer(background_genes %in% deg_genes))
    names(geneList) <- background_genes

    # topGO object
    GOdata <- new("topGOdata",
                  ontology = "BP",
                  allGenes = geneList,
                  annot = annFUN.gene2GO,
                  gene2GO = gene2GO,
                  nodeSize = 10)

    # enrichment test
    result <- runTest(GOdata,
                      algorithm = "weight01",
                      statistic = "fisher")

    # store EVERYTHING needed for plotting graph
    stage_topGO_objects[[st]] <- list(
      GOdata = GOdata,
      result = result,
      geneList = geneList
    )
  }

  return(stage_topGO_objects)
}
```

## visualisation

```{r}
#maps each term to an ancestor within a level range (range worked better than a single level -> max_level if finds more)
# level is approximated by walking up the GO graph through parents
# 1 = root (GO:0008150) -> Biological Processes

plot_stage_go_results <- function(stage_go_results,
                                  deg_list,
                                  top_n = 30,
                                  min_level = 3,
                                  max_level = 4,
                                  output_dir = NULL,
                                  prefix = "GO") {

  # map short species id to labels
  species_labels <- c(
    "blattella" = "B. germanica",
    "drosophila" = "D. melanogaster",
    "harmonia" = "H. axyridis",
    "locusta" = "L. migratoria",
    "plutella" = "P. xylostella",
    "drosophila2" = "D. melanogaster (set 2)"
  )

  #fallback to prefix if not found
  species_name <- species_labels[[prefix]]
  if (is.null(species_name)) species_name <- prefix

  plots <- list()

  # extract top_n most significant terms (by Fisher test) per stage
  # from topGO GenTable output
  extract_top_terms <- function(stage_name, obj) {
    # skip invalid or empty objects
    if (is.null(obj) || is.null(obj$GOdata) || is.null(obj$result)) return(NULL)
    # extract GO table
    gt <- tryCatch({
      GenTable(
        obj$GOdata,
        Fisher = obj$result,
        orderBy = "Fisher",
        topNodes = top_n
      )
    }, error = function(e) NULL)
    # NULL if no results
    if (is.null(gt) || nrow(gt) == 0) return(NULL)
    # dataframe with stage, GO ID, and transformed score
    data.frame(
      Stage = gsub("_vs_all", "", stage_name),
      GO = gt$GO.ID,
      score = -log10(pmax(as.numeric(gt$Fisher), 1e-300))
    )
  }

  # extraction across all stages and combine results
  top_terms <- mapply(
    FUN = extract_top_terms,
    stage_name = names(stage_go_results),
    obj = stage_go_results,
    SIMPLIFY = FALSE
  ) %>% bind_rows()

  # stop if no GO terms found
  if (nrow(top_terms) == 0) {
    warning("No GO terms found")
    return(plots)
  }

  # convert GO ID to term name
  safe_term <- function(go) {
    if (is.na(go)) return(NA)
    term <- tryCatch(Term(GOTERM[[go]]), error = function(e) NA)
    ifelse(is.na(term), go, term)
  }

  # iteratively traverse parent relationships -> estimate GO in set levels
  # find depth in the ontology tree (it can take a lot of time)
  get_go_level_fast <- function(go) {
    if (go == "GO:0008150") return(1)
    level <- 1
    current <- go

    repeat {
      parents <- GOBPPARENTS[[current]]
      if (is.null(parents) || all(is.na(parents))) break
      parents <- parents[!is.na(parents)]
      if (length(parents) == 0) break
      current <- parents[1]
      level <- level + 1
      # Safety break to avoid loooong loops
      if (level > 100) break
    }
    return(level)
  }

  # map GO term to ancestor within the level range (closest to max_level if more)
  map_to_level <- function(go, min_level, max_level) {
    ancestors <- c(go, unlist(GOBPANCESTOR[[go]]))
    ancestors <- unique(ancestors[grepl("^GO:", ancestors)])
    
    if (length(ancestors) == 0) return(NA)
    
    levels <- sapply(ancestors, get_go_level_fast)
    valid <- which(levels >= min_level & levels <= max_level)
    if (length(valid) == 0) return(NA)

    return(ancestors[valid][which.max(levels[valid])])
  }

  # get term names
  top_terms$MappedGO <- sapply(top_terms$GO, map_to_level,
                               min_level = min_level,
                               max_level = max_level)

  top_terms$MappedName <- sapply(top_terms$MappedGO, safe_term)

  # collapse redundant GOs in a stage <- keep max score per mapped term
  collapsed_df <- top_terms %>%
    filter(!is.na(MappedName)) %>%
    group_by(Stage, MappedName) %>%
    summarise(score = max(score), .groups = "drop")

  # Select top_n global most significant GOs
  global_terms <- collapsed_df %>%
    group_by(MappedName) %>%
    summarise(score = max(score), .groups = "drop") %>%
    slice_max(order_by = score, n = top_n) %>%
    pull(MappedName)

  # get scores for selected GOs for all stages
  lookup_scores <- function(stage_name, obj, go_terms) {

    res <- tryCatch(score(obj$result), error = function(e) NULL)
    if (is.null(res)) return(NULL)

    data.frame(
      Stage = stage_name,
      GO = go_terms,
      score = -log10(pmax(res[go_terms], 1e-300))
    )
  }

  # score search for each stage
  full_scores <- lapply(names(stage_go_results), function(stage) {
    obj <- stage_go_results[[stage]]

    selected_go <- top_terms %>%
      filter(MappedName %in% global_terms) %>%
      pull(GO) %>%
      unique()

    lookup_scores(gsub("_vs_all", "", stage), obj, selected_go)
  }) %>% bind_rows()

  # Map GO IDs again to ensure consistency
  full_scores$MappedGO <- sapply(full_scores$GO, map_to_level,
                                min_level = min_level,
                                max_level = max_level)

  full_scores$MappedName <- sapply(full_scores$MappedGO, safe_term)

  # heatmap matrix -> fill missing with zero
  heatmap_df <- full_scores %>%
    filter(MappedName %in% global_terms) %>%
    group_by(Stage, MappedName) %>%
    summarise(score = max(score, na.rm = TRUE), .groups = "drop") %>%
    tidyr::complete(Stage, MappedName, fill = list(score = 0))

  # heatmap of GO enrichment
  p_heatmap <- ggplot(
    heatmap_df,
    aes(
      x = Stage,
      y = reorder(str_wrap(MappedName, 40), score),
      fill = score
    )
  ) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "#E66101") +
    theme_minimal() +
    labs(
      title = paste("GO Biological Process –", species_name),
      x = "Stage",
      y = "",
      fill = "-log10(p)"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text.y = element_text(size = 7),
      panel.grid = element_blank()
    ) +
    scale_x_discrete(expand = c(0, 0))

  print(p_heatmap)
  plots$heatmap <- p_heatmap

  # save plot
  if (!is.null(output_dir)) {
    ggsave(file.path(output_dir,
                     paste0(prefix, "_GO_heatmap.png")),
           p_heatmap, width = 8, height = 6)
  }
  return(plots)
}
```

## run

```{r}
run_stage_analysis <- function(counts,
                               metadata,
                               eggnog_table,
                               stage_column = "Stage2") {
  # 1. DESEQ ANALYSIS
  res <- run_stage_vs_all_deseq(
    counts = counts,
    metadata = metadata,
    stage_column = stage_column
  )

  # extract stage-specific DEGs
  deg_stage2 <- res$deg_stage2

  # optionally keep dds (commented out to save memory)
  # dds <- res$dds

  # print number of DEGs per stage
  cat("\nDEGs per stage:\n")
  print(sapply(deg_stage2, nrow))

  # 2. OVERLAP MATRIX
  stage_gene_lists <- lapply(deg_stage2, rownames)

  overlap_matrix <- outer(
    names(stage_gene_lists),
    names(stage_gene_lists),
    Vectorize(function(x, y)
      length(intersect(stage_gene_lists[[x]], stage_gene_lists[[y]])))
  )

  cat("\nOverlap matrix:\n")
  print(overlap_matrix)

  # 3. GO ENRICHMENT
  go_results <- run_stage_go_enrichment(
    deg_list = deg_stage2,
    eggnog_table = eggnog_table
  )

  return(list(
    deg_stage2 = deg_stage2,
    go_results = go_results
  ))
}
```

## harmonia

```{r}
# load count matrix and metadata
harmonia_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/harmonia_counts.rds")
harmonia_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/harmonia_metadata.rds")

# load full metadata table
metadata <- read.csv(file="/home/cluster_user/Magisterka/metadata2.csv",
                     row.names = 1,
                     header = TRUE,
                     sep = ",")

# reorder metadata rows to match count matrix columns
harmonia_metadata <- metadata[colnames(harmonia_counts), ]

harmonia <- run_stage_analysis(
  counts = harmonia_counts,
  metadata = harmonia_metadata,
  eggnog_table = eggnog_harmonia
)

print(harmonia$go_results[[1]]$result)
print(harmonia$go_results[[2]]$result)
print(harmonia$go_results[[3]]$result)

pdf("GO_graph_harmonia.pdf")
showSigOfNodes(
  harmonia$go_results[[1]]$GOdata,
  score(harmonia$go_results[[1]]$result),
  firstSigNodes = 5,
  useInfo = "def"
)
dev.off()
```

plot

```{r}
plot_stage_go_results(
  stage_go_results = harmonia$go_results,
  deg_list = harmonia$deg_stage2,
  top_n = 30,
  min_level = 3,
  max_level = 4,
  output_dir = "/home/cluster_user/Magisterka/plots/",
  prefix = "harmonia"
)
```

## plutella

```{r}
plutella_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/plutella_counts.rds")
plutella_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/plutella_metadata.rds")
metadata <- read.csv(file="/home/cluster_user/Magisterka/metadata2.csv", row.names = 1, header = TRUE, sep = ",") 

plutella_metadata <- metadata[colnames(plutella_counts), ]

plutella <- run_stage_analysis(
  counts = plutella_counts,
  metadata = plutella_metadata,
  eggnog_table = eggnog_plutella
)
```

```{r}
plot_stage_go_results(
  stage_go_results = plutella$go_results,
  deg_list = plutella$deg_stage2,
  top_n = 30,
  min_level = 3,
  max_level = 4,
  output_dir = "/home/cluster_user/Magisterka/plots/",
  prefix = "plutella"
)
```

## 1 drosophila

```{r}
drosophila_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila_counts.rds")
drosophila_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila_metadata.rds")
metadata <- read.csv(file="/home/cluster_user/Magisterka/metadata2.csv", row.names = 1, header = TRUE, sep = ",") 

drosophila_metadata <- metadata[colnames(drosophila_counts), ]

drosophila <- run_stage_analysis(
  counts = drosophila_counts,
  metadata = drosophila_metadata,
  eggnog_table = eggnog_drosophila
)
```

```{r}
plot_stage_go_results(
  stage_go_results = drosophila$go_results,
  deg_list = drosophila$deg_stage2,
  top_n = 30,
  min_level = 3,
  max_level = 4,
  output_dir = "/home/cluster_user/Magisterka/plots/",
  prefix = "drosophila"
)
```

## 2 drosophila

```{r}
drosophila2_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila2_counts.rds")
drosophila2_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila2_metadata.rds")
metadata <- read.csv(file="/home/cluster_user/Magisterka/metadata2.csv", row.names = 1, header = TRUE, sep = ",") 

drosophila2_metadata <- metadata[colnames(drosophila2_counts), ]

drosophila2 <- run_stage_analysis(
  counts = drosophila2_counts,
  metadata = drosophila2_metadata,
  eggnog_table = eggnog_drosophila
)
```

```{r}
plot_stage_go_results(
  stage_go_results = drosophila2$go_results,
  deg_list = drosophila2$deg_stage2,
  top_n = 30,
  min_level = 3,
  max_level = 4,
  output_dir = "/home/cluster_user/Magisterka/plots/",
  prefix = "drosophila2"
)
```

## blattella

```{r}
blattella_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/blattella_counts.rds")
blattella_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/blattella_metadata.rds")
metadata <- read.csv(file="/home/cluster_user/Magisterka/metadata2.csv", row.names = 1, header = TRUE, sep = ",") 

blattella_metadata <- metadata[colnames(blattella_counts), ]

blattella <- run_stage_analysis(
  counts = blattella_counts,
  metadata = blattella_metadata,
  eggnog_table = eggnog_blattella
)
```

```{r}
plot_stage_go_results(
  stage_go_results = blattella$go_results,
  deg_list = blattella$deg_stage2,
  top_n = 30,
  min_level = 3,
  max_level = 4,
  output_dir = "/home/cluster_user/Magisterka/plots/",
  prefix = "blattella"
)

## this hemimetabolic embryo has a lot of similar processes to larvae
## nymphs has cuticle which was in pupa
```

## compare

```{r}
# function converting gene IDs to orthogroup IDs for a given species
genes_to_og <- function(genes, gene2og, species_name) {
  gene2og %>%
    filter(Species == species_name,
           gene_id %in% genes) %>%
    pull(og) %>%
    unique()
}


# convert stage-specific genes to orthogroups (Harmonia)
harmonia_stage_og <- lapply(
  harmonia$deg_stage2,
  function(df) genes_to_og(rownames(df), gene2og, "harmonia")
)

# rename stages and remove "_vs_all"
names(harmonia_stage_og) <- paste0("Harmonia_", gsub("_vs_all", "", names(harmonia_stage_og)))


# convert stage-specific genes to orthogroups (Plutella)
plutella_stage_og <- lapply(
  plutella$deg_stage2,
  function(df) genes_to_og(rownames(df), gene2og, "plutella")
)

names(plutella_stage_og) <- paste0("Plutella_", gsub("_vs_all", "", names(plutella_stage_og)))


# convert stage-specific genes to orthogroups (Drosophila dataset 1)
drosophila_stage_og <- lapply(
  drosophila$deg_stage2,
  function(df) genes_to_og(rownames(df), gene2og, "drosophila")
)

names(drosophila_stage_og) <- paste0("Drosophila_", gsub("_vs_all", "", names(drosophila_stage_og)))


# convert stage-specific genes to orthogroups (Drosophila dataset 2)
drosophila2_stage_og <- lapply(
  drosophila2$deg_stage2,
  function(df) genes_to_og(rownames(df), gene2og, "drosophila")
)

names(drosophila2_stage_og) <- paste0("Drosophila2_", gsub("_vs_all", "", names(drosophila2_stage_og)))


# convert stage-specific genes to orthogroups (Blattella)
blattella_stage_og <- lapply(
  blattella$deg_stage2,
  function(df) genes_to_og(rownames(df), gene2og, "blattella")
)

names(blattella_stage_og) <- paste0("Blattella_", gsub("_vs_all", "", names(blattella_stage_og)))


# combine orthogroup sets from all species
stage_gene_sets <- c(
  harmonia_stage_og,
  plutella_stage_og,
  drosophila_stage_og,
  drosophila2_stage_og,
  blattella_stage_og
)

# ensure all orthogroups exist in gene2og table
stage_gene_sets <- lapply(stage_gene_sets, function(x) intersect(x, gene2og$og))

# check number of orthogroups per stage
sapply(stage_gene_sets, length)
```

## Jaccard similarity

```{r}
#shared genes / total unique genes across both stages
#-> mitigates bias from comparing groups with many DEGs vs comparing groups with less DEGs
#they may have similar overlap but it's different ratio
#because it's not all genes (orthogroups) but only DEGs orthogroups

# function calculating Jaccard similarity between two sets
jaccard <- function(a, b) {
  length(intersect(a, b)) / length(union(a, b))
}

sets <- names(stage_gene_sets)

# initialize similarity matrix
mat <- matrix(0, length(sets), length(sets))
rownames(mat) <- sets
colnames(mat) <- sets


# compute pairwise Jaccard similarity between stages
for(i in seq_along(sets)){
  for(j in seq_along(sets)){
    mat[i,j] <- jaccard(
      stage_gene_sets[[sets[i]]],
      stage_gene_sets[[sets[j]]]
    )
  }
}

# remove self-comparisons
diag(mat) <- NA

# define color scale for heatmap
breaks <- seq(0, 0.4, length.out = 100)

colors <- colorRampPalette(
  c("white", "#FDB863", "#E66101", "#B2182B")
)(99)


# plot heatmap of stage overlap
pheatmap(
  mat,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  color = colors,
  breaks = breaks,
  angle_col = 45,
  border_color = NA,
  na_col = "grey90",
  main = "Jaccard similarity of stage-specific DEGs"
)


# save heatmap to file
png("/home/cluster_user/Magisterka/plots/DEG_overlap_heatmap.png",
    width = 2200,
    height = 1800,
    res = 300)

pheatmap(
  mat,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  color = colors,
  breaks = breaks,
  angle_col = 45,
  border_color = NA,
  na_col = "grey90",
  main = "Jaccard similarity of stage-specific DEGs",
  fontsize_row = 7,
  fontsize_col = 7
)

dev.off()
```
