---
title: "6. all species gene expression analysis"
format:
  html:
    embed-resources: true
    toc: true
    toc_float: true
    toc-depth: 6
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## libs

```{r}
#| echo: true
#| eval: true
#| results: false
library(DESeq2)
library(ggplot2)
library(tidyr)
#library(umap)
library(uwot) 
library(pheatmap)
library(ggrepel)
library(readr)
library(stringr)
library(dplyr)
library(tibble)
library(Biostrings)
library(Mfuzz)
library(Biobase)
library(limma)
library(grid)
library(princurve)
library(patchwork)
library(sva)
library(vegan)
library(viridis)
library(lme4)
library(rlang)
library(emmeans)

conflicted::conflict_prefer("rename", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("intersect", "dplyr")
conflicted::conflict_prefer("unname", "base")
conflicted::conflict_prefer("umap", "uwot")
conflicted::conflict_prefer("exprs", "Biobase")
conflicted::conflict_prefer("setdiff", "dplyr")
```

## matrix

read previously prepared counts and orthogroups files

```{r}
locusta_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/locusta_counts.rds")
harmonia_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/harmonia_counts.rds")
plutella_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/plutella_counts.rds")
drosophila_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila_counts.rds")
blattella_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/blattella_counts.rds")
drosophila2_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila2_counts.rds")
```

```{r}
orthogroups <- read_tsv("/home/cluster_user/Magisterka/proteins/OrthoFinder/Results_Jun25/Orthogroups/Orthogroups.tsv")

single_copy <- read_lines("/home/cluster_user/Magisterka/proteins/OrthoFinder/Results_Jun25/Orthogroups/Orthogroups_SingleCopyOrthologues.txt")

#information about genes in orthogroups
head(orthogroups)

#single-copy orthogroups
head(single_copy)
```

### orthofinder

making matrix with SRR IDs as columns and orthogroups as rows, so for each gene with counts I sum them into orthogroup

```{r}
#| echo: true
#| eval: true
#| results: false
#list of dataframes with counts (per RNA-seq dataset)
all_counts <- list(
  locusta = locusta_counts,
  harmonia = harmonia_counts,
  plutella = plutella_counts,
  drosophila = drosophila_counts,
  blattella = blattella_counts,
  drosophila2 = drosophila2_counts
)

#combine all counts tables into one long table
#rows: all gene names; columns: all SRR and species; values: counts 
all_counts_long <- lapply(names(all_counts), function(species) {
  df <- all_counts[[species]]
  df$gene_id <- rownames(df)
  df$species <- species
  df
}) %>%
  bind_rows()

head(all_counts_long)

```

```{r}
#| echo: true
#| eval: true
#| results: false
#table: orthogroup, species, gene
orthogroup_mapping <- orthogroups %>%
  pivot_longer(-Orthogroup, names_to = "species", values_to = "genes") %>%
  filter(!is.na(genes)) %>%
  separate_rows(genes, sep = ",") %>%
  #clean gene names
  mutate(
    genes = str_trim(genes),
    species = str_replace(species, "_proteins", ""),
    genes = str_replace(genes, "\\..*", "") 
  )

#duplicate mapping for the second drosophila dataset
orthogroup_mapping_drosophila2 <- orthogroup_mapping %>%
  filter(species == "drosophila") %>%
  mutate(species = "drosophila2")

orthogroup_mapping <- bind_rows(orthogroup_mapping, orthogroup_mapping_drosophila2)

orthogroup_mapping %>%
  dplyr::count(species)

#numbers of genes in orthogroups (based on genomes)
# blattella	22002			
# drosophila	11435			
# drosophila2	11435			
# harmonia	18095			
# locusta	90809			
# plutella	23332			

head(orthogroup_mapping)


#join count data with orthogroup mapping
merged <- all_counts_long %>%
  inner_join(
    orthogroup_mapping,
    by = c("species", "gene_id" = "genes")
  )

head(merged)
merged %>%
  dplyr::count(species) %>%
  print()

#numbers of genes after mapping counts to orthogroups
#(should be equal or less than in the original genome annotation)
# blattella	19505			
# drosophila	11435			
# drosophila2	11394			
# harmonia	14846			
# locusta	65328			
# plutella	23143
```

### orthosnap

```{r}
#orthosnap -> found additional 1:1 orthogroups

snap_files <- list.files("/home/cluster_user/Magisterka/proteins/OrthoFinder/Results_Jun25/output_snap_ogs/", pattern = "\\.fa$", full.names = TRUE)

#read files and extract gene IDs
snap_orthologs <- lapply(snap_files, function(file) {
  fasta <- readAAStringSet(file)
  genes <- names(fasta)
  
  tibble(
    orthogroup = str_remove(basename(file), "\\.fa.orthosnap.0.fa$"),
    gene_id = str_replace(genes, "^.*?_", "") %>%
          str_replace("\\..*", "")
  )
}) %>% bind_rows()

#merge with expression counts
snap_expr <- snap_orthologs %>%
  inner_join(all_counts_long, by = "gene_id")

#sum gene counts into orthogroups
orthosnap_counts <- snap_expr %>%
  select(-gene_id, -species) %>%
  group_by(orthogroup) %>%
  summarise(across(everything(), ~sum(.x, na.rm = TRUE))) %>%
  column_to_rownames(var = "orthogroup")

#rows = orthogroups, columns = SRR samples
orthosnap_counts %>% dim()

#numbers of SRR samples per species
srr_per_species <- sapply(all_counts, ncol)
print(srr_per_species)

#total number of SRR samples across all datadets
total_srr <- sum(srr_per_species)
print(total_srr)

```

### 1:1 orthogroups

```{r}
#| echo: true
#| eval: true
#| results: false
#gene expression counts for all orthogroups (not only 1:1)
orthogroup_counts <- merged %>%
  select(Orthogroup, where(is.numeric)) %>%
  group_by(Orthogroup) %>%
  summarise(across(everything(), ~sum(.x, na.rm = TRUE)))

head(orthogroup_counts)

orthogroup_counts <- orthogroup_counts %>%
  column_to_rownames(var = "Orthogroup")

head(orthogroup_counts)
dim(orthogroup_counts)

colnames(orthogroup_counts)
```

```{r}
#narrow to only 1:1 orthogroups
orthogroup_counts_1to1 <- orthogroup_counts %>%
  filter(rownames(.) %in% single_copy)

#check if anything overlaps between 1:1 from orthofinder and orthosnap (no)
overlapping_genes <- intersect(rownames(orthogroup_counts_1to1), rownames(orthosnap_counts))

length(overlapping_genes)
print(overlapping_genes)

#add orthosnap
combined_counts <- bind_rows(
  orthogroup_counts_1to1,
  orthosnap_counts[!rownames(orthosnap_counts) %in% rownames(orthogroup_counts_1to1), ]
)

combined_counts[is.na(combined_counts)] <- 0
```

```{r}
#number of 1:1 orthogroups
cat("orthogroup_counts_1to1: ", dim(orthogroup_counts_1to1)[1], "orthogroups ×", dim(orthogroup_counts_1to1)[2], "SRRs\n")
cat("orthosnap_counts: ", dim(orthosnap_counts)[1], "orthogroups ×", dim(orthosnap_counts)[2], "SRRs\n")
cat("combined_counts: ", dim(combined_counts)[1], "orthogroups ×", dim(combined_counts)[2], "SRRs\n")

```

## deseq2

```{r}
#read metadata: Stage, Replicate, Sample.name, Species, Stage2, Sample2
#_2 are generalized names across all datasets / across replicates
metadata <- read.csv(file="/home/cluster_user/Magisterka/metadata2.csv", row.names = 1, header = TRUE, sep = ",") 

metadata <- metadata[colnames(orthogroup_counts), ]

dim(metadata)
head(metadata)
```

```{r}
# dds <- DESeqDataSetFromMatrix(countData = combat_seq_counts,
#                               colData = metadata,
#                               design = ~ 1)

dds <- DESeqDataSetFromMatrix(countData = combined_counts,
                              colData = metadata,
                              design = ~ Stage2 + Species)

#fit the model and estimate differential expression
dds <- DESeq(dds)

head(results(dds, tidy=TRUE))
summary(results(dds))

saveRDS(dds, file = "/home/cluster_user/link_to_datadisk/copy/counts/dds.rds")
```

### pca

```{r}
#variance-stabilizing transformation (reduces dependence of variance on mean expression)
vst <- varianceStabilizingTransformation(dds, blind = FALSE)
vst_mat <- assay(vst)

#get metadata
sample_info <- as.data.frame(colData(vst))
sample_info$Sample <- rownames(sample_info)
sample_info <- sample_info %>% select(Sample, Sample2, Stage2, Species)

#rows: samples; columns: genes
vst_df <- as.data.frame(t(vst_mat))
vst_df$Sample <- rownames(vst_df)
#add metadata
vst_df <- left_join(vst_df, sample_info, by = "Sample")

#median on replicates (before pca)
vst_median <- vst_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

#prepare for pca
expr_median <- vst_median %>% select(-Stage2, -Species)
samples <- vst_median$Sample2
expr_values <- vst_median %>% select(-Sample2, -Stage2, -Species)

exprs_mat <- t(as.matrix(expr_values))
colnames(exprs_mat) <- vst_median$Sample2
rownames(exprs_mat) <- colnames(expr_values)
exprs_mat <- t(exprs_mat)

#ExpressionSet object
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(exprs_mat),
                                       row.names = colnames(exprs_mat)))
eset <- ExpressionSet(assayData = exprs_mat, phenoData = pheno)

#standardize (mean = 0, sd = 1)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#principal component analysis
pca <- prcomp(expr_scaled)
#explained variance
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)

#plot
pcaData <- as.data.frame(pca$x)
pcaData$Sample2 <- vst_median$Sample2
pcaData$Stage2 <- vst_median$Stage2
pcaData$Species <- vst_median$Species

p <- ggplot(pcaData, aes(PC1, PC2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of gene expression") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_all_stage_species.png",
       p, width = 11, height = 8, dpi = 300)
```

### heatmap

```{r}
#annotations
annotation_df <- vst_median %>%
  select(Sample2, Stage2, Species) %>%
  rename(Stage = Stage2) %>% 
  column_to_rownames("Sample2")

species_labels <- c(
  "blattella"  = "B. germanica",
  "drosophila" = "D. melanogaster",
  "harmonia"   = "H. axyridis",
  "locusta"    = "L. migratoria",
  "plutella"   = "P. xylostella"
)

#change species names acording to map above
annotation_df$Species <- species_labels[ as.character(annotation_df$Species) ]
annotation_df$Species <- factor(annotation_df$Species, levels = unname(species_labels))

#stages colours the same as in pca
stages_order <- c("Adult Female", "Adult Male", "Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa")
annotation_df$Stage <- factor(annotation_df$Stage, levels = stages_order)
stage_colors <- setNames(scales::hue_pal()(length(stages_order)), stages_order)

#different colours for species
species_levels <- unique(annotation_df$Species)
species_colors <- setNames(
  RColorBrewer::brewer.pal(n = max(3, length(species_levels)), "Set1")[1:length(species_levels)], species_levels
)

annotation_colors <- list(
  Stage = stage_colors,
  Species = species_colors
)

heat_colors <- colorRampPalette(rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")))(100)

#heatmap: columns: samples; rows: orthogroups; values: standardised gene expression
ph <- pheatmap(t(expr_scaled),
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         annotation_col = annotation_df,
         annotation_colors = annotation_colors,
         show_rownames = FALSE,
         show_colnames = TRUE,
         fontsize_col = 8,
         fontsize = 12,
         border_color = NA,
         angle_col = 45,
         main = "Heatmap of gene expression",
         color = heat_colors
         )

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/pheatmap_all_samples.png", ph, width = 11, height = 8)
```

### DEGs heatmap

```{r}
#heatmap of differential expression between stages
unique_stages <- levels(dds$Stage2)

#matrix for storing DEG counts between stages
mat <- matrix(0, nrow = length(unique_stages), ncol = length(unique_stages),
              dimnames = list(unique_stages, unique_stages))

#run all pairwise comparisons of DEGs
pairwise_comparisons <- combn(unique_stages, 2, simplify = FALSE)
deg_list <- list()

for (comp in pairwise_comparisons) {
  group1 <- comp[1]
  group2 <- comp[2]

  res <- tryCatch({
    results(dds, contrast = c("Stage2", group1, group2))
  }, error = function(e) {
    message(paste("Contrast", group1, "vs", group2, "failed:", e$message))
    return(NULL)
  })

  if (!is.null(res)) {
    deg_list[[paste0(group1, "_vs_", group2)]] <- as.data.frame(res)
    n_deg <- sum(res$padj < 0.05, na.rm = TRUE)

    mat[group1, group2] <- n_deg
    mat[group2, group1] <- n_deg
  }
}

#heatmap
deg_colors <- colorRampPalette(RColorBrewer::brewer.pal(n = 9, name = "Oranges"))(100)

ph <- pheatmap(mat,
         fontsize_row = 10,
         fontsize_col = 10,
         angle_col = 45,
         cellwidth = 20,
         cellheight = 20,
         border_color = NA,
         main = "Number of DEGs between stages",
         color = deg_colors
)

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/pheatmap_all_stage_species.png", ph, width = 5, height = 4)
```

### umap

```{r}
#Uniform Manifold Approximation and Projection
#Mfuzz standardise (z-score on orthogroups)
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(vst_mat),
                                       row.names = colnames(vst_mat)))
eset <- ExpressionSet(assayData = vst_mat, phenoData = pheno)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#dataframe with matadata
expr_df <- as.data.frame(t(expr_scaled)) 
expr_df$Sample <- colnames(vst_mat) 
sample_info <- as.data.frame(colData(vst)) 
sample_info$Sample <- rownames(sample_info) 
expr_df <- left_join(expr_df, sample_info, by = "Sample") 

#medians
expr_median <- expr_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

#UMAP
expr_median_mat <- expr_median %>%
  ungroup() %>%
  select(-Sample2, -Stage2, -Species) %>%
  as.matrix()

#standardise
expr_median_mat <- t(scale(t(expr_median_mat)))

umap_result <- umap(expr_median_mat, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")
umap_df <- as.data.frame(umap_result)
colnames(umap_df) <- c("UMAP1", "UMAP2")
umap_df$Sample2 <- expr_median$Sample2
umap_df$Stage2 <- expr_median$Stage2
umap_df$Species <- expr_median$Species

#plot
p <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab("UMAP1") +
  ylab("UMAP2") +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  ggtitle("UMAP of gene expression") +
  labs(color = "Stage", shape = "Species") +
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_all_stage_species.png",
       p, width = 11, height = 8, dpi = 300)
```

### pseudotime

```{r}
print(percentVar)
print(sum(percentVar[0:6]))

#from pca matrix, first 6 main components (69.4% variance explained)
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:6)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
#standardise
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

#check if pseudotime is according to the stages order
stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")

pcaData$Stage2 <- factor(pcaData$Stage2, levels = stages_order)
pcaData$Stage2 <- droplevels(pcaData$Stage2)

#check correlation with the stages order and flip if needed
cor_test <- cor(as.numeric(factor(pcaData$Stage2, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")
print(cor_test)
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

species_levels <- sort(c("blattella", "drosophila", "harmonia", "locusta", "plutella"))
pcaData$Species <- factor(pcaData$Species, levels = species_levels)
species_colors <- setNames(
  RColorBrewer::brewer.pal(length(species_levels), "Set1"),
  species_levels
)

#distribution of pseudotime values across developmental stages
p1 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    ) +
  labs(
    title = "Pseudotime by stage - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_all_stage_species.png",
       p1, width = 5, height = 5, dpi = 300)

#Pseudotime by stage and speciess across stages and species
p2 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_all_stage_species.png",
       p2, width = 7, height = 5, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1)) +
  plot_annotation(
    title = "Pseudotime distribution - PCA",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_all_stage_species_combined.png",
       combined, width = 11, height = 5, dpi = 300)

pseudotime_summary <- pcaData %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )

pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

#median pseudotime trajectory showing developmental progression
p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pseudotime trajectory - PCA",
    y = "Median pseudotime",
    x = "Developmental stage"
  )

print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_all_stage_species.png",
       p3, width = 11, height = 8, dpi = 300)
```

#### pca

```{r}
curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:6)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(x = PC1, y = PC2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression PCA",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_stage_species.png",
        p_curve, width = 11, height = 8, dpi = 300)
```

#### umap

```{r}
#same workflow as above, for UMAP data
umap_mat_all <- as.matrix(umap_df[, c("UMAP1", "UMAP2")])
rownames(umap_mat_all) <- umap_df$Sample2

#match curve on umap
umap_curve_all <- principal_curve(umap_mat_all)

#count pseudotime
lambda_all <- umap_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))
umap_df$Pseudotime <- lambda_norm_all

#check pseudotime direction
stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")
umap_df$Stage2 <- factor(umap_df$Stage2, levels = stages_order)
umap_df$Stage2 <- droplevels(umap_df$Stage2)

cor_test <- cor(as.numeric(factor(umap_df$Stage2, levels = stages_order)),
                umap_df$Pseudotime, use = "complete.obs")
print(cor_test)

#swap pseudotime if needed
if (cor_test < 0) {
  umap_df$Pseudotime <- 1 - umap_df$Pseudotime
}

p1 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime by stage - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_umap_stage_species.png", p1, width = 5, height = 5, dpi = 300)

p2 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_umap_stage_species.png", p2, width = 7, height = 5, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1)) +
  plot_annotation(
    title = "Pseudotime distribution - UMAP",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_all_stage_species_combined.png",
       combined, width = 11, height = 5, dpi = 300)

pseudotime_summary <- umap_df %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )
pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pseudotime trajectory - UMAP",
    y = "Median pseudotime",
    x = "Developmental stage"
  )
print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_umap_stage_species.png", p3, width = 11, height = 8, dpi = 300)

#trajectory curve
curve_line <- as.data.frame(umap_curve_all$s)
colnames(curve_line) <- c("UMAP1", "UMAP2")
curve_line <- curve_line[order(umap_df$Pseudotime), ]

p_umap_curve <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(
    data = curve_line,
    aes(x = UMAP1, y = UMAP2),
    inherit.aes = FALSE,
    color = "black",
    size = 1.2,
    alpha = 0.7
  ) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression UMAP",
    x = "UMAP1",
    y = "UMAP2",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_umap_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_stage_species.png",
       p_umap_curve, width = 11, height = 8, dpi = 300)
```

## no species effect

### matrix

```{r}
#| echo: true
#| eval: true
#| results: false

drosophila_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila_joined_counts.rds")

#list of dataframes with counts (per RNA-seq dataset)
all_counts <- list(
  locusta = locusta_counts,
  harmonia = harmonia_counts,
  plutella = plutella_counts,
  drosophila = drosophila_counts,
  blattella = blattella_counts
)

#combine all counts tables into one long table
#rows: all gene names; columns: all SRR and species; values: counts 
all_counts_long <- lapply(names(all_counts), function(species) {
  df <- as.data.frame(all_counts[[species]])
  df$gene_id <- rownames(df)
  df$species <- species
  df
}) %>%
  bind_rows()

head(all_counts_long)

#table: orthogroup, species, gene
orthogroup_mapping <- orthogroups %>%
  pivot_longer(-Orthogroup, names_to = "species", values_to = "genes") %>%
  filter(!is.na(genes)) %>%
  separate_rows(genes, sep = ",") %>%
  #clean gene names
  mutate(
    genes = str_trim(genes),
    species = str_replace(species, "_proteins", ""),
    genes = str_replace(genes, "\\..*", "") 
  )

#duplicate mapping for the second drosophila dataset
orthogroup_mapping_drosophila2 <- orthogroup_mapping %>%
  filter(species == "drosophila") %>%
  mutate(species = "drosophila2")

orthogroup_mapping <- bind_rows(orthogroup_mapping, orthogroup_mapping_drosophila2)

orthogroup_mapping %>%
  dplyr::count(species)

head(orthogroup_mapping)


#join count data with orthogroup mapping
merged <- all_counts_long %>%
  inner_join(
    orthogroup_mapping,
    by = c("species", "gene_id" = "genes")
  )

head(merged)
merged %>%
  dplyr::count(species) %>%
  print()

orthogroup_counts <- merged %>%
  select(Orthogroup, where(is.numeric)) %>%
  group_by(Orthogroup) %>%
  summarise(across(everything(), ~sum(.x, na.rm = TRUE)))

head(orthogroup_counts)

orthogroup_counts <- orthogroup_counts %>%
  column_to_rownames(var = "Orthogroup")

head(orthogroup_counts)
dim(orthogroup_counts)

colnames(orthogroup_counts)

#narrow to only 1:1 orthogroups
orthogroup_counts_1to1 <- orthogroup_counts %>%
  filter(rownames(.) %in% single_copy)

#check if anything overlaps between 1:1 from orthofinder and orthosnap (no)
overlapping_genes <- intersect(rownames(orthogroup_counts_1to1), rownames(orthosnap_counts))

length(overlapping_genes)
print(overlapping_genes)

#add orthosnap
combined_counts <- bind_rows(
  orthogroup_counts_1to1,
  orthosnap_counts[!rownames(orthosnap_counts) %in% rownames(orthogroup_counts_1to1), ]
)

combined_counts[is.na(combined_counts)] <- 0
```

### deseq2

```{r}
metadata <- read.csv(file="/home/cluster_user/Magisterka/metadata2.csv", row.names = 1, header = TRUE, sep = ",") 

metadata <- metadata[colnames(orthogroup_counts), ]

dds <- DESeqDataSetFromMatrix(
  countData = combined_counts,
  colData   = metadata,
  design    = ~ Species + Stage2
)

dds <- estimateSizeFactors(dds, type = "poscounts")

#dds <- DESeq(dds) without this step as I'll do only visualisation as i have modified counts for drosophila

saveRDS(dds, file = "/home/cluster_user/link_to_datadisk/copy/counts/dds_batch.rds")
```

### pca

```{r}
#removing species effect by doing pca on residuas from model with species factor
# VST
vst <- varianceStabilizingTransformation(dds, blind = FALSE)
vst_mat <- assay(vst)

#linear model to remove the species effect (applied for each orthogroup separately)
residuals_mat <- apply(vst_mat, 1, function(x) {
  lm(x ~ colData(vst)$Species)$residuals
})
residuals_mat <- t(residuals_mat)

# residuals_mat <- removeBatchEffect(
#     vst_mat,
#     batch = colData(vst)$Species
# )

# residuals_mat <- ComBat(dat = vst_mat,
#                         batch = colData(vst)$Species)

#add metadata
sample_info <- as.data.frame(colData(vst))
sample_info$Sample <- rownames(sample_info)
sample_info <- sample_info %>% select(Sample, Sample2, Stage2, Species)

#data.frame residuals + Sample
residuals_df <- as.data.frame(t(residuals_mat))
residuals_df$Sample <- rownames(residuals_df)
residuals_df <- left_join(residuals_df, sample_info, by = "Sample")

#median for each Sample2/Stage2
residuals_median <- residuals_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

#make matrix
expr_median <- residuals_median %>% select(-Stage2, -Species)
samples <- expr_median$Sample2
expr_values <- expr_median %>% select(-Sample2)
exprs_mat <- t(as.matrix(expr_values))
colnames(exprs_mat) <- expr_median$Sample2
rownames(exprs_mat) <- colnames(expr_values)
exprs_mat <- t(exprs_mat)

#standardize Mfuzz -> z-score
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(exprs_mat), row.names = colnames(exprs_mat)))
eset <- ExpressionSet(assayData = exprs_mat, phenoData = pheno)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#PCA
pca <- prcomp(expr_scaled)
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)
pcaData <- as.data.frame(pca$x)
pcaData$Sample2 <- residuals_median$Sample2
pcaData$Stage2 <- residuals_median$Stage2
pcaData$Species <- residuals_median$Species

p <- ggplot(pcaData, aes(PC1, PC2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of gene expression (removed species effect)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_all_stage.png",
       p, width = 11, height = 8, dpi = 300)

```

#### test

```{r}
# dist_res  <- dist(t(residuals_mat))
# 
# bd_stage <- betadisper(dist_res, metadata$Stage2)
# anova(bd_stage)
# 
# bd_species <- betadisper(dist_res, metadata$Species)
# anova(bd_species)
# 
# adonis2(dist_mat ~ Stage2 + Species, data = metadata, permutations = 999, by = "margin")
# adonis2(dist_res  ~ Stage2 + Species, data = metadata, permutations = 999, by = "margin")
# 
# p_before <- sapply(1:nrow(vst_mat), function(i) {
#   an <- anova(lm(vst_mat[i, ] ~ metadata$Stage2))
#   an["metadata$Stage2","Pr(>F)"]
# })
# p_after <- sapply(1:nrow(residuals_mat), function(i) {
#   an <- anova(lm(residuals_mat[i, ] ~ metadata$Stage2))
#   an["metadata$Stage2","Pr(>F)"]
# })
# sum(p.adjust(p_before, "BH") < 0.05)
# sum(p.adjust(p_after, "BH")  < 0.05)
```

### heatmap

```{r}
#annotations
annotation_df <- residuals_median %>%
  select(Sample2, Stage2, Species) %>%
  rename(Stage = Stage2) %>% 
  column_to_rownames("Sample2")

#change species names to full
annotation_df$Species <- species_labels[ as.character(annotation_df$Species) ]
annotation_df$Species <- factor(annotation_df$Species, levels = unname(species_labels))

#stages colours the same as in pca
stages_order <- c("Adult Female", "Adult Male", "Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa")
annotation_df$Stage <- factor(annotation_df$Stage, levels = stages_order)
stage_colors <- setNames(scales::hue_pal()(length(stages_order)), stages_order)

#different colours for species
species_levels <- unique(annotation_df$Species)
species_colors <- setNames(
  RColorBrewer::brewer.pal(n = max(3, length(species_levels)), "Set1")[1:length(species_levels)],
  species_levels
)

annotation_colors <- list(
  Stage = stage_colors,
  Species = species_colors
)

ph <- pheatmap(t(expr_scaled),
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         annotation_col = annotation_df,
         annotation_colors = annotation_colors,
         show_rownames = FALSE,
         show_colnames = TRUE,
         fontsize_col = 8,
         fontsize = 12,
         border_color = NA,
         angle_col = 45,
         main = "Heatmap of gene expression (removed species effect)",
         color = heat_colors
         )

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/heatmap_all_stage.png", ph, width = 12, height = 8)
```

### umap

```{r}
#Mfuzz standardise (z-score on orthogroups)
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(residuals_mat),
                                       row.names = colnames(residuals_mat)))
eset <- ExpressionSet(assayData = residuals_mat, phenoData = pheno)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#dataframe with matadata
expr_df <- as.data.frame(t(expr_scaled)) 
expr_df$Sample <- colnames(vst_mat) 
sample_info <- as.data.frame(colData(vst)) 
sample_info$Sample <- rownames(sample_info) 
expr_df <- left_join(expr_df, sample_info, by = "Sample") 

#medians
expr_median <- expr_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

#UMAP
expr_median_mat <- expr_median %>%
  ungroup() %>%
  select(-Sample2, -Stage2, -Species) %>%
  as.matrix()

#standardise
expr_median_mat <- t(scale(t(expr_median_mat)))

umap_result <- umap(expr_median_mat, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")
umap_df <- as.data.frame(umap_result)
colnames(umap_df) <- c("UMAP1", "UMAP2")
umap_df$Sample2 <- expr_median$Sample2
umap_df$Stage2 <- expr_median$Stage2
umap_df$Species <- expr_median$Species

#plot
p <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab("UMAP1") +
  ylab("UMAP2") +
  ggtitle("UMAP of gene expression (removed species effect)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_all_stage.png",
       p, width = 11, height = 8, dpi = 300)

```

### cluster

```{r}
expr_values <- expr_median %>%
  ungroup() %>%
  select(-Sample2, -Stage2) %>%
  as.matrix()

rownames(expr_values) <- expr_median$Sample2

dist_mat <- dist(expr_values, method = "euclidean")

hc <- hclust(dist_mat, method = "average")

#maps Sample2 -> Stage2
labels_df <- expr_median %>% select(Sample2, Stage2)
stages_vec <- setNames(labels_df$Stage2, labels_df$Sample2)

stage_levels <- sort(unique(labels_df$Stage2))

dend <- as.dendrogram(hc)

dend <- dend %>%
  dendextend::set("labels_colors", value = stage_colors[stages_vec[labels(dend)]])

png("/home/cluster_user/Magisterka/plots/dendrogram_all_stage.png",
    width = 1600, height = 800, res = 150)

par(mar = c(10, 4, 4, 4))
plot(dend, main = "Hierarchical clustering (removed species effect)")
legend("topright", inset = c(0.05, 0), legend = names(stage_colors),
       fill = stage_colors, border = NA, bty = "n")

dev.off() 
```

```{r}
#so this is incorrect
# #pheatmap stage
# pheno <- as.data.frame(colData(dds))
# design <- model.matrix(~ 0 + Stage2, data = pheno)
# 
# fit <- lmFit(residuals_mat, design)
# 
# #contrasts by stage2
# contrast_list <- lapply(combn(stage_levels, 2, simplify = FALSE), function(x) {
#   contrast <- rep(0, length(stage_levels))
#   names(contrast) <- stage_levels
#   contrast[x[1]] <- 1
#   contrast[x[2]] <- -1
#   contrast
# })
# 
# contrast_matrix <- do.call(cbind, contrast_list)
# colnames(contrast_matrix) <- sapply(combn(stage_levels, 2, simplify = FALSE), function(x) paste(x, collapse = "_vs_"))
# 
# fit2 <- contrasts.fit(fit, contrast_matrix)
# fit2 <- eBayes(fit2)
# 
# deg_list <- list()
# mat <- matrix(0, nrow = length(stage_levels), ncol = length(stage_levels),
#               dimnames = list(stage_levels, stage_levels))
# 
# for (comp_name in colnames(contrast_matrix)) {
#   res <- topTable(fit2, coef = comp_name, number = Inf, adjust.method = "BH", sort.by = "none")
#   deg_list[[comp_name]] <- res
#   comps <- strsplit(comp_name, "_vs_")[[1]]
#   
#   n_deg <- sum(res$adj.P.Val < 0.05, na.rm = TRUE)
#   mat[comps[1], comps[2]] <- n_deg
#   mat[comps[2], comps[1]] <- n_deg
# }
# 
# # Heatmapa
# ph <- pheatmap(mat,
#          fontsize_row = 10,
#          fontsize_col = 10,
#          angle_col = 45,
#          cellwidth = 20,
#          cellheight = 20,
#          border_color = NA,
#          main = "Number of DEGs between stages\n(removed species effect)",
#          color = deg_colors
# )
# 
# grid.newpage()
# grid.draw(ph$gtable)
# 
# ggsave("/home/cluster_user/Magisterka/plots/pheatmap_all_stage.png", ph, width = 5, height = 4.5)
```

### pseudotime

```{r}
print(percentVar)
print(sum(percentVar[0:6]))

#from pca matrix, first 6 main components (59.5% variance explained)
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:6)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")

pcaData$Stage2 <- factor(pcaData$Stage2, levels = stages_order)
pcaData$Stage2 <- droplevels(pcaData$Stage2)

cor_test <- cor(as.numeric(factor(pcaData$Stage2, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")

print(cor_test)
#make sure pseudotime goes from eggs to adults and not from adult to eggs
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

p1 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime by stage - PCA\n(removed species effect)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_all_stage.png",
       p1, width = 5, height = 5, dpi = 300)

species_levels <- sort(c("blattella", "drosophila", "harmonia", "locusta", "plutella"))
pcaData$Species <- factor(pcaData$Species, levels = species_levels)
species_colors <- setNames(
  RColorBrewer::brewer.pal(length(species_levels), "Set1"),
  species_levels
)

p2 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = species_colors,
    #breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - PCA\n(removed species effect)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_all_stage.png",
       p2, width = 7, height = 5, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1)) +
  plot_annotation(
    title = "Pseudotime distribution - PCA (removed species effect)",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_all_stage_combined.png",
       combined, width = 11, height = 5, dpi = 300)

pseudotime_summary <- pcaData %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )

pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = species_colors,
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pseudotime trajectory (removed species effect) - PCA",
    y = "Median pseudotime",
    x = "Developmental stage"
  )

print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_all_stage.png",
       p3, width = 11, height = 8, dpi = 300)
```

#### pca

```{r}
curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:6)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(x = PC1, y = PC2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression PCA (removed species effect)",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_all_stage.png",
        p_curve, width = 11, height = 8, dpi = 300)
```

#### umap

```{r}
umap_mat_all <- as.matrix(umap_df[, c("UMAP1", "UMAP2")])
rownames(umap_mat_all) <- umap_df$Sample2

#match curve on umap
umap_curve_all <- principal_curve(umap_mat_all)

#count pseudotime
lambda_all <- umap_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))
umap_df$Pseudotime <- lambda_norm_all

#check pseudotime direction
stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")
umap_df$Stage2 <- factor(umap_df$Stage2, levels = stages_order)
umap_df$Stage2 <- droplevels(umap_df$Stage2)

cor_test <- cor(as.numeric(factor(umap_df$Stage2, levels = stages_order)),
                umap_df$Pseudotime, use = "complete.obs")
print(cor_test)

#swap pseudotime if needed
if (cor_test < 0) {
  umap_df$Pseudotime <- 1 - umap_df$Pseudotime
}

#plots

#pseudotime vs stage
p1 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime by stage - UMAP\n(removed species effect)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_umap_all_stage.png", p1, width = 5, height = 5, dpi = 300)

#pseudotime vs stage+species
p2 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - UMAP\n(removed species effect)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_umap_all_stage.png", p2, width = 7, height = 5, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1)) +
  plot_annotation(
    title = "Pseudotime distribution - UMAP (removed species effect)",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_all_stage_combined.png",
       combined, width = 11, height = 5, dpi = 300)

#pseudotime summary line plot
pseudotime_summary <- umap_df %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )
pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pseudotime trajectory (removed species effect) - UMAP",
    y = "Median pseudotime",
    x = "Developmental stage"
  )
print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_umap_all_stage.png", p3, width = 11, height = 8, dpi = 300)

#trajectory curve
curve_line <- as.data.frame(umap_curve_all$s)
colnames(curve_line) <- c("UMAP1", "UMAP2")
curve_line <- curve_line[order(umap_df$Pseudotime), ]

#umap plot
p_umap_curve <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(
    data = curve_line,
    aes(x = UMAP1, y = UMAP2),
    inherit.aes = FALSE,
    color = "black",
    size = 1.2,
    alpha = 0.7
  ) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression UMAP (removed species effect)",
    x = "UMAP1",
    y = "UMAP2",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_umap_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_all_stage.png",
       p_umap_curve, width = 11, height = 8, dpi = 300)
```

## statistical tests

```{r}
# Convert DESeq2 colData to data.frame
sample_info <- as.data.frame(colData(vst))
sample_info$Sample <- rownames(sample_info)

# Keep all relevant columns including instar (Stage)
sample_info <- sample_info %>%
  select(Sample, Sample2, Stage, Stage2, Species)

# Transpose residuals matrix to data.frame
residuals_df <- as.data.frame(t(residuals_mat))
residuals_df$Sample <- rownames(residuals_df)

# Join metadata
residuals_df <- left_join(residuals_df, sample_info, by = "Sample")

residuals_median_instars <- residuals_df %>%
  group_by(Sample2, Stage, Stage2, Species) %>% 
  summarise(across(where(is.numeric), median), .groups = "drop")

expr_dist <- residuals_median_instars %>%
  select(-Stage, -Stage2, -Species) %>%
  column_to_rownames("Sample2") %>%
  as.matrix()

#Calculate Spearman correlation distance
dist_mat <- 1 - cor(t(expr_dist), method = "spearman")

# Save sample names
samples <- rownames(dist_mat)

# Convert distance matrix to long-format data.frame
# the data is non independent!!! (distance of samples pairs)
dist_df <- as.data.frame(as.table(dist_mat))
colnames(dist_df) <- c("Sample1", "Sample2", "distance")

# Add metadata for both samples
meta <- residuals_median_instars %>%
  select(Sample2, Stage, Stage2, Species) %>%
  distinct()

# Join metadata for Sample1
dist_df <- dist_df %>%
  left_join(meta, by = c("Sample1" = "Sample2")) %>%
  rename(Stage1 = Stage,
         Stage2_1 = Stage2,
         Species1 = Species)

# Join metadata for Sample2
dist_df <- dist_df %>%
  left_join(meta, by = c("Sample2" = "Sample2")) %>%
  rename(Stage2 = Stage,
         Stage2_2 = Stage2,
         Species2 = Species)

# Remove self-comparisons
dist_df <- dist_df %>% filter(Sample1 != Sample2)

#. Create broader developmental categories (Meta-stages)
meta_stage <- function(stage) {
  case_when(
    grepl("^Larva", stage) ~ "Larva",
    grepl("^Nymph", stage) ~ "Nymph",
    grepl("^Prepupa", stage) ~ "Prepupa",
    grepl("^Pupa", stage) ~ "Pupa",
    grepl("^Embryo|Egg", stage) ~ "Embryo",
    grepl("^Adult", stage) ~ "Adult",
    TRUE ~ NA_character_
  )
}

dist_df <- dist_df %>%
  mutate(
    instar_number = case_when(
      grepl("Nymph", Stage1) ~ as.numeric(gsub(".*?(\\d+).*", "\\1", Stage1)),
      grepl("instar", Stage1) ~ as.numeric(gsub(".*?(\\d+).*", "\\1", Stage1)),
      TRUE ~ NA_real_
    )
  )

#  Compare nymphs to Larva, Prepupa, Pupa
nymph_comp <- dist_df %>%
  filter(Stage2_1 == "Nymph" & Stage2_2 %in% c("Larva", "Prepupa", "Pupa"))

# Boxplot - overview of pairwise distances
p <- ggplot(nymph_comp, aes(x = Stage2_2, y = distance, fill = Stage2_2)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black", size = 1.5) +
  scale_fill_manual(values = stage_colors) +
  theme_classic(base_size = 14) +
  labs(
    x = "Developmental stage",
    y = "Spearman distance",
    title = "Nymph distances to other developmental stages"
  ) +
  theme(legend.position = "none")

print(p)
ggsave("/home/cluster_user/Magisterka/plots/spearman_dist_violin.png",
       p, width = 8, height = 5, dpi = 300)
```

#### early vs late

##### prepupa

```{r}
#Early vs late nymph instars; locusta 1,2,3,4,5 blattella: 1,3,5,6
dist_df <- dist_df %>%
  mutate(
    NymphType1 = case_when(
      instar_number %in% 1:3 ~ "Early",
      instar_number %in% 4:6 ~ "Late",
      TRUE ~ NA_character_
    ),
    NymphType1 = factor(NymphType1, levels = c("Early", "Late"))
  )

nymph_trend <- dist_df %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 == "Prepupa",
    !is.na(NymphType1)
  )

p <- ggplot(nymph_trend, aes(x = NymphType1, y = distance)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymph instar type (1-3 vs 4-6)",
    y = "Spearman distance to prepupa",
    title = "Early vs late nymph distances to prepupa"
  ) +
  theme(legend.position = "right")

print(p)
ggsave("/home/cluster_user/Magisterka/plots/early_vs_late_prepupa.png",
       p, width = 6, height = 5, dpi = 300)

model_early_late <- lm(
  distance ~ NymphType1 + Species2,
  data = nymph_trend
)
anova(model_early_late)
summary(model_early_late)

#for more extreme stages
dist_df <- dist_df %>%
  mutate(
    NymphType1_12_56 = case_when(
      instar_number %in% c(1, 2) ~ "Early (1-2)",
      instar_number %in% c(5, 6) ~ "Late (5-6)",
      TRUE ~ NA_character_
    )
  )

nymph_trend_12_56 <- dist_df %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 == "Prepupa",
    !is.na(NymphType1_12_56)
  )

p <- ggplot(nymph_trend_12_56, aes(x = NymphType1_12_56, y = distance)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymph instar type (1-2 vs 5-6)",
    y = "Spearman distance to prepupa",
    title = "Early vs late nymph distances to prepupa"
  ) +
  theme(legend.position = "right")

print(p)
ggsave("/home/cluster_user/Magisterka/plots/very_early_vs_late_prepupa.png",
       p, width = 6, height = 5, dpi = 300)

model_extreme <- lm(
  distance ~ NymphType1_12_56 + Species2,
  data = nymph_trend_12_56
)
anova(model_extreme)
summary(model_extreme)

#late nymphs are closer to prepupa
```

##### larva

```{r}
nymph_trend <- dist_df %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 == "Larva",
    !is.na(NymphType1)
  )

p <- ggplot(nymph_trend, aes(x = NymphType1, y = distance)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymph instar type (1-3 vs 4-6)",
    y = "Spearman distance to larva",
    title = "Early vs late nymph distances to larva"
  ) +
  theme(legend.position = "right")

print(p)
ggsave("/home/cluster_user/Magisterka/plots/early_vs_late_larva.png",
       p, width = 6, height = 5, dpi = 300)

model_early_late <- lm(
  distance ~ NymphType1 + Species2,
  data = nymph_trend
)
anova(model_early_late)
summary(model_early_late)

#for more extreme stages
dist_df <- dist_df %>%
  mutate(
    NymphType1_12_56 = case_when(
      instar_number %in% c(1, 2) ~ "Early (1-2)",
      instar_number %in% c(5, 6) ~ "Late (5-6)",
      TRUE ~ NA_character_
    )
  )

nymph_trend_12_56 <- dist_df %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 == "Larva",
    !is.na(NymphType1_12_56)
  )

p <- ggplot(nymph_trend_12_56, aes(x = NymphType1_12_56, y = distance)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymph instar type (1-2 vs 5-6)",
    y = "Spearman distance to larva",
    title = "Early vs late nymph distances to larva"
  ) +
  theme(legend.position = "right")

print(p)
ggsave("/home/cluster_user/Magisterka/plots/very_early_vs_late_larva.png",
       p, width = 6, height = 5, dpi = 300)

model_extreme <- lm(
  distance ~ NymphType1_12_56 + Species2,
  data = nymph_trend_12_56
)
anova(model_extreme)
summary(model_extreme)

#larva are closer to early nymphs
```

##### pupa

```{r}
nymph_trend <- dist_df %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 == "Pupa",
    !is.na(NymphType1)
  )

p <- ggplot(nymph_trend, aes(x = NymphType1, y = distance)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymph instar type (1-3 vs 4-6)",
    y = "Spearman distance to pupa",
    title = "Early vs late nymph distances to pupa"
  ) +
  theme(legend.position = "right")

print(p)
ggsave("/home/cluster_user/Magisterka/plots/early_vs_late_pupa.png",
       p, width = 6, height = 5, dpi = 300)

model_early_late <- lm(
  distance ~ NymphType1 + Species2,
  data = nymph_trend
)
anova(model_early_late)
summary(model_early_late)

#for more extreme stages
dist_df <- dist_df %>%
  mutate(
    NymphType1_12_56 = case_when(
      instar_number %in% c(1, 2) ~ "Early (1-2)",
      instar_number %in% c(5, 6) ~ "Late (5-6)",
      TRUE ~ NA_character_
    )
  )

nymph_trend_12_56 <- dist_df %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 == "Pupa",
    !is.na(NymphType1_12_56)
  )

p <- ggplot(nymph_trend_12_56, aes(x = NymphType1_12_56, y = distance)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymph instar type (1-2 vs 5-6)",
    y = "Spearman distance to pupa",
    title = "Early vs late nymph distances to pupa"
  ) +
  theme(legend.position = "right")

print(p)
ggsave("/home/cluster_user/Magisterka/plots/very_early_vs_late_pupa.png",
       p, width = 6, height = 5, dpi = 300)

model_extreme <- lm(
  distance ~ NymphType1_12_56 + Species2,
  data = nymph_trend_12_56
)
anova(model_extreme)
summary(model_extreme)

#connection to nymph instars isnt significant
```

#### instars

```{r}
trend_df <- dist_df %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 %in% c("Larva", "Prepupa", "Pupa"),
    !is.na(instar_number)
  )

cor_results <- trend_df %>%
  group_by(Stage2_2) %>%
  summarise(
    rho = cor(instar_number, distance, method = "spearman"),
    n = n(),
    .groups = "drop"
  )

print(cor_results)

#counted separate for each metastage
#   Stage2_2      rho     n
# 1 Larva     0.242     126 --> distance growths in next stages)
# 2 Prepupa  -0.314     117 --> distance falls in next stages
# 3 Pupa     -0.00894    99 --> only very small decrease, not visible

cor_labels <- cor_results %>%
  mutate(
    label = paste0("rho == ", round(rho, 2))
  )

p <- ggplot(trend_df, aes(instar_number, distance, color = Stage2_2)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  facet_wrap(~ Stage2_2, nrow = 1, scales = "free_y") +
  geom_text(
    data = cor_labels,
    aes(
      x = -Inf, y = Inf, label = label
    ),
    inherit.aes = FALSE,
    hjust = -0.1, vjust = 1.1,
    size = 4,
    parse = TRUE
  ) +
  scale_color_manual(values = stage_colors) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 13, face = "bold")
  ) +
  labs(
    x = "Nymph instar number",
    y = "Spearman distance",
    title = "Distance to to other stages across nymphal development"
  )

p <- ggplot(trend_df, aes(instar_number, distance, color = Stage2_2)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ Stage2_2, nrow = 1, scales = "free_y") +
  geom_text(
    data = cor_results,
    aes(x = -Inf, y = Inf, label = paste0("rho = ", round(rho, 2))),
    inherit.aes = FALSE,
    hjust = -0.1, vjust = 1.1
  ) +
  theme_classic()

print(p)
ggsave("/home/cluster_user/Magisterka/plots/spearman_dist_instar.png",
       p, width = 8, height = 5, dpi = 300)
```

#### models

```{r}
lm_model <- lmer(distance ~ instar_number * Stage2_2 + (1 | Sample2), data = trend_df)
anova(lm_model)
print(lm_model)

# Analysis of Variance Table
#                        npar  Sum Sq  Mean Sq F value
# instar_number             1 0.00072 0.000724  0.0277
# Stage2_2                  2 0.18490 0.092452  3.5330
# instar_number:Stage2_2    2 0.52726 0.263632 10.0746 --> big effect
# Linear mixed model fit by REML ['lmerMod']
# Formula: distance ~ instar_number * Stage2_2 + (1 | Sample2)
#    Data: trend_df
# REML criterion at convergence: -224.0655
# Random effects:
#  Groups   Name        Std.Dev.
#  Sample2  (Intercept) 0.03938 
#  Residual             0.16177 
# Number of obs: 342, groups:  Sample2, 38
# Fixed Effects:
#(Intercept)                  instar_number                Stage2_2Prepupa  
#0.90374                        0.02525                        0.20472  
#Stage2_2Pupa  instar_number:Stage2_2Prepupa     instar_number:Stage2_2Pupa  
#0.03501                       -0.05484                       -0.02538  

lm_model2 <- lmer(distance ~ instar_number * Stage2_2 + Species2 + (1 | Sample2),
                  data = trend_df) #<-- this should be the best (most correct) model
#(1 | Sample2) -> intercept for Sample2 (not parameter for each sample) -> accounts for replications of samples
#(1 | Sample1) -> has variance close to 0, so i removed it
#Species2 controls for variation between species
#lmer linear mixed-effects model  - fixed effects + random effects (sample related)
anova(lm_model2)
print(lm_model2)

# Analysis of Variance Table
#                        npar  Sum Sq  Mean Sq F value
# instar_number             1 0.00072 0.000724  0.0277
# Stage2_2                  2 0.20075 0.100375  3.8358
# Species2                  2 0.13082 0.065410  2.4996
# instar_number:Stage2_2    2 0.52726 0.263632 10.0746 --> strong effect
# Linear mixed model fit by REML ['lmerMod']
# Formula: distance ~ instar_number * Stage2_2 + Species2 + (1 | Sample2)
#    Data: trend_df
# REML criterion at convergence: -217.6775
# Random effects:
#  Groups   Name        Std.Dev.
#  Sample2  (Intercept) 0.03462 
#  Residual             0.16177 
# Number of obs: 342, groups:  Sample2, 38
# Fixed Effects:
#(Intercept)                  instar_number                Stage2_2Prepupa  
#0.87659                        0.02525                        0.19863  
#Stage2_2Pupa               Species2harmonia               Species2plutella  
#0.03151                        0.05209                        0.04294  
#instar_number:Stage2_2Prepupa     instar_number:Stage2_2Pupa  
#-0.05484                       -0.02538  

#larva +0.02525
#prepupa +0.02525-0.05484=-0.02959
#pupa +0.02525-0.02538=-0.00013
```

#### embryo

```{r}
#is larva closer to embryo or nymph
embryo_test <- dist_df %>%
  filter(
    Stage2_1 == "Larva",
    Stage2_2 %in% c("Embryo", "Nymph", "Prepupa", "Pupa")
  )

model <- lmer(distance ~ Stage2_2 + Species2 + (1 | Sample2), data = embryo_test)

anova(model)
print(model)

# Analysis of Variance Table
#          npar Sum Sq Mean Sq F value
# Stage2_2    3 1.8602 0.62006 22.7411 -----> strong effect
# Species2    4 0.1172 0.02930  1.0746
# Linear mixed model fit by REML ['lmerMod']
# Formula: distance ~ Stage2_2 + Species2 + (1 | Sample2)
#    Data: embryo_test
# REML criterion at convergence: -395.8238
# Random effects:
#  Groups   Name        Std.Dev.
#  Sample2  (Intercept) 0.05447 
#  Residual             0.16512 
# Number of obs: 616, groups:  Sample2, 44
# Fixed Effects:
#        (Intercept)       Stage2_2Nymph     Stage2_2Prepupa        Stage2_2Pupa  Species2drosophila  
#            1.04049            -0.06651             0.13236             0.09133             0.01637  
#   Species2harmonia     Species2locusta    Species2plutella  
#            0.01746             0.02506             0.07888 

# Larva–Embryo	1.040
# Larva–Nymph	0.974
# Larva–Prepupa	1.173
# Larva–Pupa	1.131

#larva is 0.06651 closer to nymph than to embryo --> Berlese-
#also further to pre/pupa than to embryo 
#closest to early nymph (connected to previous test) -> Hinton+
#and it's okay larva are supposed to be higly specialised embryo in Berlese so they don't need to be super close, but then it shouldn't be closer to nymphs

emmeans(model, pairwise ~ Stage2_2)

# $emmeans
#  Stage2_2 emmean     SE df lower.CL upper.CL
#  Embryo     1.07 0.0263 36     1.01     1.12
#  Nymph      1.00 0.0403 36     0.92     1.08
#  Prepupa    1.20 0.0289 36     1.14     1.26
#  Pupa       1.16 0.0296 36     1.10     1.22
# 
# Results are averaged over the levels of: Species2 
# Degrees-of-freedom method: kenward-roger 
# Confidence level used: 0.95 
# 
# $contrasts
#  contrast         estimate     SE df t.ratio p.value
#  Embryo - Nymph     0.0665 0.0470 36   1.414  0.4989 --> not significant
#  Embryo - Prepupa  -0.1324 0.0405 36  -3.268  0.0122
#  Embryo - Pupa     -0.0913 0.0405 36  -2.255  0.1280
#  Nymph - Prepupa   -0.1989 0.0621 36  -3.204  0.0144
#  Nymph - Pupa      -0.1578 0.0621 36  -2.543  0.0700
#  Prepupa - Pupa     0.0410 0.0288 36   1.426  0.4919
# 
# Results are averaged over the levels of: Species2 
# Degrees-of-freedom method: kenward-roger 
# P value adjustment: tukey method for comparing a family of 4 estimates 

#it's not significant that nymph is closer to larva than embryo but still Berlese needs that embryo is closer -> they are similar so it's still more Hinton where the transfers between stages are more fluid
```

## no embryo&adults

```{r}
metadata_filtered <- metadata[!(metadata$Stage2 %in% c("Embryo", "Eggs","Adult Female", "Adult Male" )), ]
counts_filtered <- combined_counts[, rownames(metadata_filtered)]

dds <- DESeqDataSetFromMatrix(countData = counts_filtered,
                              colData = metadata_filtered,
                              design = ~ Stage2)

dds <- estimateSizeFactors(dds)
#dds <- DESeq(dds)
```

### pca

```{r}
vst <- varianceStabilizingTransformation(dds, blind = FALSE)
vst_mat <- assay(vst)

#linear model to delete the species effect
residuals_mat <- apply(vst_mat, 1, function(x) {
  lm(x ~ colData(vst)$Species)$residuals
})
residuals_mat <- t(residuals_mat)

# residuals_mat <- removeBatchEffect(
#     vst_mat,
#     batch = colData(vst)$Species,
#     design = model.matrix(~ colData(vst)$Stage2)
# )

# residuals_mat <- ComBat(dat = vst_mat,
#                         batch = colData(vst)$Species)

sample_info <- as.data.frame(colData(vst))
sample_info$Sample <- rownames(sample_info)
sample_info <- sample_info %>% select(Sample, Sample2, Stage2, Species)

#data.frame residuals + Sample
residuals_df <- as.data.frame(t(residuals_mat))
residuals_df$Sample <- rownames(residuals_df)
residuals_df <- left_join(residuals_df, sample_info, by = "Sample")

#median for each Sample2/Stage2
residuals_median <- residuals_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

expr_median <- residuals_median %>% select(-Stage2, -Species)
samples <- expr_median$Sample2
expr_values <- expr_median %>% select(-Sample2)
exprs_mat <- t(as.matrix(expr_values))
colnames(exprs_mat) <- expr_median$Sample2
rownames(exprs_mat) <- colnames(expr_values)
exprs_mat <- t(exprs_mat)

#standardize Mfuzz
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(exprs_mat), row.names = colnames(exprs_mat)))
eset <- ExpressionSet(assayData = exprs_mat, phenoData = pheno)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#PCA
pca <- prcomp(expr_scaled)
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)

pcaData <- as.data.frame(pca$x)
pcaData$Sample2 <- residuals_median$Sample2
pcaData$Stage2 <- residuals_median$Stage2
pcaData$Species <- residuals_median$Species

#stages colours the same as in previous pca
stages_order <- c("Larva", "Nymph", "Prepupa", "Pupa")
annotation_df$Stage <- factor(annotation_df$Stage, levels = stages_order)
stage_colors <- setNames(scales::hue_pal()(length(stages_order) + 4)[5:(length(stages_order) + 4)], stages_order)

p <- ggplot(pcaData, aes(PC1, PC2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of gene expression (removed species effect, chosen stages)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  scale_color_manual(values = stage_colors)+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_no_e_stage.png",
       p, width = 11, height = 8, dpi = 300)
```

### heatmap

```{r}
#annotations
annotation_df <- residuals_median %>%
  select(Sample2, Stage2, Species) %>%
  rename(Stage = Stage2) %>% 
  column_to_rownames("Sample2")

#change species names to full
annotation_df$Species <- factor(species_labels[annotation_df$Species],
                                levels = species_labels)

#different colours for species
species_levels <- unique(annotation_df$Species)
species_colors <- setNames(
  RColorBrewer::brewer.pal(n = max(3, length(species_levels)), "Set1")[1:length(species_levels)],
  species_levels
)

annotation_colors <- list(
  Stage = stage_colors,
  Species = species_colors
)

ph <- pheatmap(t(expr_scaled),
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         annotation_col = annotation_df,
         annotation_colors = annotation_colors,
         show_rownames = FALSE,
         show_colnames = TRUE,
         fontsize_col = 8,
         fontsize = 12,
         border_color = NA,
         angle_col = 45,
         main = "Heatmap of gene expression (removed species effect, chosen stages)",
         color = heat_colors
         )

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/heatmap_no_e_stage.png", ph, width = 12, height = 8)
```

### umap

```{r}
#Mfuzz standardise (z-score on orthogroups)
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(residuals_mat),
                                       row.names = colnames(residuals_mat)))
eset <- ExpressionSet(assayData = residuals_mat, phenoData = pheno)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#dataframe with matadata
expr_df <- as.data.frame(t(expr_scaled)) 
expr_df$Sample <- colnames(vst_mat) 
sample_info <- as.data.frame(colData(vst)) 
sample_info$Sample <- rownames(sample_info) 
expr_df <- left_join(expr_df, sample_info, by = "Sample") 

#medians
expr_median <- expr_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

#UMAP
expr_median_mat <- expr_median %>%
  ungroup() %>%
  select(-Sample2, -Stage2, -Species) %>%
  as.matrix()

#standardise
expr_median_mat <- t(scale(t(expr_median_mat)))

umap_result <- umap(expr_median_mat, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")
umap_df <- as.data.frame(umap_result)
colnames(umap_df) <- c("UMAP1", "UMAP2")
umap_df$Sample2 <- expr_median$Sample2
umap_df$Stage2 <- expr_median$Stage2
umap_df$Species <- expr_median$Species

#plot
p <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab("UMAP1") +
  ylab("UMAP2") +
  ggtitle("UMAP of gene expression (removed species effect, chosen stages)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  scale_color_manual(values = stage_colors)+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_no_e_stage.png",
       p, width = 11, height = 8, dpi = 300)
```

### cluster

```{r}
expr_values <- expr_median %>%
  ungroup() %>%
  select(-Sample2, -Stage2) %>%
  as.matrix()

rownames(expr_values) <- expr_median$Sample2

dist_mat <- dist(expr_values, method = "euclidean")

hc <- hclust(dist_mat, method = "average")

#maps Sample2 -> Stage2
labels_df <- expr_median %>% select(Sample2, Stage2)
stages_vec <- setNames(labels_df$Stage2, labels_df$Sample2)

stage_levels <- sort(unique(labels_df$Stage2))

dend <- as.dendrogram(hc)

dend <- dend %>%
  dendextend::set("labels_colors", value = stage_colors[stages_vec[labels(dend)]])

png("/home/cluster_user/Magisterka/plots/dendrogram_no_e_stage.png",
    width = 1600, height = 800, res = 150)

par(mar = c(10, 4, 4, 4))
plot(dend, main = "Hierarchical clustering (removed species effect, chosen stages)")
legend("topright", inset = c(0.05, 0), legend = names(stage_colors),
       fill = stage_colors, border = NA, bty = "n")

dev.off() 
```

```{r}
# #pheatmap stage
# pheno <- as.data.frame(colData(dds))
# design <- model.matrix(~ 0 + Stage2, data = pheno)
# 
# fit <- lmFit(residuals_mat, design)
# 
# #contrasts by stage2
# contrast_list <- lapply(combn(stage_levels, 2, simplify = FALSE), function(x) {
#   contrast <- rep(0, length(stage_levels))
#   names(contrast) <- stage_levels
#   contrast[x[1]] <- 1
#   contrast[x[2]] <- -1
#   contrast
# })
# 
# contrast_matrix <- do.call(cbind, contrast_list)
# colnames(contrast_matrix) <- sapply(combn(stage_levels, 2, simplify = FALSE), function(x) paste(x, collapse = "_vs_"))
# 
# fit2 <- contrasts.fit(fit, contrast_matrix)
# fit2 <- eBayes(fit2)
# 
# deg_list <- list()
# mat <- matrix(0, nrow = length(stage_levels), ncol = length(stage_levels),
#               dimnames = list(stage_levels, stage_levels))
# 
# for (comp_name in colnames(contrast_matrix)) {
#   res <- topTable(fit2, coef = comp_name, number = Inf, adjust.method = "BH", sort.by = "none")
#   deg_list[[comp_name]] <- res
#   comps <- strsplit(comp_name, "_vs_")[[1]]
#   
#   n_deg <- sum(res$adj.P.Val < 0.05, na.rm = TRUE)
#   mat[comps[1], comps[2]] <- n_deg
#   mat[comps[2], comps[1]] <- n_deg
# }
# 
# # Heatmapa
# ph <- pheatmap(mat,
#          fontsize_row = 10,
#          fontsize_col = 10,
#          angle_col = 45,
#          cellwidth = 20,
#          cellheight = 20,
#          border_color = NA,
#          main = "Number of DEGs between stages\n(removed species effect, chosen stages)",
#          color = deg_colors
# )
# 
# grid.newpage()
# grid.draw(ph$gtable)
# 
# ggsave("/home/cluster_user/Magisterka/plots/pheatmap_no_e_stage.png", ph, width = 5, height = 4.5)
```

### pseudotime

```{r}
print(percentVar)
print(sum(percentVar[0:6]))

#from pca matrix, first 6 main components (64.4% variance explained)
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:6)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

pcaData$Stage2 <- factor(pcaData$Stage2, levels = stages_order)
pcaData$Stage2 <- droplevels(pcaData$Stage2)

cor_test <- cor(as.numeric(factor(pcaData$Stage2, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")

print(cor_test)
#make sure pseudotime goes from eggs to adults and not from adult to eggs
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

p1 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime by stage - PCA\n(removed species effect, chosen stages)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_no_stage.png",
       p1, width = 5, height = 5, dpi = 300)

species_levels <- sort(c("blattella", "drosophila", "harmonia", "locusta", "plutella"))
pcaData$Species <- factor(pcaData$Species, levels = species_levels)
species_colors <- setNames(
  RColorBrewer::brewer.pal(length(species_levels), "Set1"),
  species_levels
)

p2 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - PCA\n(removed species effect, chosen stages)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_no_stage.png",
       p2, width = 7, height = 5, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1)) +
  plot_annotation(
    title = "Pseudotime distribution - PCA (removed species effect, chosen stages)",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_no_combined.png",
       combined, width = 11, height = 5, dpi = 300)


pseudotime_summary <- pcaData %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )

pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pseudotime trajectory (removed species effect, chosen stages) - PCA",
    y = "Median pseudotime",
    x = "Developmental stage"
  )

print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_no_stage.png",
       p3, width = 11, height = 8, dpi = 300)
```

#### pca

```{r}
curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:6)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(x = PC1, y = PC2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression PCA (removed species effect, chosen stages)",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_no_stage.png",
        p_curve, width = 11, height = 8, dpi = 300)
```

#### umap

```{r}
umap_mat_all <- as.matrix(umap_df[, c("UMAP1", "UMAP2")])
rownames(umap_mat_all) <- umap_df$Sample2

#match curve on umap
umap_curve_all <- principal_curve(umap_mat_all)

#count pseudotime
lambda_all <- umap_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))
umap_df$Pseudotime <- lambda_norm_all

#check pseudotime direction
stages_order <- c("Larva", "Nymph", "Prepupa", "Pupa")
umap_df$Stage2 <- factor(umap_df$Stage2, levels = stages_order)
umap_df$Stage2 <- droplevels(umap_df$Stage2)

cor_test <- cor(as.numeric(factor(umap_df$Stage2, levels = stages_order)),
                umap_df$Pseudotime, use = "complete.obs")
print(cor_test)

#swap pseudotime if needed
if (cor_test < 0) {
  umap_df$Pseudotime <- 1 - umap_df$Pseudotime
}

#plots

#pseudotime vs stage
p1 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime by stage - UMAP\n(removed species effect, chosen stages)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_umap_no_stage.png", p1, width = 5, height = 5, dpi = 300)

#pseudotime vs stage+species
p2 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - UMAP\n(removed species effect, chosen stages)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_umap_no_stage.png", p2, width = 7, height = 5, dpi = 300)

#pseudotime summary line plot
pseudotime_summary <- umap_df %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )
pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "   Median pseudotime trajectory (removed species effect, chosen stages) - UMAP",
    y = "Median pseudotime",
    x = "Developmental stage"
  )
print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_umap_no_stage.png", p3, width = 11, height = 8, dpi = 300)

#trajectory curve
curve_line <- as.data.frame(umap_curve_all$s)
colnames(curve_line) <- c("UMAP1", "UMAP2")
curve_line <- curve_line[order(umap_df$Pseudotime), ]

#umap plot
p_umap_curve <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(
    data = curve_line,
    aes(x = UMAP1, y = UMAP2),
    inherit.aes = FALSE,
    color = "black",
    size = 1.2,
    alpha = 0.7
  ) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression UMAP (removed species effect, chosen stages)",
    x = "UMAP1",
    y = "UMAP2",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_umap_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_no_stage.png",
       p_umap_curve, width = 11, height = 8, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1)) +
  plot_annotation(
    title = "Pseudotime distribution - UMAP (removed species effect, chosen stages)",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_no_combined.png",
       combined, width = 11, height = 5, dpi = 300)
```
