---
title: "6. all species gene expression analysis"
format:
  html:
    embed-resources: true
    toc: true
    toc_float: true
    toc-depth: 6
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## libs

```{r}
#| echo: true
#| eval: true
#| results: false
library(DESeq2)
library(ggplot2)
library(tidyr)
#library(umap)
library(uwot) 
library(pheatmap)
library(ggrepel)
library(readr)
library(stringr)
library(dplyr)
library(tibble)
library(Biostrings)
library(Mfuzz)
library(Biobase)
library(limma)
library(grid)
library(princurve)
library(patchwork)
library(sva)
library(vegan)
library(viridis)
library(lme4)
library(rlang)
library(emmeans)

conflicted::conflict_prefer("rename", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("intersect", "dplyr")
conflicted::conflict_prefer("unname", "base")
conflicted::conflict_prefer("umap", "uwot")
conflicted::conflict_prefer("exprs", "Biobase")
conflicted::conflict_prefer("setdiff", "dplyr")
```

## matrix

read previously prepared counts and orthogroups files

```{r}
locusta_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/locusta_counts.rds")
harmonia_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/harmonia_counts.rds")
plutella_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/plutella_counts.rds")
drosophila_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila_counts.rds")
blattella_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/blattella_counts.rds")
drosophila2_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila2_counts.rds")
```

```{r}
orthogroups <- read_tsv("/home/cluster_user/Magisterka/proteins/OrthoFinder/Results_Jun25/Orthogroups/Orthogroups.tsv")

single_copy <- read_lines("/home/cluster_user/Magisterka/proteins/OrthoFinder/Results_Jun25/Orthogroups/Orthogroups_SingleCopyOrthologues.txt")

#information about genes in orthogroups
head(orthogroups)

#single-copy orthogroups
head(single_copy)
```

### orthofinder

making matrix with SRR IDs as columns and orthogroups as rows, so for each gene with counts I sum them into orthogroup

```{r}
#| echo: true
#| eval: true
#| results: false
#list of dataframes with counts (per RNA-seq dataset)
all_counts <- list(
  locusta = locusta_counts,
  harmonia = harmonia_counts,
  plutella = plutella_counts,
  drosophila = drosophila_counts,
  blattella = blattella_counts,
  drosophila2 = drosophila2_counts
)

#combine all counts tables into one long table
#rows: all gene names; columns: all SRR and species; values: counts 
all_counts_long <- lapply(names(all_counts), function(species) {
  df <- all_counts[[species]]
  df$gene_id <- rownames(df)
  df$species <- species
  df
}) %>%
  bind_rows()

head(all_counts_long)

```

```{r}
#| echo: true
#| eval: true
#| results: false
#table: orthogroup, species, gene
orthogroup_mapping <- orthogroups %>%
  pivot_longer(-Orthogroup, names_to = "species", values_to = "genes") %>%
  filter(!is.na(genes)) %>%
  separate_rows(genes, sep = ",") %>%
  #clean gene names
  mutate(
    genes = str_trim(genes),
    species = str_replace(species, "_proteins", ""),
    genes = str_replace(genes, "\\..*", "") 
  )

#duplicate mapping for the second drosophila dataset
orthogroup_mapping_drosophila2 <- orthogroup_mapping %>%
  filter(species == "drosophila") %>%
  mutate(species = "drosophila2")

orthogroup_mapping <- bind_rows(orthogroup_mapping, orthogroup_mapping_drosophila2)

orthogroup_mapping %>%
  dplyr::count(species)

#numbers of genes in orthogroups (based on genomes)
# blattella	22002			
# drosophila	11435			
# drosophila2	11435			
# harmonia	18095			
# locusta	90809			
# plutella	23332			

head(orthogroup_mapping)


#join count data with orthogroup mapping
merged <- all_counts_long %>%
  inner_join(
    orthogroup_mapping,
    by = c("species", "gene_id" = "genes")
  )

head(merged)
merged %>%
  dplyr::count(species) %>%
  print()

#numbers of genes after mapping counts to orthogroups
#(should be equal or less than in the original genome annotation)
# blattella	19505			
# drosophila	11435			
# drosophila2	11394			
# harmonia	14846			
# locusta	65328			
# plutella	23143
```

### orthosnap

```{r}
#orthosnap -> found additional 1:1 orthogroups

snap_files <- list.files("/home/cluster_user/Magisterka/proteins/OrthoFinder/Results_Jun25/output_snap_ogs/", pattern = "\\.fa$", full.names = TRUE)

#read files and extract gene IDs
snap_orthologs <- lapply(snap_files, function(file) {
  fasta <- readAAStringSet(file)
  genes <- names(fasta)
  
  tibble(
    orthogroup = str_remove(basename(file), "\\.fa.orthosnap.0.fa$"),
    gene_id = str_replace(genes, "^.*?_", "") %>%
          str_replace("\\..*", "")
  )
}) %>% bind_rows()

#merge with expression counts
snap_expr <- snap_orthologs %>%
  inner_join(all_counts_long, by = "gene_id")

#sum gene counts into orthogroups
orthosnap_counts <- snap_expr %>%
  select(-gene_id, -species) %>%
  group_by(orthogroup) %>%
  summarise(across(everything(), ~sum(.x, na.rm = TRUE))) %>%
  column_to_rownames(var = "orthogroup")

#rows = orthogroups, columns = SRR samples
orthosnap_counts %>% dim()

#numbers of SRR samples per species
srr_per_species <- sapply(all_counts, ncol)
print(srr_per_species)

#total number of SRR samples across all datadets
total_srr <- sum(srr_per_species)
print(total_srr)

```

### 1:1 orthogroups

```{r}
#| echo: true
#| eval: true
#| results: false
#gene expression counts for all orthogroups (not only 1:1)
orthogroup_counts <- merged %>%
  select(Orthogroup, where(is.numeric)) %>%
  group_by(Orthogroup) %>%
  summarise(across(everything(), ~sum(.x, na.rm = TRUE)))

head(orthogroup_counts)

orthogroup_counts <- orthogroup_counts %>%
  column_to_rownames(var = "Orthogroup")

head(orthogroup_counts)
dim(orthogroup_counts)

colnames(orthogroup_counts)
```

```{r}
#narrow to only 1:1 orthogroups
orthogroup_counts_1to1 <- orthogroup_counts %>%
  filter(rownames(.) %in% single_copy)

#check if anything overlaps between 1:1 from orthofinder and orthosnap (no)
overlapping_genes <- intersect(rownames(orthogroup_counts_1to1), rownames(orthosnap_counts))

length(overlapping_genes)
print(overlapping_genes)

#add orthosnap
combined_counts <- bind_rows(
  orthogroup_counts_1to1,
  orthosnap_counts[!rownames(orthosnap_counts) %in% rownames(orthogroup_counts_1to1), ]
)

combined_counts[is.na(combined_counts)] <- 0
```

```{r}
#number of 1:1 orthogroups
cat("orthogroup_counts_1to1: ", dim(orthogroup_counts_1to1)[1], "orthogroups ×", dim(orthogroup_counts_1to1)[2], "SRRs\n")
cat("orthosnap_counts: ", dim(orthosnap_counts)[1], "orthogroups ×", dim(orthosnap_counts)[2], "SRRs\n")
cat("combined_counts: ", dim(combined_counts)[1], "orthogroups ×", dim(combined_counts)[2], "SRRs\n")

```

## deseq2

```{r}
#read metadata: Stage, Replicate, Sample.name, Species, Stage2, Sample2
#_2 are generalized names across all datasets / across replicates
metadata <- read.csv(file="/home/cluster_user/Magisterka/metadata2.csv", row.names = 1, header = TRUE, sep = ",") 

metadata <- metadata[colnames(orthogroup_counts), ]

dim(metadata)
head(metadata)
```

```{r}
# dds <- DESeqDataSetFromMatrix(countData = combat_seq_counts,
#                               colData = metadata,
#                               design = ~ 1)

dds <- DESeqDataSetFromMatrix(countData = combined_counts,
                              colData = metadata,
                              design = ~ Stage2 + Species)

#fit the model and estimate differential expression
dds <- DESeq(dds)

head(results(dds, tidy=TRUE))
summary(results(dds))

saveRDS(dds, file = "/home/cluster_user/link_to_datadisk/copy/counts/dds.rds")
```

### pca

```{r}
#variance-stabilizing transformation (reduces dependence of variance on mean expression)
vst <- varianceStabilizingTransformation(dds, blind = FALSE)
vst_mat <- assay(vst)

#get metadata
sample_info <- as.data.frame(colData(vst))
sample_info$Sample <- rownames(sample_info)
sample_info <- sample_info %>% select(Sample, Sample2, Stage2, Species)

#rows: samples; columns: genes
vst_df <- as.data.frame(t(vst_mat))
vst_df$Sample <- rownames(vst_df)
#add metadata
vst_df <- left_join(vst_df, sample_info, by = "Sample")

#median on replicates (before pca)
vst_median <- vst_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

#prepare for pca
expr_median <- vst_median %>% select(-Stage2, -Species)
samples <- vst_median$Sample2
expr_values <- vst_median %>% select(-Sample2, -Stage2, -Species)

exprs_mat <- t(as.matrix(expr_values))
colnames(exprs_mat) <- vst_median$Sample2
rownames(exprs_mat) <- colnames(expr_values)
exprs_mat <- t(exprs_mat)

#ExpressionSet object
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(exprs_mat),
                                       row.names = colnames(exprs_mat)))
eset <- ExpressionSet(assayData = exprs_mat, phenoData = pheno)

#standardize (mean = 0, sd = 1)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#principal component analysis
pca <- prcomp(expr_scaled)
#explained variance
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)

#plot
pcaData <- as.data.frame(pca$x)
pcaData$Sample2 <- vst_median$Sample2
pcaData$Stage2 <- vst_median$Stage2
pcaData$Species <- vst_median$Species

p <- ggplot(pcaData, aes(PC1, PC2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000, show.legend = FALSE) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of gene expression") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_all_stage_species.png",
       p, width = 11, height = 8, dpi = 300)
```

### heatmap

```{r}
#annotations
annotation_df <- vst_median %>%
  select(Sample2, Stage2, Species) %>%
  rename(Stage = Stage2) %>% 
  column_to_rownames("Sample2")

species_labels <- c(
  "blattella"  = "B. germanica",
  "drosophila" = "D. melanogaster",
  "harmonia"   = "H. axyridis",
  "locusta"    = "L. migratoria",
  "plutella"   = "P. xylostella"
)

#change species names acording to map above
annotation_df$Species <- species_labels[ as.character(annotation_df$Species) ]
annotation_df$Species <- factor(annotation_df$Species, levels = unname(species_labels))

#stages colours the same as in pca
stages_order <- c("Adult Female", "Adult Male", "Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa")
annotation_df$Stage <- factor(annotation_df$Stage, levels = stages_order)
stage_colors <- setNames(scales::hue_pal()(length(stages_order)), stages_order)

#different colours for species
species_levels <- unique(annotation_df$Species)
species_colors <- setNames(
  RColorBrewer::brewer.pal(n = max(3, length(species_levels)), "Set1")[1:length(species_levels)], species_levels
)

annotation_colors <- list(
  Stage = stage_colors,
  Species = species_colors
)

heat_colors <- colorRampPalette(rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")))(100)

#heatmap: columns: samples; rows: orthogroups; values: standardised gene expression
ph <- pheatmap(t(expr_scaled),
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         annotation_col = annotation_df,
         annotation_colors = annotation_colors,
         show_rownames = FALSE,
         show_colnames = TRUE,
         fontsize_col = 8,
         fontsize = 12,
         border_color = NA,
         angle_col = 45,
         main = "Heatmap of gene expression",
         color = heat_colors
         )

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/pheatmap_all_samples.png", ph, width = 11, height = 8)
```

### DEGs heatmap

```{r}
#heatmap of differential expression between stages
unique_stages <- levels(dds$Stage2)

#matrix for storing DEG counts between stages
mat <- matrix(0, nrow = length(unique_stages), ncol = length(unique_stages),
              dimnames = list(unique_stages, unique_stages))

#run all pairwise comparisons of DEGs
pairwise_comparisons <- combn(unique_stages, 2, simplify = FALSE)
deg_list <- list()

for (comp in pairwise_comparisons) {
  group1 <- comp[1]
  group2 <- comp[2]

  res <- tryCatch({
    results(dds, contrast = c("Stage2", group1, group2))
  }, error = function(e) {
    message(paste("Contrast", group1, "vs", group2, "failed:", e$message))
    return(NULL)
  })

  if (!is.null(res)) {
    deg_list[[paste0(group1, "_vs_", group2)]] <- as.data.frame(res)
    n_deg <- sum(res$padj < 0.05, na.rm = TRUE)

    mat[group1, group2] <- n_deg
    mat[group2, group1] <- n_deg
  }
}

#heatmap
deg_colors <- colorRampPalette(RColorBrewer::brewer.pal(n = 9, name = "Oranges"))(100)

ph <- pheatmap(mat,
         fontsize_row = 10,
         fontsize_col = 10,
         angle_col = 45,
         cellwidth = 20,
         cellheight = 20,
         border_color = NA,
         main = "Number of DEGs between stages",
         color = deg_colors
)

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/pheatmap_all_stage_species.png", ph, width = 5, height = 4)
```

### umap

```{r}
#Uniform Manifold Approximation and Projection
#Mfuzz standardise (z-score on orthogroups)
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(vst_mat),
                                       row.names = colnames(vst_mat)))
eset <- ExpressionSet(assayData = vst_mat, phenoData = pheno)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#dataframe with matadata
expr_df <- as.data.frame(t(expr_scaled)) 
expr_df$Sample <- colnames(vst_mat) 
sample_info <- as.data.frame(colData(vst)) 
sample_info$Sample <- rownames(sample_info) 
expr_df <- left_join(expr_df, sample_info, by = "Sample") 

#medians
expr_median <- expr_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

#UMAP
expr_median_mat <- expr_median %>%
  ungroup() %>%
  select(-Sample2, -Stage2, -Species) %>%
  as.matrix()

#standardise
expr_median_mat <- t(scale(t(expr_median_mat)))

umap_result <- umap(expr_median_mat, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")
umap_df <- as.data.frame(umap_result)
colnames(umap_df) <- c("UMAP1", "UMAP2")
umap_df$Sample2 <- expr_median$Sample2
umap_df$Stage2 <- expr_median$Stage2
umap_df$Species <- expr_median$Species

#plot
p <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000, show.legend = FALSE) +
  xlab("UMAP1") +
  ylab("UMAP2") +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  ggtitle("UMAP of gene expression") +
  labs(color = "Stage", shape = "Species") +
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_all_stage_species.png",
       p, width = 11, height = 8, dpi = 300)
```

### pseudotime

```{r}
print(percentVar)
print(sum(percentVar[0:6]))

#from pca matrix, first 6 main components (69.4% variance explained)
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:6)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
#standardise
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

#check if pseudotime is according to the stages order
stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")

pcaData$Stage2 <- factor(pcaData$Stage2, levels = stages_order)
pcaData$Stage2 <- droplevels(pcaData$Stage2)

#check correlation with the stages order and flip if needed
cor_test <- cor(as.numeric(factor(pcaData$Stage2, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")
print(cor_test)
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

species_levels <- sort(c("blattella", "drosophila", "harmonia", "locusta", "plutella"))
pcaData$Species <- factor(pcaData$Species, levels = species_levels)
species_colors <- setNames(
  RColorBrewer::brewer.pal(length(species_levels), "Set1"),
  species_levels
)

#distribution of pseudotime values across developmental stages
p1 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_point(
    position = position_jitter(width = 0.3),
    aes(fill = Stage2),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Stage2),
    fun = median,
    geom = "point",
    shape = 95,
    size = 14,
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  scale_color_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    ) +
  labs(
    title = "Pseudotime by stage - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_all_stage_species.png",
       p1, width = 5, height = 5, dpi = 300)

#Pseudotime by stage and speciess across stages and species
p2 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  geom_point(
    position = position_jitterdodge(
      jitter.width = 0.3,
      dodge.width = 0.8
    ),
    aes(fill = Species),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Species, group = Species),
    fun = median,
    geom = "point",
    shape = 95,
    size = 12,
    position = position_dodge(width = 0.8),
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - PCA",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_all_stage_species.png",
       p2, width = 7, height = 5, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1.5)) +
  plot_annotation(
    title = "Pseudotime distribution - PCA",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_all_stage_species_combined.png",
       combined, width = 11, height = 5, dpi = 300)

pseudotime_summary <- pcaData %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )

pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

#median pseudotime trajectory showing developmental progression
p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pseudotime trajectory - PCA",
    y = "Median pseudotime",
    x = "Developmental stage"
  )

print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_all_stage_species.png",
       p3, width = 11, height = 8, dpi = 300)
```

#### stats

```{r}
pseudo_stats <- pcaData %>%
  group_by(Stage2) %>%
  summarise(
    n = n(),
    median = median(Pseudotime),
    mean = mean(Pseudotime),
    sd = sd(Pseudotime),
    min = min(Pseudotime),
    max = max(Pseudotime),
    Q1 = quantile(Pseudotime, 0.25),
    Q3 = quantile(Pseudotime, 0.75),
    .groups = "drop"
  )

print(pseudo_stats)

pseudo_species <- pcaData %>%
  group_by(Stage2, Species) %>%
  summarise(
    median = median(Pseudotime),
    min = min(Pseudotime),
    max = max(Pseudotime),
    .groups = "drop"
  )

print(pseudo_species)

sorted_samples <- pcaData %>%
  arrange(Pseudotime) %>%
  select(Sample2, Species, Stage2, Pseudotime)

print(head(sorted_samples, 10))
print(tail(sorted_samples, 10))

stages_by_variation <- pcaData %>%
  group_by(Stage2) %>%
  summarise(
    range = max(Pseudotime) - min(Pseudotime),
    sd = sd(Pseudotime)
  ) %>%
  arrange(desc(range))

print(stages_by_variation)
```

#### pca

```{r}
curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:6)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(x = PC1, y = PC2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000, show.legend = FALSE) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression PCA",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_stage_species.png",
        p_curve, width = 11, height = 8, dpi = 300)
```

#### umap

```{r}
#same workflow as above, for UMAP data
umap_mat_all <- as.matrix(umap_df[, c("UMAP1", "UMAP2")])
rownames(umap_mat_all) <- umap_df$Sample2

#match curve on umap
umap_curve_all <- principal_curve(umap_mat_all)

#count pseudotime
lambda_all <- umap_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))
umap_df$Pseudotime <- lambda_norm_all

#check pseudotime direction
stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")
umap_df$Stage2 <- factor(umap_df$Stage2, levels = stages_order)
umap_df$Stage2 <- droplevels(umap_df$Stage2)

cor_test <- cor(as.numeric(factor(umap_df$Stage2, levels = stages_order)),
                umap_df$Pseudotime, use = "complete.obs")
print(cor_test)

#swap pseudotime if needed
if (cor_test < 0) {
  umap_df$Pseudotime <- 1 - umap_df$Pseudotime
}

p1 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_point(
    position = position_jitter(width = 0.3),
    aes(fill = Stage2),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Stage2),
    fun = median,
    geom = "point",
    shape = 95,
    size = 14,
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  scale_color_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime by stage - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_umap_stage_species.png", p1, width = 5, height = 5, dpi = 300)

p2 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  geom_point(
    position = position_jitterdodge(
      jitter.width = 0.3,
      dodge.width = 0.8
    ),
    aes(fill = Species),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Species, group = Species),
    fun = median,
    geom = "point",
    shape = 95,
    size = 12,
    position = position_dodge(width = 0.8),
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - UMAP",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_umap_stage_species.png", p2, width = 7, height = 5, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1.5)) +
  plot_annotation(
    title = "Pseudotime distribution - UMAP",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_all_stage_species_combined.png",
       combined, width = 11, height = 5, dpi = 300)

pseudotime_summary <- umap_df %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )
pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pseudotime trajectory - UMAP",
    y = "Median pseudotime",
    x = "Developmental stage"
  )
print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_umap_stage_species.png", p3, width = 11, height = 8, dpi = 300)

#trajectory curve
curve_line <- as.data.frame(umap_curve_all$s)
colnames(curve_line) <- c("UMAP1", "UMAP2")
curve_line <- curve_line[order(umap_df$Pseudotime), ]

p_umap_curve <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000, show.legend = FALSE) +
  geom_path(
    data = curve_line,
    aes(x = UMAP1, y = UMAP2),
    inherit.aes = FALSE,
    color = "black",
    size = 1.2,
    alpha = 0.7
  ) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression UMAP",
    x = "UMAP1",
    y = "UMAP2",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_umap_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_stage_species.png",
       p_umap_curve, width = 11, height = 8, dpi = 300)
```

## no species effect

### matrix

```{r}
#| echo: true
#| eval: true
#| results: false

drosophila_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila_joined_counts.rds")

#list of dataframes with counts (per RNA-seq dataset)
all_counts <- list(
  locusta = locusta_counts,
  harmonia = harmonia_counts,
  plutella = plutella_counts,
  drosophila = drosophila_counts,
  blattella = blattella_counts
)

#combine all counts tables into one long table
#rows: all gene names; columns: all SRR and species; values: counts 
all_counts_long <- lapply(names(all_counts), function(species) {
  df <- as.data.frame(all_counts[[species]])
  df$gene_id <- rownames(df)
  df$species <- species
  df
}) %>%
  bind_rows()

head(all_counts_long)

#table: orthogroup, species, gene
orthogroup_mapping <- orthogroups %>%
  pivot_longer(-Orthogroup, names_to = "species", values_to = "genes") %>%
  filter(!is.na(genes)) %>%
  separate_rows(genes, sep = ",") %>%
  #clean gene names
  mutate(
    genes = str_trim(genes),
    species = str_replace(species, "_proteins", ""),
    genes = str_replace(genes, "\\..*", "") 
  )

#duplicate mapping for the second drosophila dataset
orthogroup_mapping_drosophila2 <- orthogroup_mapping %>%
  filter(species == "drosophila") %>%
  mutate(species = "drosophila2")

orthogroup_mapping <- bind_rows(orthogroup_mapping, orthogroup_mapping_drosophila2)

orthogroup_mapping %>%
  dplyr::count(species)

head(orthogroup_mapping)


#join count data with orthogroup mapping
merged <- all_counts_long %>%
  inner_join(
    orthogroup_mapping,
    by = c("species", "gene_id" = "genes")
  )

head(merged)
merged %>%
  dplyr::count(species) %>%
  print()

orthogroup_counts <- merged %>%
  select(Orthogroup, where(is.numeric)) %>%
  group_by(Orthogroup) %>%
  summarise(across(everything(), ~sum(.x, na.rm = TRUE)))

head(orthogroup_counts)

orthogroup_counts <- orthogroup_counts %>%
  column_to_rownames(var = "Orthogroup")

head(orthogroup_counts)
dim(orthogroup_counts)

colnames(orthogroup_counts)

#narrow to only 1:1 orthogroups
orthogroup_counts_1to1 <- orthogroup_counts %>%
  filter(rownames(.) %in% single_copy)

#check if anything overlaps between 1:1 from orthofinder and orthosnap (no)
overlapping_genes <- intersect(rownames(orthogroup_counts_1to1), rownames(orthosnap_counts))

length(overlapping_genes)
print(overlapping_genes)

#add orthosnap
combined_counts <- bind_rows(
  orthogroup_counts_1to1,
  orthosnap_counts[!rownames(orthosnap_counts) %in% rownames(orthogroup_counts_1to1), ]
)

combined_counts[is.na(combined_counts)] <- 0
```

### deseq2

```{r}
metadata <- read.csv(file="/home/cluster_user/Magisterka/metadata2.csv", row.names = 1, header = TRUE, sep = ",") 

metadata <- metadata[colnames(orthogroup_counts), ]

dds <- DESeqDataSetFromMatrix(
  countData = combined_counts,
  colData   = metadata,
  design    = ~ Species + Stage2
)

dds <- estimateSizeFactors(dds, type = "poscounts")

#dds <- DESeq(dds) without this step as I'll do only visualisation as i have modified counts for drosophila

saveRDS(dds, file = "/home/cluster_user/link_to_datadisk/copy/counts/dds_batch.rds")
```

### pca

```{r}
#removing species effect by doing pca on residuas from model with species factor
# VST
vst <- varianceStabilizingTransformation(dds, blind = FALSE)
vst_mat <- assay(vst)

#linear model to remove the species effect (applied for each orthogroup separately)
residuals_mat <- apply(vst_mat, 1, function(x) {
  lm(x ~ colData(vst)$Species)$residuals
})
residuals_mat <- t(residuals_mat)

# residuals_mat <- removeBatchEffect(
#     vst_mat,
#     batch = colData(vst)$Species
# )

# residuals_mat <- ComBat(dat = vst_mat,
#                         batch = colData(vst)$Species)

#add metadata
sample_info <- as.data.frame(colData(vst))
sample_info$Sample <- rownames(sample_info)
sample_info <- sample_info %>% select(Sample, Sample2, Stage2, Species)

#data.frame residuals + Sample
residuals_df <- as.data.frame(t(residuals_mat))
residuals_df$Sample <- rownames(residuals_df)
residuals_df <- left_join(residuals_df, sample_info, by = "Sample")

#median for each Sample2/Stage2
residuals_median <- residuals_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

#make matrix
expr_median <- residuals_median %>% select(-Stage2, -Species)
samples <- expr_median$Sample2
expr_values <- expr_median %>% select(-Sample2)
exprs_mat <- t(as.matrix(expr_values))
colnames(exprs_mat) <- expr_median$Sample2
rownames(exprs_mat) <- colnames(expr_values)
exprs_mat <- t(exprs_mat)

#standardize Mfuzz -> z-score
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(exprs_mat), row.names = colnames(exprs_mat)))
eset <- ExpressionSet(assayData = exprs_mat, phenoData = pheno)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#PCA
pca <- prcomp(expr_scaled)
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)
pcaData <- as.data.frame(pca$x)
pcaData$Sample2 <- residuals_median$Sample2
pcaData$Stage2 <- residuals_median$Stage2
pcaData$Species <- residuals_median$Species

p <- ggplot(pcaData, aes(PC1, PC2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000, show.legend = FALSE) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of gene expression (removed species effect)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_all_stage.png",
       p, width = 11, height = 8, dpi = 300)

```

### heatmap

```{r}
#annotations
annotation_df <- residuals_median %>%
  select(Sample2, Stage2, Species) %>%
  rename(Stage = Stage2) %>% 
  column_to_rownames("Sample2")

#change species names to full
annotation_df$Species <- species_labels[ as.character(annotation_df$Species) ]
annotation_df$Species <- factor(annotation_df$Species, levels = unname(species_labels))

#stages colours the same as in pca
stages_order <- c("Adult Female", "Adult Male", "Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa")
annotation_df$Stage <- factor(annotation_df$Stage, levels = stages_order)
stage_colors <- setNames(scales::hue_pal()(length(stages_order)), stages_order)

#different colours for species
species_levels <- unique(annotation_df$Species)
species_colors <- setNames(
  RColorBrewer::brewer.pal(n = max(3, length(species_levels)), "Set1")[1:length(species_levels)],
  species_levels
)

annotation_colors <- list(
  Stage = stage_colors,
  Species = species_colors
)

ph <- pheatmap(t(expr_scaled),
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         annotation_col = annotation_df,
         annotation_colors = annotation_colors,
         show_rownames = FALSE,
         show_colnames = TRUE,
         fontsize_col = 8,
         fontsize = 12,
         border_color = NA,
         angle_col = 45,
         main = "Heatmap of gene expression (removed species effect)",
         color = heat_colors
         )

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/heatmap_all_stage.png", ph, width = 12, height = 8)
```

### umap

```{r}
#Mfuzz standardise (z-score on orthogroups)
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(residuals_mat),
                                       row.names = colnames(residuals_mat)))
eset <- ExpressionSet(assayData = residuals_mat, phenoData = pheno)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#dataframe with matadata
expr_df <- as.data.frame(t(expr_scaled)) 
expr_df$Sample <- colnames(vst_mat) 
sample_info <- as.data.frame(colData(vst)) 
sample_info$Sample <- rownames(sample_info) 
expr_df <- left_join(expr_df, sample_info, by = "Sample") 

#medians
expr_median <- expr_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

#UMAP
expr_median_mat <- expr_median %>%
  ungroup() %>%
  select(-Sample2, -Stage2, -Species) %>%
  as.matrix()

#standardise
expr_median_mat <- t(scale(t(expr_median_mat)))

umap_result <- umap(expr_median_mat, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")
umap_df <- as.data.frame(umap_result)
colnames(umap_df) <- c("UMAP1", "UMAP2")
umap_df$Sample2 <- expr_median$Sample2
umap_df$Stage2 <- expr_median$Stage2
umap_df$Species <- expr_median$Species

#plot
p <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 4.5, max.overlaps = 1000, show.legend = FALSE) +
  xlab("UMAP1") +
  ylab("UMAP2") +
  ggtitle("UMAP of gene expression (removed species effect)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_all_stage.png",
       p, width = 11, height = 8, dpi = 300)
```

```{r}
#plot for poster
p <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 4.5) +
  geom_text_repel(size = 5.5, max.overlaps = 1000, show.legend = FALSE) +
  xlab("UMAP1") +
  ylab("UMAP2") +
  ggtitle("UMAP of gene expression (removed species effect)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
    axis.title = element_text(size = 19, face = "bold"),
    axis.text = element_text(size = 18),
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 19, face = "bold")
  )+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_all_stage_big.png",
       p, width = 11, height = 8, dpi = 300)
```

### cluster

```{r}
expr_values <- expr_median %>%
  ungroup() %>%
  select(-Sample2, -Stage2) %>%
  as.matrix()

rownames(expr_values) <- expr_median$Sample2

dist_mat <- dist(expr_values, method = "euclidean")

hc <- hclust(dist_mat, method = "average")

#maps Sample2 -> Stage2
labels_df <- expr_median %>% select(Sample2, Stage2)
stages_vec <- setNames(labels_df$Stage2, labels_df$Sample2)

stage_levels <- sort(unique(labels_df$Stage2))

dend <- as.dendrogram(hc)

dend <- dend %>%
  dendextend::set("labels_colors", value = stage_colors[stages_vec[labels(dend)]])

png("/home/cluster_user/Magisterka/plots/dendrogram_all_stage.png",
    width = 1600, height = 800, res = 150)

par(mar = c(10, 4, 4, 4))
plot(dend, main = "Hierarchical clustering (removed species effect)")
legend("topright", inset = c(0.05, 0), legend = names(stage_colors),
       fill = stage_colors, border = NA, bty = "n")

dev.off() 
```

```{r}
#so this is incorrect
# #pheatmap stage
# pheno <- as.data.frame(colData(dds))
# design <- model.matrix(~ 0 + Stage2, data = pheno)
# 
# fit <- lmFit(residuals_mat, design)
# 
# #contrasts by stage2
# contrast_list <- lapply(combn(stage_levels, 2, simplify = FALSE), function(x) {
#   contrast <- rep(0, length(stage_levels))
#   names(contrast) <- stage_levels
#   contrast[x[1]] <- 1
#   contrast[x[2]] <- -1
#   contrast
# })
# 
# contrast_matrix <- do.call(cbind, contrast_list)
# colnames(contrast_matrix) <- sapply(combn(stage_levels, 2, simplify = FALSE), function(x) paste(x, collapse = "_vs_"))
# 
# fit2 <- contrasts.fit(fit, contrast_matrix)
# fit2 <- eBayes(fit2)
# 
# deg_list <- list()
# mat <- matrix(0, nrow = length(stage_levels), ncol = length(stage_levels),
#               dimnames = list(stage_levels, stage_levels))
# 
# for (comp_name in colnames(contrast_matrix)) {
#   res <- topTable(fit2, coef = comp_name, number = Inf, adjust.method = "BH", sort.by = "none")
#   deg_list[[comp_name]] <- res
#   comps <- strsplit(comp_name, "_vs_")[[1]]
#   
#   n_deg <- sum(res$adj.P.Val < 0.05, na.rm = TRUE)
#   mat[comps[1], comps[2]] <- n_deg
#   mat[comps[2], comps[1]] <- n_deg
# }
# 
# # Heatmapa
# ph <- pheatmap(mat,
#          fontsize_row = 10,
#          fontsize_col = 10,
#          angle_col = 45,
#          cellwidth = 20,
#          cellheight = 20,
#          border_color = NA,
#          main = "Number of DEGs between stages\n(removed species effect)",
#          color = deg_colors
# )
# 
# grid.newpage()
# grid.draw(ph$gtable)
# 
# ggsave("/home/cluster_user/Magisterka/plots/pheatmap_all_stage.png", ph, width = 5, height = 4.5)
```

### pseudotime

```{r}
print(percentVar)
print(sum(percentVar[0:6]))

#from pca matrix, first 6 main components (59.5% variance explained)
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:6)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")

pcaData$Stage2 <- factor(pcaData$Stage2, levels = stages_order)
pcaData$Stage2 <- droplevels(pcaData$Stage2)

cor_test <- cor(as.numeric(factor(pcaData$Stage2, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")

print(cor_test)
#make sure pseudotime goes from eggs to adults and not from adult to eggs
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

p1 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_point(
    position = position_jitter(width = 0.3),
    aes(fill = Stage2),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Stage2),
    fun = median,
    geom = "point",
    shape = 95,
    size = 14,
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  scale_color_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime by stage - PCA\n(removed species effect)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_all_stage.png",
       p1, width = 5, height = 5, dpi = 300)

species_levels <- sort(c("blattella", "drosophila", "harmonia", "locusta", "plutella"))
pcaData$Species <- factor(pcaData$Species, levels = species_levels)
species_colors <- setNames(
  RColorBrewer::brewer.pal(length(species_levels), "Set1"),
  species_levels
)

p2 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  geom_point(
    position = position_jitterdodge(
      jitter.width = 0.3,
      dodge.width = 0.8
    ),
    aes(fill = Species),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Species, group = Species),
    fun = median,
    geom = "point",
    shape = 95,
    size = 12,
    position = position_dodge(width = 0.8),
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - PCA\n(removed species effect)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_all_stage.png",
       p2, width = 7, height = 5, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1.5)) +
  plot_annotation(
    title = "Pseudotime distribution - PCA (removed species effect)",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_all_stage_combined.png",
       combined, width = 11, height = 5, dpi = 300)

pseudotime_summary <- pcaData %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )

pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = species_colors,
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pseudotime trajectory (removed species effect) - PCA",
    y = "Median pseudotime",
    x = "Developmental stage"
  )

print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_all_stage.png",
       p3, width = 11, height = 8, dpi = 300)
```

poster

```{r}
#---poster plot
p1 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_point(
    position = position_jitter(width = 0.3),
    aes(fill = Stage2),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Stage2),
    fun = median,
    geom = "point",
    shape = 95,
    size = 14,
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  scale_color_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 17, face = "bold"),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 15)
    ) +
  labs(
    title = "Pseudotime by stage - PCA\n(removed species effect)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_all_stage_big.png",
       p1, width = 5, height = 5, dpi = 300)
```

#### stats

```{r}
pseudo_stats <- pcaData %>%
  group_by(Stage2) %>%
  summarise(
    n = n(),
    median = median(Pseudotime),
    mean = mean(Pseudotime),
    sd = sd(Pseudotime),
    min = min(Pseudotime),
    max = max(Pseudotime),
    Q1 = quantile(Pseudotime, 0.25),
    Q3 = quantile(Pseudotime, 0.75),
    .groups = "drop"
  )

print(pseudo_stats)

pseudo_species <- pcaData %>%
  group_by(Stage2, Species) %>%
  summarise(
    median = median(Pseudotime),
    min = min(Pseudotime),
    max = max(Pseudotime),
    .groups = "drop"
  )

print(pseudo_species)

sorted_samples <- pcaData %>%
  arrange(Pseudotime) %>%
  select(Sample2, Species, Stage2, Pseudotime)

print(head(sorted_samples, 10))
print(tail(sorted_samples, 10))

stages_by_variation <- pcaData %>%
  group_by(Stage2) %>%
  summarise(
    range = max(Pseudotime) - min(Pseudotime),
    sd = sd(Pseudotime)
  ) %>%
  arrange(desc(range))

print(stages_by_variation)
```

#### pca

```{r}
curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:6)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(x = PC1, y = PC2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000, show.legend = FALSE) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression PCA (removed species effect)",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_all_stage.png",
        p_curve, width = 11, height = 8, dpi = 300)
```

#### umap

```{r}
umap_mat_all <- as.matrix(umap_df[, c("UMAP1", "UMAP2")])
rownames(umap_mat_all) <- umap_df$Sample2

#match curve on umap
umap_curve_all <- principal_curve(umap_mat_all)

#count pseudotime
lambda_all <- umap_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))
umap_df$Pseudotime <- lambda_norm_all

#check pseudotime direction
stages_order <- c("Eggs", "Embryo", "Larva", "Nymph", "Prepupa", "Pupa", "Adult Male", "Adult Female")
umap_df$Stage2 <- factor(umap_df$Stage2, levels = stages_order)
umap_df$Stage2 <- droplevels(umap_df$Stage2)

cor_test <- cor(as.numeric(factor(umap_df$Stage2, levels = stages_order)),
                umap_df$Pseudotime, use = "complete.obs")
print(cor_test)

#swap pseudotime if needed
if (cor_test < 0) {
  umap_df$Pseudotime <- 1 - umap_df$Pseudotime
}

#plots

#pseudotime vs stage
p1 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_point(
    position = position_jitter(width = 0.3),
    aes(fill = Stage2),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Stage2),
    fun = median,
    geom = "point",
    shape = 95,
    size = 14,
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  scale_color_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime by stage - UMAP\n(removed species effect)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_umap_all_stage.png", p1, width = 5, height = 5, dpi = 300)

#pseudotime vs stage+species
p2 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  geom_point(
    position = position_jitterdodge(
      jitter.width = 0.3,
      dodge.width = 0.8
    ),
    aes(fill = Species),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Species, group = Species),
    fun = median,
    geom = "point",
    shape = 95,
    size = 12,
    position = position_dodge(width = 0.8),
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - UMAP\n(removed species effect)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_umap_all_stage.png", p2, width = 7, height = 5, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1.5)) +
  plot_annotation(
    title = "Pseudotime distribution - UMAP (removed species effect)",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_all_stage_combined.png",
       combined, width = 11, height = 5, dpi = 300)

#pseudotime summary line plot
pseudotime_summary <- umap_df %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )
pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pseudotime trajectory (removed species effect) - UMAP",
    y = "Median pseudotime",
    x = "Developmental stage"
  )
print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_umap_all_stage.png", p3, width = 11, height = 8, dpi = 300)

#trajectory curve
curve_line <- as.data.frame(umap_curve_all$s)
colnames(curve_line) <- c("UMAP1", "UMAP2")
curve_line <- curve_line[order(umap_df$Pseudotime), ]

#umap plot
p_umap_curve <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000, show.legend = FALSE) +
  geom_path(
    data = curve_line,
    aes(x = UMAP1, y = UMAP2),
    inherit.aes = FALSE,
    color = "black",
    size = 1.2,
    alpha = 0.7
  ) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression UMAP (removed species effect)",
    x = "UMAP1",
    y = "UMAP2",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_umap_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_all_stage.png",
       p_umap_curve, width = 11, height = 8, dpi = 300)
```

## statistical tests

```{r}
# Convert DESeq2 colData to data.frame
sample_info <- as.data.frame(colData(vst))
sample_info$Sample <- rownames(sample_info)

# Keep all relevant columns including instar (Stage)
sample_info <- sample_info %>%
  select(Sample, Sample2, Stage, Stage2, Species)

# Transpose residuals matrix to data.frame
residuals_df <- as.data.frame(t(residuals_mat))
residuals_df$Sample <- rownames(residuals_df)

# Join metadata
residuals_df <- left_join(residuals_df, sample_info, by = "Sample")

residuals_median_instars <- residuals_df %>%
  group_by(Sample2, Stage, Stage2, Species) %>% 
  summarise(across(where(is.numeric), median), .groups = "drop")

expr_dist <- residuals_median_instars %>%
  select(-Stage, -Stage2, -Species) %>%
  column_to_rownames("Sample2") %>%
  as.matrix()

#Calculate Spearman correlation distance
dist_mat <- 1 - cor(t(expr_dist), method = "spearman")

# Save sample names
samples <- rownames(dist_mat)

# Convert distance matrix to long-format data.frame
# the data is non independent!!! (distance of samples pairs)
dist_df <- as.data.frame(as.table(dist_mat))
colnames(dist_df) <- c("Sample1", "Sample2", "distance")

# Add metadata for both samples
meta <- residuals_median_instars %>%
  select(Sample2, Stage, Stage2, Species) %>%
  distinct()

# Join metadata for Sample1
dist_df <- dist_df %>%
  left_join(meta, by = c("Sample1" = "Sample2")) %>%
  rename(Stage1 = Stage,
         Stage2_1 = Stage2,
         Species1 = Species)

# Join metadata for Sample2
dist_df <- dist_df %>%
  left_join(meta, by = c("Sample2" = "Sample2")) %>%
  rename(Stage2 = Stage,
         Stage2_2 = Stage2,
         Species2 = Species)

# Remove self-comparisons
dist_df <- dist_df %>% filter(Sample1 != Sample2)

#. Create broader developmental categories (Meta-stages)
meta_stage <- function(stage) {
  case_when(
    grepl("^Larva", stage) ~ "Larva",
    grepl("^Nymph", stage) ~ "Nymph",
    grepl("^Prepupa", stage) ~ "Prepupa",
    grepl("^Pupa", stage) ~ "Pupa",
    grepl("^Embryo|Egg", stage) ~ "Embryo",
    grepl("^Adult", stage) ~ "Adult",
    TRUE ~ NA_character_
  )
}

dist_df <- dist_df %>%
  mutate(
    instar_number = case_when(
      grepl("Nymph", Stage1) ~ as.numeric(gsub(".*?(\\d+).*", "\\1", Stage1)),
      grepl("instar", Stage1) ~ as.numeric(gsub(".*?(\\d+).*", "\\1", Stage1)),
      TRUE ~ NA_real_
    )
  )

#  Compare nymphs to Larva, Prepupa, Pupa
nymph_comp <- dist_df %>%
  filter(Stage2_1 == "Nymph" & Stage2_2 %in% c("Larva", "Prepupa", "Pupa"))

# Boxplot - overview of pairwise distances
p <- ggplot(nymph_comp, aes(x = Stage2_2, y = distance, fill = Stage2_2)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, color = "black", size = 1.5) +
  scale_fill_manual(values = stage_colors) +
  theme_classic(base_size = 14) +
  labs(
    x = "Developmental stage",
    y = "Spearman distance to nymph",
    title = "Nymphal distances to holometabolic stages"
  ) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"))

print(p)
ggsave("/home/cluster_user/Magisterka/plots/spearman_dist_violin.png",
       p, width = 7, height = 5, dpi = 300)
```

#### early vs late

##### prepupa

```{r}
#Early vs late nymphal instars; locusta 1,2,3,4,5 blattella: 1,3,5,6
# dist_df <- dist_df %>%
#   mutate(
#     NymphType1 = case_when(
#       instar_number %in% 1:3 ~ "Early",
#       instar_number %in% 4:6 ~ "Late",
#       TRUE ~ NA_character_
#     ),
#     NymphType1 = factor(NymphType1, levels = c("Early", "Late"))
#   )
# 
# nymph_trend <- dist_df %>%
#   filter(
#     Stage2_1 == "Nymph",
#     Stage2_2 == "Prepupa",
#     !is.na(NymphType1)
#   )
# 
# p <- ggplot(nymph_trend, aes(x = NymphType1, y = distance)) +
#   geom_violin(trim = FALSE, alpha = 0.7) +
#   geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
#   geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
#   theme_classic(base_size = 14) +
#   labs(
#     x = "Nymphal instar category",
#     y = "Spearman distance to prepupa",
#     title = "Early vs late nymph distances to prepupa",
#     color = "Nymph\nsample"
#   ) +
#   theme(legend.position = "right")
# 
# print(p)
# ggsave("/home/cluster_user/Magisterka/plots/early_vs_late_prepupa.png",
#        p, width = 6, height = 5, dpi = 300)
# 
# model_early_late <- lmer(distance ~ NymphType1 + (1 | Sample2),
#                          data = nymph_trend)
# summary(model_early_late)
# emmeans(model_early_late, ~ NymphType1)

#for more extreme stages
dist_df <- dist_df %>%
  mutate(
    NymphType1_12_56 = case_when(
      instar_number %in% c(1, 2) ~ "Early (1-2)",
      instar_number %in% c(5, 6) ~ "Late (5-6)",
      TRUE ~ NA_character_
    )
  )

nymph_trend_12_56 <- dist_df %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 == "Prepupa",
    !is.na(NymphType1_12_56)
  )

p <- ggplot(nymph_trend_12_56, aes(x = NymphType1_12_56, y = distance)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymphal instar category",
    y = "Spearman distance to prepupa",
    title = "Early vs late nymph distances to prepupa",
    color = "Nymph\nsample"
  ) +
  theme(legend.position = "right")

print(p)
ggsave("/home/cluster_user/Magisterka/plots/very_early_vs_late_prepupa.png",
       p, width = 6, height = 5, dpi = 300)

model_extreme <- lmer(distance ~ NymphType1_12_56 + (1 | Sample2),
                         data = nymph_trend_12_56)
summary(model_extreme)
emmeans(model_extreme, ~ NymphType1_12_56)

emm <- emmeans(model_extreme, ~ NymphType1_12_56)
summary(pairs(emm))

#late nymphs are closer to prepupa
```

##### larva

```{r}
# nymph_trend <- dist_df %>%
#   filter(
#     Stage2_1 == "Nymph",
#     Stage2_2 == "Larva",
#     !is.na(NymphType1)
#   )
# 
# p <- ggplot(nymph_trend, aes(x = NymphType1, y = distance)) +
#   geom_violin(trim = FALSE, alpha = 0.7) +
#   geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
#   geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
#   theme_classic(base_size = 14) +
#   labs(
#     x = "Nymphal instar category",
#     y = "Spearman distance to larva",
#     title = "Early vs late nymph distances to larva",
#     color = "Nymph\nsample"
#   ) +
#   theme(legend.position = "right")
# 
# print(p)
# ggsave("/home/cluster_user/Magisterka/plots/early_vs_late_larva.png",
#        p, width = 6, height = 5, dpi = 300)
# 
# model_early_late <- lmer(distance ~ NymphType1 + (1 | Sample2),
#                          data = nymph_trend)
# summary(model_early_late)
# emmeans(model_early_late, ~ NymphType1)
# 
#for more extreme stages
dist_df <- dist_df %>%
  mutate(
    NymphType1_12_56 = case_when(
      instar_number %in% c(1, 2) ~ "Early (1-2)",
      instar_number %in% c(5, 6) ~ "Late (5-6)",
      TRUE ~ NA_character_
    )
  )

nymph_trend_12_56 <- dist_df %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 == "Larva",
    !is.na(NymphType1_12_56)
  )

p <- ggplot(nymph_trend_12_56, aes(x = NymphType1_12_56, y = distance)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymphal instar category",
    y = "Spearman distance to larva",
    title = "Early vs late nymph distances to larva",
    color = "Nymph\nsample"
  ) +
  theme(legend.position = "right")

print(p)
ggsave("/home/cluster_user/Magisterka/plots/very_early_vs_late_larva.png",
       p, width = 6, height = 5, dpi = 300)

model_extreme <- lmer(distance ~ NymphType1_12_56 + (1 | Sample2),
                         data = nymph_trend_12_56)
summary(model_extreme)
emmeans(model_extreme, ~ NymphType1_12_56)

emm <- emmeans(model_extreme, ~ NymphType1_12_56)
summary(pairs(emm))

# Linear mixed model fit by REML ['lmerMod']
# Formula: distance ~ NymphType1_12_56 + (1 | Sample2)
#    Data: nymph_trend_12_56
# 
# REML criterion at convergence: -40
# 
# Scaled residuals: 
#     Min      1Q  Median      3Q     Max 
# -1.8791 -0.7604 -0.1407  0.8021  2.0038 
# 
# Random effects:
#  Groups   Name        Variance Std.Dev.
#  Sample2  (Intercept) 0.001878 0.04334 
#  Residual             0.031271 0.17684 
# Number of obs: 84, groups:  Sample2, 14
# 
# Fixed effects:
#                            Estimate Std. Error t value
# (Intercept)                 0.92007    0.02964  31.038
# NymphType1_12_56Late (5-6)  0.10233    0.03859   2.652
# 
# Correlation of Fixed Effects:
#             (Intr)
# NT1_12_56L( -0.651
#  NymphType1_12_56 emmean     SE   df lower.CL upper.CL
#  Early (1-2)        0.92 0.0296 35.5    0.860     0.98
#  Late (5-6)         1.02 0.0296 35.5    0.962     1.08
# 
# Degrees-of-freedom method: kenward-roger 
# Confidence level used: 0.95 
#  contrast                     estimate     SE df t.ratio p.value
#  (Early (1-2)) - (Late (5-6))   -0.102 0.0386 69  -2.652  0.0099
# 
# Degrees-of-freedom method: kenward-roger 

#larva are closer to early nymphs
```

##### pupa

```{r}
# nymph_trend <- dist_df %>%
#   filter(
#     Stage2_1 == "Nymph",
#     Stage2_2 == "Pupa",
#     !is.na(NymphType1)
#   )
# 
# p <- ggplot(nymph_trend, aes(x = NymphType1, y = distance)) +
#   geom_violin(trim = FALSE, alpha = 0.7) +
#   geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
#   geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
#   theme_classic(base_size = 14) +
#   labs(
#     x = "Nymphal instar category",
#     y = "Spearman distance to pupa",
#     title = "Early vs late nymph distances to pupa",
#     color = "Nymph\nsample"
#   ) +
#   theme(legend.position = "right")
# 
# print(p)
# ggsave("/home/cluster_user/Magisterka/plots/early_vs_late_pupa.png",
#        p, width = 6, height = 5, dpi = 300)
# 
# model_early_late <- lmer(distance ~ NymphType1 + (1 | Sample2),
#                          data = nymph_trend)
# summary(model_early_late)
# emmeans(model_early_late, ~ NymphType1)

#for more extreme stages
dist_df <- dist_df %>%
  mutate(
    NymphType1_12_56 = case_when(
      instar_number %in% c(1, 2) ~ "Early (1-2)",
      instar_number %in% c(5, 6) ~ "Late (5-6)",
      TRUE ~ NA_character_
    )
  )

nymph_trend_12_56 <- dist_df %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 == "Pupa",
    !is.na(NymphType1_12_56)
  )

p <- ggplot(nymph_trend_12_56, aes(x = NymphType1_12_56, y = distance)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6) +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymphal instar category",
    y = "Spearman distance to pupa",
    title = "Early vs late nymph distances to pupa",
    color = "Nymph\nsample"
  ) +
  theme(legend.position = "right")

print(p)
ggsave("/home/cluster_user/Magisterka/plots/very_early_vs_late_pupa.png",
       p, width = 6, height = 5, dpi = 300)

model_extreme <- lmer(distance ~ NymphType1_12_56 + (1 | Sample2),
                         data = nymph_trend_12_56)
summary(model_extreme)
emmeans(model_extreme, ~ NymphType1_12_56)

emm <- emmeans(model_extreme, ~ NymphType1_12_56)
summary(pairs(emm))

#connection to nymphal instars isnt significant
```

##### combined plot

```{r}
combined_nymph <- dist_df %>%
  mutate(
    NymphType = case_when(
      instar_number %in% c(1, 2) ~ "Early (1–2)",
      instar_number %in% c(5, 6) ~ "Late (5–6)",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 %in% c("Larva", "Prepupa", "Pupa"),
    !is.na(NymphType)
  )

combined_nymph$Stage2_2 <- factor(
  combined_nymph$Stage2_2,
  levels = c("Larva", "Prepupa", "Pupa")
)

p_combined <- ggplot(combined_nymph,
                     aes(x = NymphType, y = distance)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6, size = 1.2) +
  facet_wrap(~ Stage2_2, nrow = 1, scales = "free_y") +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymphal instar category",
    y = "Spearman distance",
    title = "Comparison of holometabolic stages distances to early and late nymphal instars",
    color = "Nymph"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 15, face = "bold"),
    strip.text = element_text(face = "bold", size = 12),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 12, face = "bold")
  )

print(p_combined)
ggsave("/home/cluster_user/Magisterka/plots/very_early_vs_late_combined.png",
       p_combined, width = 12, height = 5, dpi = 300)

#---poster plot

p_combined <- ggplot(combined_nymph,
                     aes(x = NymphType, y = distance)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(aes(color = Sample1), width = 0.15, alpha = 0.6, size = 3) +
  facet_wrap(~ Stage2_2, nrow = 1, scales = "free_y") +
  theme_classic(base_size = 18) +
  labs(
    x = "Nymphal instar category",
    y = "Spearman distance",
    title = "Comparison of holometabolic stages distances\nto early and late nymphal instars",
    color = "Nymph"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
    strip.text = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 17),
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 17, face = "bold")
  )

print(p_combined)
ggsave("/home/cluster_user/Magisterka/plots/very_early_vs_late_combined_big.png",
       p_combined, width = 12, height = 5, dpi = 300)
```

#### instars

```{r}
trend_df <- dist_df %>%
  filter(
    Stage2_1 == "Nymph",
    Stage2_2 %in% c("Larva", "Prepupa", "Pupa"),
    !is.na(instar_number)
  )

cor_results <- trend_df %>%
  group_by(Stage2_2) %>%
  summarise(
    rho = cor(instar_number, distance, method = "spearman"),
    n = n(),
    .groups = "drop"
  )

print(cor_results)

#counted separate for each metastage
#   Stage2_2      rho     n
# 1 Larva     0.242     126 --> distance growths in next stages)
# 2 Prepupa  -0.314     117 --> distance falls in next stages
# 3 Pupa     -0.00894    99 --> only very small decrease, not visible

# subset colors only for used stages
stage_colors_sub <- stage_colors[c("Larva", "Prepupa", "Pupa")]

# correlation labels
cor_labels <- cor_results %>%
  mutate(label = paste0("rho = ", round(rho, 2)))

p <- ggplot(trend_df, aes(instar_number, distance, color = Stage2_2)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  facet_wrap(~ Stage2_2, nrow = 1, scales = "free_y") +
  geom_text(
    data = cor_labels,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.1, vjust = 1.1,
    size = 4
  ) +
  scale_color_manual(values = stage_colors_sub) +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymphal instar number",
    y = "Spearman distance",
    title = "Similarity to nymphal instars"
  ) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 13, face = "bold"),
    plot.title = element_text(face = "bold")
  )

print(p)
ggsave("/home/cluster_user/Magisterka/plots/spearman_dist_instar.png",
       p, width = 7, height = 5, dpi = 300)

p_sample <- ggplot(trend_df, aes(instar_number, distance)) +
  geom_point(aes(color = Sample1), alpha = 0.6, size = 2) +
  geom_smooth(
    method = "lm", se = TRUE, linewidth = 1.2
  ) +
  facet_wrap(~ Stage2_2, nrow = 1, scales = "free_y") +
  theme_classic(base_size = 14) +
  labs(
    x = "Nymphal instar number",
    y = "Spearman distance",
    title = "Similarity to nymphal instars",
    color = "Nymph\nsample"
  ) +
  theme(
    legend.position = "right",
    strip.text = element_text(size = 13, face = "bold"),
    plot.title = element_text(face = "bold")
  )

print(p_sample)

ggsave("/home/cluster_user/Magisterka/plots/spearman_dist_instar_sample.png",
       p_sample, width = 9, height = 5, dpi = 300)
```

#### models

```{r}
lm_model <- lmer(distance ~ instar_number * Stage2_2 + (1 | Sample2), data = trend_df)
summary(lm_model)

#(1 | Sample2) -> intercept for Sample2 (not parameter for each sample) -> accounts for replications of samples
#(1 | Sample1) -> has variance close to 0, so i removed it
#Species2 controls for variation between species ->> no, because it was already from residua then ---> MODEL1
#lmer linear mixed-effects model  - fixed effects + random effects (sample related)

# Linear mixed model fit by REML ['lmerMod']
# Formula: distance ~ instar_number * Stage2_2 + (1 | Sample2)
#    Data: trend_df
# 
# REML criterion at convergence: -224.1
# 
# Scaled residuals: 
#      Min       1Q   Median       3Q      Max 
# -2.10157 -0.76390  0.01936  0.75001  2.24429 
# 
# Random effects:
#  Groups   Name        Variance Std.Dev.
#  Sample2  (Intercept) 0.001551 0.03938 
#  Residual             0.026168 0.16177 
# Number of obs: 342, groups:  Sample2, 38
# 
# Fixed effects:
#                                Estimate Std. Error t value
# (Intercept)                    0.903741   0.033425  27.038
# instar_number                  0.025250   0.008479   2.978
# Stage2_2Prepupa                0.204722   0.048170   4.250
# Stage2_2Pupa                   0.035013   0.050390   0.695
# instar_number:Stage2_2Prepupa -0.054838   0.012219  -4.488
# instar_number:Stage2_2Pupa    -0.025376   0.012782  -1.985
# 
# Correlation of Fixed Effects:
#                     (Intr) instr_ Stg2_2Pr Stg2_2Pp instr_nmbr:Stg2_2Pr
# instar_nmbr         -0.846                                             
# Stag2_2Prpp         -0.694  0.587                                      
# Stage2_2Pup         -0.663  0.561  0.460                               
# instr_nmbr:Stg2_2Pr  0.587 -0.694 -0.846   -0.389                      
# instr_nmbr:Stg2_2Pp  0.561 -0.663 -0.389   -0.846    0.460 

#  contrast        estimate     SE  df t.ratio p.value
#  Larva - Prepupa   0.0548 0.0122 301   4.488 <0.0001
#  Larva - Pupa      0.0254 0.0128 301   1.985  0.1176
#  Prepupa - Pupa   -0.0295 0.0130 301  -2.267  0.0621
# 
# Degrees-of-freedom method: kenward-roger 
# P value adjustment: tukey method for comparing a family of 3 estimates 

#  Stage2_2 instar_number.trend      SE  df lower.CL upper.CL t.ratio p.value
#  Larva               0.025250 0.00848 301  0.00856   0.0419   2.978  0.0031
#  Prepupa            -0.029588 0.00880 301 -0.04690  -0.0123  -3.363  0.0009
#  Pupa               -0.000126 0.00957 301 -0.01895   0.0187  -0.013  0.9895
# 
# Degrees-of-freedom method: kenward-roger 
# Confidence level used: 0.95 

emm <- emtrends(lm_model, ~ Stage2_2, var = "instar_number")
print(summary(emm, infer = c(TRUE, TRUE)))
contrast_results <- pairs(emm)
print(contrast_results)

trend_emm <- as.data.frame(emtrends(lm_model, ~ Stage2_2, var = "instar_number"))

p_trend <- ggplot(trend_emm, aes(x = Stage2_2, y = instar_number.trend, color = Stage2_2)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  scale_color_manual(values = stage_colors[c("Larva", "Prepupa", "Pupa")]) +
  theme_classic(base_size = 14) +
  labs(
    x = "Developmental stage",
    y = "Change in distance per instar",
    title = "Changes in distances across nymphal instars"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )

print(p_trend)

ggsave("/home/cluster_user/Magisterka/plots/instar_trends_emmeans.png",
       p_trend, width = 7, height = 5, dpi = 300)
```

#### embryo

```{r}
#compare embryo from hemi and holo separately
hemi_species <- c("locusta", "blattella")
holo_species <- c("harmonia", "plutella", "drosophila")

dist_df <- dist_df %>%
  mutate(
    EmbryoType = case_when(
      Stage2_2 == "Embryo" & Species2 %in% hemi_species ~ "Embryo_hemi",
      Stage2_2 == "Embryo" & Species2 %in% holo_species ~ "Embryo_holo",
      TRUE ~ Stage2_2
    )
  )

embryo_test <- dist_df %>%
  filter(
    Stage2_1 == "Larva",
    EmbryoType %in% c("Embryo_hemi", "Embryo_holo", "Nymph", "Prepupa", "Pupa")
  )

model <- lmer(distance ~ EmbryoType + (1 | Sample2), data = embryo_test)
summary(model)

# Linear mixed model fit by REML ['lmerMod']
# Formula: distance ~ EmbryoType + (1 | Sample2)
#    Data: embryo_test
# 
# REML criterion at convergence: -406
# 
# Scaled residuals: 
#      Min       1Q   Median       3Q      Max 
# -2.97903 -0.71763 -0.02411  0.70653  2.56588 
# 
# Random effects:
#  Groups   Name        Variance Std.Dev.
#  Sample2  (Intercept) 0.003112 0.05578 
#  Residual             0.027266 0.16512 
# Number of obs: 616, groups:  Sample2, 44
# 
# Fixed effects:
#                       Estimate Std. Error t value
# (Intercept)            1.04049    0.03181  32.710
# EmbryoTypeEmbryo_holo  0.01637    0.04307   0.380
# EmbryoTypeNymph       -0.05258    0.03967  -1.325
# EmbryoTypePrepupa      0.16838    0.03743   4.499
# EmbryoTypePupa         0.12515    0.03836   3.262
# 
# Correlation of Fixed Effects:
#             (Intr) EmbTE_ EmbrTN EmbryTypPr
# EmbryTypEm_ -0.739                         
# EmbryTypNym -0.802  0.592                  
# EmbryTypPrp -0.850  0.628  0.681           
# EmbryoTypPp -0.829  0.612  0.665  0.705    
#  EmbryoType  emmean     SE df lower.CL upper.CL t.ratio p.value
#  Embryo_hemi  1.040 0.0318 39    0.976     1.10  32.710 <0.0001
#  Embryo_holo  1.057 0.0290 39    0.998     1.12  36.396 <0.0001
#  Nymph        0.988 0.0237 39    0.940     1.04  41.668 <0.0001
#  Prepupa      1.209 0.0197 39    1.169     1.25  61.279 <0.0001
#  Pupa         1.166 0.0214 39    1.122     1.21  54.353 <0.0001
# 
# Degrees-of-freedom method: kenward-roger 
# Confidence level used: 0.95 
#  contrast                  estimate     SE df t.ratio p.value
#  Embryo_hemi - Embryo_holo  -0.0164 0.0431 39  -0.380  0.9954
#  Embryo_hemi - Nymph         0.0526 0.0397 39   1.325  0.6773
    #  Embryo_hemi - Prepupa      -0.1684 0.0374 39  -4.499  0.0005
    #  Embryo_hemi - Pupa         -0.1251 0.0384 39  -3.262  0.0184
#  Embryo_holo - Nymph         0.0690 0.0375 39   1.839  0.3664
    #  Embryo_holo - Prepupa      -0.1520 0.0351 39  -4.330  0.0009
    #  Embryo_holo - Pupa         -0.1088 0.0361 39  -3.013  0.0345
    #  Nymph - Prepupa            -0.2210 0.0308 39  -7.164 <0.0001
    #  Nymph - Pupa               -0.1777 0.0320 39  -5.559 <0.0001
#  Prepupa - Pupa              0.0432 0.0291 39   1.484  0.5788
# 
# Degrees-of-freedom method: kenward-roger 
# P value adjustment: tukey method for comparing a family of 5 estimates 


#larva is 0.closer to nymph than to embryo --> Berlese-
#also further to pre/pupa than to embryo 
#closest to early nymph (connected to previous test) -> Hinton+
#and it's okay that larva are supposed to be higly specialised embryo in Berlese so they don't need to be super close, but then it shouldn't be closer to nymphs
 
#it's not significant that nymph is closer to larva than embryo but still Berlese needs that embryo is closer -> they are similar 

#so it's still more Hinton where the transfers between stages are more fluid

#PLOTS
embryo_test$EmbryoType <- factor(
  embryo_test$EmbryoType,
  levels = c("Embryo_hemi", "Embryo_holo", "Nymph", "Prepupa", "Pupa")
)

embryo_test$EmbryoType <- recode(
  embryo_test$EmbryoType,
  "Embryo_hemi" = "Embryo (hemi)",
  "Embryo_holo" = "Embryo (holo)"
)

stage_colors_extended <- c(
  "Embryo (hemi)" = "#009E5A",   # darker green
  "Embryo (holo)" = "#33CC80",   # lighter green
  "Nymph" = "#00A9FF",
  "Prepupa" = "#C77CFF",
  "Pupa" = "#FF61CC"
)

p <- ggplot(embryo_test, aes(x = EmbryoType, y = distance, fill = EmbryoType)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.5) +
  scale_fill_manual(values = stage_colors_extended) +
  theme_classic(base_size = 14) +
  labs(
    x = "Developmental stage",
    y = "Spearman distance to larva",
    title = "Larval distance to developmental stages"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )

print(p)
ggsave("/home/cluster_user/Magisterka/plots/spearman_dist_embryo.png",
       p, width = 8, height = 5, dpi = 300)

#emmeans
emm <- emmeans(model, ~ EmbryoType)
print(summary(emm, infer = c(TRUE, TRUE)))
contrast_results <- pairs(emm)
print(contrast_results)

emm_df <- as.data.frame(emmeans(model, ~ EmbryoType))

# fix names
emm_df$EmbryoType <- recode(
  emm_df$EmbryoType,
  "Embryo_hemi" = "Embryo (hemi)",
  "Embryo_holo" = "Embryo (holo)"
)

emm_df$EmbryoType <- factor(
  emm_df$EmbryoType,
  levels = c("Embryo (hemi)", "Embryo (holo)", "Nymph", "Prepupa", "Pupa")
)

p_emm <- ggplot(emm_df, aes(x = EmbryoType, y = emmean, color = EmbryoType)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  scale_color_manual(values = stage_colors_extended) +
  theme_classic(base_size = 17.5) +
  labs(
    x = "Developmental stage",
    y = "Estimated Spearman distance",
    title = "Larval relationships across stages"
  ) +
  theme(
    legend.position = "none",
    #axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 13)
  )

print(p_emm)

ggsave("/home/cluster_user/Magisterka/plots/emmeans_larva.png",
       p_emm, width = 8, height = 5, dpi = 300)
```

```{r}
stage_colors_extended <- c(
  "Embryo\n(hemi)" = "#009E5A",   # darker green
  "Embryo\n(holo)" = "#33CC80",   # lighter green
  "Nymph" = "#00A9FF",
  "Prepupa" = "#C77CFF",
  "Pupa" = "#FF61CC"
)

emm <- emmeans(model, ~ EmbryoType)
print(summary(emm, infer = c(TRUE, TRUE)))
contrast_results <- pairs(emm)
print(contrast_results)

emm_df <- as.data.frame(emmeans(model, ~ EmbryoType))

emm_df$EmbryoType <- recode(
  emm_df$EmbryoType,
  "Embryo_hemi" = "Embryo\n(hemi)",
  "Embryo_holo" = "Embryo\n(holo)"
)

emm_df$EmbryoType <- factor(
  emm_df$EmbryoType,
  levels = c("Embryo\n(hemi)", "Embryo\n(holo)", "Nymph", "Prepupa", "Pupa")
)

#--poster plot
p_emm <- ggplot(emm_df, aes(x = EmbryoType, y = emmean, color = EmbryoType)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  scale_color_manual(values = stage_colors_extended) +
  theme_classic(base_size = 18) +
  labs(
    x = "Developmental stage",
    y = "Estimated Spearman distance",
    title = "Larval relationships across stages"
  ) +
  theme(
    legend.position = "none",
    #axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 17)
  )

print(p_emm)

ggsave("/home/cluster_user/Magisterka/plots/emmeans_larva_big.png",
       p_emm, width = 8, height = 5, dpi = 300)
```

### all

```{r}
# create separate dataframe for heatmap
heatmap_dist_df <- dist_df %>%
  mutate(
    Stage2_1 = case_when(
      Stage2_1 == "Embryo" & Species1 %in% hemi_species ~ "Embryo (hemi)",
      Stage2_1 == "Embryo" & Species1 %in% holo_species ~ "Embryo (holo)",
      TRUE ~ Stage2_1
    ),
    
    Stage2_2 = case_when(
      Stage2_2 == "Embryo" & Species2 %in% hemi_species ~ "Embryo (hemi)",
      Stage2_2 == "Embryo" & Species2 %in% holo_species ~ "Embryo (holo)",
      TRUE ~ Stage2_2
    )
  )

# calculate median distances between stage pairs
heatmap_df <- heatmap_dist_df %>%
  group_by(Stage2_1, Stage2_2) %>%
  summarise(
    median_distance = median(distance, na.rm = TRUE),
    .groups = "drop"
  )

# order stages
stage_order <- c(
  "Eggs",
  "Embryo (hemi)",
  "Embryo (holo)",
  "Nymph",
  "Larva",
  "Prepupa",
  "Pupa",
  "Adult Female",
  "Adult Male"
)

# keep only lower triangle
heatmap_df <- heatmap_df %>%
  mutate(
    x_id = match(Stage2_1, stage_order),
    y_id = match(Stage2_2, stage_order)
  ) %>%
  filter(y_id >= x_id)

heatmap_df$Stage2_1 <- factor(heatmap_df$Stage2_1, levels = stage_order)
heatmap_df$Stage2_2 <- factor(heatmap_df$Stage2_2, levels = rev(stage_order))

# heatmap
p_heatmap <- ggplot(
  heatmap_df,
  aes(x = Stage2_1, y = Stage2_2, fill = median_distance)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(median_distance, 2)), size = 4) +

  scale_fill_gradient2(
    low = "#2166AC",     # more similar
    mid = "white",       # neutral (distance = 1)
    high = "#B2182B",    # more divergent
    midpoint = 1,
    limits = c(0.5, 1.5),
    name = "Median\nSpearman\ndistance"
  ) +

  coord_fixed() +
  theme_classic(base_size = 14) +
  labs(
    x = "Stage 1",
    y = "Stage 2",
    title = "Transcriptomic distances\nbetween developmental stages"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
print(p_heatmap)

ggsave(
  "/home/cluster_user/Magisterka/plots/stage_distance_heatmap.png",
  p_heatmap,
  width = 7,
  height = 6,
  dpi = 300
)
```

### partial correlation

```{r}
cor_matrix <- cor(t(expr_dist), method = "spearman")

#partial correlation
inv_cor <- solve(cor_matrix)
pcor <- -cov2cor(inv_cor)

pcor_df <- as.data.frame(as.table(pcor))
colnames(pcor_df) <- c("Sample1", "Sample2", "distance")

# Join metadata for Sample1
pcor_df <- pcor_df %>%
  left_join(meta, by = c("Sample1" = "Sample2")) %>%
  rename(Stage1 = Stage,
         Stage2_1 = Stage2,
         Species1 = Species)

# Join metadata for Sample2
pcor_df <- pcor_df %>%
  left_join(meta, by = c("Sample2" = "Sample2")) %>%
  rename(Stage2 = Stage,
         Stage2_2 = Stage2,
         Species2 = Species)

# Remove self-comparisons
pcor_df <- pcor_df %>% filter(Sample1 != Sample2)

# divide embryo
pcor_df <- pcor_df %>%
  mutate(
    Stage2_1 = case_when(
      Stage2_1 == "Embryo" & Species1 %in% hemi_species ~ "Embryo (hemi)",
      Stage2_1 == "Embryo" & Species1 %in% holo_species ~ "Embryo (holo)",
      TRUE ~ Stage2_1
    ),
    
    Stage2_2 = case_when(
      Stage2_2 == "Embryo" & Species2 %in% hemi_species ~ "Embryo (hemi)",
      Stage2_2 == "Embryo" & Species2 %in% holo_species ~ "Embryo (holo)",
      TRUE ~ Stage2_2
    )
  )

# calculate median distances between stage pairs
pcor_df <- pcor_df %>%
  group_by(Stage2_1, Stage2_2) %>%
  summarise(
    median_distance = median(distance, na.rm = TRUE),
    .groups = "drop"
  )

# keep only lower triangle
pcor_df <- pcor_df %>%
  mutate(
    x_id = match(Stage2_1, stage_order),
    y_id = match(Stage2_2, stage_order)
  ) %>%
  filter(y_id >= x_id)

pcor_df$Stage2_1 <- factor(pcor_df$Stage2_1, levels = stage_order)
pcor_df$Stage2_2 <- factor(pcor_df$Stage2_2, levels = rev(stage_order))

# heatmap
p_heatmap <- ggplot(
  pcor_df,
  aes(x = Stage2_1, y = Stage2_2, fill = median_distance)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(median_distance, 2)), size = 4) +

  scale_fill_gradient2(
    low = "#2166AC",     # more similar
    mid = "white",       # neutral (distance = 1)
    high = "#B2182B",    # more divergent
    midpoint = 0,
    limits = c(-1, 1),
    name = "Median\npartial\ncorrelation"
  ) +

  coord_fixed() +
  theme_classic(base_size = 14) +
  labs(
    x = "Stage 1",
    y = "Stage 2",
    title = "Partial Spearman correlation\nbetween developmental stages"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
print(p_heatmap)

ggsave(
  "/home/cluster_user/Magisterka/plots/partial_heatmap.png",
  p_heatmap,
  width = 7,
  height = 6,
  dpi = 300
)
```

## no embryo&adults

```{r}
metadata_filtered <- metadata[!(metadata$Stage2 %in% c("Embryo", "Eggs","Adult Female", "Adult Male" )), ]
counts_filtered <- combined_counts[, rownames(metadata_filtered)]

dds <- DESeqDataSetFromMatrix(countData = counts_filtered,
                              colData = metadata_filtered,
                              design = ~ Stage2)

dds <- estimateSizeFactors(dds)
#dds <- DESeq(dds)
```

### pca

```{r}
vst <- varianceStabilizingTransformation(dds, blind = FALSE)
vst_mat <- assay(vst)

#linear model to delete the species effect
residuals_mat <- apply(vst_mat, 1, function(x) {
  lm(x ~ colData(vst)$Species)$residuals
})
residuals_mat <- t(residuals_mat)

# residuals_mat <- removeBatchEffect(
#     vst_mat,
#     batch = colData(vst)$Species,
#     design = model.matrix(~ colData(vst)$Stage2)
# )

# residuals_mat <- ComBat(dat = vst_mat,
#                         batch = colData(vst)$Species)

sample_info <- as.data.frame(colData(vst))
sample_info$Sample <- rownames(sample_info)
sample_info <- sample_info %>% select(Sample, Sample2, Stage2, Species)

#data.frame residuals + Sample
residuals_df <- as.data.frame(t(residuals_mat))
residuals_df$Sample <- rownames(residuals_df)
residuals_df <- left_join(residuals_df, sample_info, by = "Sample")

#median for each Sample2/Stage2
residuals_median <- residuals_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

expr_median <- residuals_median %>% select(-Stage2, -Species)
samples <- expr_median$Sample2
expr_values <- expr_median %>% select(-Sample2)
exprs_mat <- t(as.matrix(expr_values))
colnames(exprs_mat) <- expr_median$Sample2
rownames(exprs_mat) <- colnames(expr_values)
exprs_mat <- t(exprs_mat)

#standardize Mfuzz
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(exprs_mat), row.names = colnames(exprs_mat)))
eset <- ExpressionSet(assayData = exprs_mat, phenoData = pheno)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#PCA
pca <- prcomp(expr_scaled)
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)

pcaData <- as.data.frame(pca$x)
pcaData$Sample2 <- residuals_median$Sample2
pcaData$Stage2 <- residuals_median$Stage2
pcaData$Species <- residuals_median$Species

#stages colours the same as in previous pca
stages_order <- c("Larva", "Nymph", "Prepupa", "Pupa")
annotation_df$Stage <- factor(annotation_df$Stage, levels = stages_order)
stage_colors <- setNames(scales::hue_pal()(length(stages_order) + 4)[5:(length(stages_order) + 4)], stages_order)

p <- ggplot(pcaData, aes(PC1, PC2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000, show.legend = FALSE) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of gene expression (removed species effect, chosen stages)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  scale_color_manual(values = stage_colors)+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_no_e_stage.png",
       p, width = 11, height = 8, dpi = 300)
```

### heatmap

```{r}
#annotations
annotation_df <- residuals_median %>%
  select(Sample2, Stage2, Species) %>%
  rename(Stage = Stage2) %>% 
  column_to_rownames("Sample2")

#change species names to full
annotation_df$Species <- factor(species_labels[annotation_df$Species],
                                levels = species_labels)

#different colours for species
species_levels <- unique(annotation_df$Species)
species_colors <- setNames(
  RColorBrewer::brewer.pal(n = max(3, length(species_levels)), "Set1")[1:length(species_levels)],
  species_levels
)

annotation_colors <- list(
  Stage = stage_colors,
  Species = species_colors
)

ph <- pheatmap(t(expr_scaled),
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         annotation_col = annotation_df,
         annotation_colors = annotation_colors,
         show_rownames = FALSE,
         show_colnames = TRUE,
         fontsize_col = 8,
         fontsize = 12,
         border_color = NA,
         angle_col = 45,
         main = "Heatmap of gene expression (removed species effect, chosen stages)",
         color = heat_colors
         )

grid.newpage()
grid.draw(ph$gtable)

ggsave("/home/cluster_user/Magisterka/plots/heatmap_no_e_stage.png", ph, width = 12, height = 8)
```

### umap

```{r}
#Mfuzz standardise (z-score on orthogroups)
pheno <- AnnotatedDataFrame(data.frame(Sample = colnames(residuals_mat),
                                       row.names = colnames(residuals_mat)))
eset <- ExpressionSet(assayData = residuals_mat, phenoData = pheno)
eset_std <- standardise(eset)
expr_scaled <- exprs(eset_std)

#dataframe with matadata
expr_df <- as.data.frame(t(expr_scaled)) 
expr_df$Sample <- colnames(vst_mat) 
sample_info <- as.data.frame(colData(vst)) 
sample_info$Sample <- rownames(sample_info) 
expr_df <- left_join(expr_df, sample_info, by = "Sample") 

#medians
expr_median <- expr_df %>%
  group_by(Sample2, Stage2, Species) %>%
  summarise(across(where(is.numeric), median), .groups = "drop")

#UMAP
expr_median_mat <- expr_median %>%
  ungroup() %>%
  select(-Sample2, -Stage2, -Species) %>%
  as.matrix()

#standardise
expr_median_mat <- t(scale(t(expr_median_mat)))

umap_result <- umap(expr_median_mat, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")
umap_df <- as.data.frame(umap_result)
colnames(umap_df) <- c("UMAP1", "UMAP2")
umap_df$Sample2 <- expr_median$Sample2
umap_df$Stage2 <- expr_median$Stage2
umap_df$Species <- expr_median$Species

#plot
p <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000, show.legend = FALSE) +
  xlab("UMAP1") +
  ylab("UMAP2") +
  ggtitle("UMAP of gene expression (removed species effect, chosen stages)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  scale_color_manual(values = stage_colors)+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_no_e_stage.png",
       p, width = 11, height = 8, dpi = 300)
```

```{r}
#---- poster plot
p <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Stage2, shape = Species, label = Sample2)) +
  geom_point(size = 4.5) +
  geom_text_repel(size = 5.5, max.overlaps = 1000, show.legend = FALSE) +
  xlab("UMAP1") +
  ylab("UMAP2") +
  ggtitle("UMAP of gene expression\n(removed species effect, chosen stages)") +
  theme_minimal() +
  labs(color = "Stage", shape = "Species") +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
    axis.title = element_text(size = 19, face = "bold"),
    axis.text = element_text(size = 18),
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 19, face = "bold")
  )+
  scale_color_manual(values = stage_colors)+
  scale_shape_manual(
    values = c(
      "drosophila" = 16,
      "blattella" = 17,
      "plutella" = 15,
      "locusta" = 3,
      "harmonia" = 7
    ),
    labels = c(
      "drosophila" = expression(italic("D. melanogaster")),
      "blattella" = expression(italic("B. germanica")),
      "plutella" = expression(italic("P. xylostella")),
      "locusta" = expression(italic("L. migratoria")),
      "harmonia" = expression(italic("H. axyridis"))
    )
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/UMAP_no_e_stage_big.png",
       p, width = 11, height = 8, dpi = 300)
```

### cluster

```{r}
expr_values <- expr_median %>%
  ungroup() %>%
  select(-Sample2, -Stage2) %>%
  as.matrix()

rownames(expr_values) <- expr_median$Sample2

dist_mat <- dist(expr_values, method = "euclidean")

hc <- hclust(dist_mat, method = "average")

#maps Sample2 -> Stage2
labels_df <- expr_median %>% select(Sample2, Stage2)
stages_vec <- setNames(labels_df$Stage2, labels_df$Sample2)

stage_levels <- sort(unique(labels_df$Stage2))

dend <- as.dendrogram(hc)

dend <- dend %>%
  dendextend::set("labels_colors", value = stage_colors[stages_vec[labels(dend)]])

png("/home/cluster_user/Magisterka/plots/dendrogram_no_e_stage.png",
    width = 1600, height = 800, res = 150)

par(mar = c(10, 4, 4, 4))
plot(dend, main = "Hierarchical clustering (removed species effect, chosen stages)")
legend("topright", inset = c(0.05, 0), legend = names(stage_colors),
       fill = stage_colors, border = NA, bty = "n")

dev.off() 
```

```{r}
# #pheatmap stage
# pheno <- as.data.frame(colData(dds))
# design <- model.matrix(~ 0 + Stage2, data = pheno)
# 
# fit <- lmFit(residuals_mat, design)
# 
# #contrasts by stage2
# contrast_list <- lapply(combn(stage_levels, 2, simplify = FALSE), function(x) {
#   contrast <- rep(0, length(stage_levels))
#   names(contrast) <- stage_levels
#   contrast[x[1]] <- 1
#   contrast[x[2]] <- -1
#   contrast
# })
# 
# contrast_matrix <- do.call(cbind, contrast_list)
# colnames(contrast_matrix) <- sapply(combn(stage_levels, 2, simplify = FALSE), function(x) paste(x, collapse = "_vs_"))
# 
# fit2 <- contrasts.fit(fit, contrast_matrix)
# fit2 <- eBayes(fit2)
# 
# deg_list <- list()
# mat <- matrix(0, nrow = length(stage_levels), ncol = length(stage_levels),
#               dimnames = list(stage_levels, stage_levels))
# 
# for (comp_name in colnames(contrast_matrix)) {
#   res <- topTable(fit2, coef = comp_name, number = Inf, adjust.method = "BH", sort.by = "none")
#   deg_list[[comp_name]] <- res
#   comps <- strsplit(comp_name, "_vs_")[[1]]
#   
#   n_deg <- sum(res$adj.P.Val < 0.05, na.rm = TRUE)
#   mat[comps[1], comps[2]] <- n_deg
#   mat[comps[2], comps[1]] <- n_deg
# }
# 
# # Heatmapa
# ph <- pheatmap(mat,
#          fontsize_row = 10,
#          fontsize_col = 10,
#          angle_col = 45,
#          cellwidth = 20,
#          cellheight = 20,
#          border_color = NA,
#          main = "Number of DEGs between stages\n(removed species effect, chosen stages)",
#          color = deg_colors
# )
# 
# grid.newpage()
# grid.draw(ph$gtable)
# 
# ggsave("/home/cluster_user/Magisterka/plots/pheatmap_no_e_stage.png", ph, width = 5, height = 4.5)
```

### pseudotime

```{r}
print(percentVar)
print(sum(percentVar[0:6]))

#from pca matrix, first 6 main components (64.4% variance explained)
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:6)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

pcaData$Stage2 <- factor(pcaData$Stage2, levels = stages_order)
pcaData$Stage2 <- droplevels(pcaData$Stage2)

cor_test <- cor(as.numeric(factor(pcaData$Stage2, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")

print(cor_test)
#make sure pseudotime goes from eggs to adults and not from adult to eggs
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

p1 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, color = Stage2)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_point(
    position = position_jitter(width = 0.3),
    aes(fill = Stage2),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Stage2),
    fun = median,
    geom = "point",
    shape = 95,
    size = 14,
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  scale_color_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime by stage - PCA\n(removed species effect, chosen stages)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_no_stage.png",
       p1, width = 5, height = 5, dpi = 300)

species_levels <- sort(c("blattella", "drosophila", "harmonia", "locusta", "plutella"))
pcaData$Species <- factor(pcaData$Species, levels = species_levels)
species_colors <- setNames(
  RColorBrewer::brewer.pal(length(species_levels), "Set1"),
  species_levels
)

p2 <- ggplot(pcaData, aes(x = Stage2, y = Pseudotime, color = Species)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  geom_point(
    position = position_jitterdodge(
      jitter.width = 0.3,
      dodge.width = 0.8
    ),
    aes(fill = Species),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Species, group = Species),
    fun = median,
    geom = "point",
    shape = 95,
    size = 12,
    position = position_dodge(width = 0.8),
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - PCA\n(removed species effect, chosen stages)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )

print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_no_stage.png",
       p2, width = 7, height = 5, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1.5)) +
  plot_annotation(
    title = "Pseudotime distribution - PCA (removed species effect, chosen stages)",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_no_combined.png",
       combined, width = 11, height = 5, dpi = 300)


pseudotime_summary <- pcaData %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )

pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Median pseudotime trajectory (removed species effect, chosen stages) - PCA",
    y = "Median pseudotime",
    x = "Developmental stage"
  )

print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_no_stage.png",
       p3, width = 11, height = 8, dpi = 300)
```

#### stats

```{r}
pseudo_stats <- pcaData %>%
  group_by(Stage2) %>%
  summarise(
    n = n(),
    median = median(Pseudotime),
    mean = mean(Pseudotime),
    sd = sd(Pseudotime),
    min = min(Pseudotime),
    max = max(Pseudotime),
    Q1 = quantile(Pseudotime, 0.25),
    Q3 = quantile(Pseudotime, 0.75),
    .groups = "drop"
  )

print(pseudo_stats)

pseudo_species <- pcaData %>%
  group_by(Stage2, Species) %>%
  summarise(
    median = median(Pseudotime),
    min = min(Pseudotime),
    max = max(Pseudotime),
    .groups = "drop"
  )

print(pseudo_species)

sorted_samples <- pcaData %>%
  arrange(Pseudotime) %>%
  select(Sample2, Species, Stage2, Pseudotime)

print(head(sorted_samples, 10))
print(tail(sorted_samples, 10))

stages_by_variation <- pcaData %>%
  group_by(Stage2) %>%
  summarise(
    range = max(Pseudotime) - min(Pseudotime),
    sd = sd(Pseudotime)
  ) %>%
  arrange(desc(range))

print(stages_by_variation)
```

#### pca

```{r}
curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:6)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(x = PC1, y = PC2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000, show.legend = FALSE) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression PCA (removed species effect, chosen stages)",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_no_stage.png",
        p_curve, width = 11, height = 8, dpi = 300)
```

#### umap

```{r}
umap_mat_all <- as.matrix(umap_df[, c("UMAP1", "UMAP2")])
rownames(umap_mat_all) <- umap_df$Sample2

#match curve on umap
umap_curve_all <- principal_curve(umap_mat_all)

#count pseudotime
lambda_all <- umap_curve_all$lambda
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))
umap_df$Pseudotime <- lambda_norm_all

#check pseudotime direction
stages_order <- c("Larva", "Nymph", "Prepupa", "Pupa")
umap_df$Stage2 <- factor(umap_df$Stage2, levels = stages_order)
umap_df$Stage2 <- droplevels(umap_df$Stage2)

cor_test <- cor(as.numeric(factor(umap_df$Stage2, levels = stages_order)),
                umap_df$Pseudotime, use = "complete.obs")
print(cor_test)

#swap pseudotime if needed
if (cor_test < 0) {
  umap_df$Pseudotime <- 1 - umap_df$Pseudotime
}

#plots

#pseudotime vs stage
p1 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Stage2)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_point(
    position = position_jitter(width = 0.3),
    aes(fill = Stage2),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Stage2),
    fun = median,
    geom = "point",
    shape = 95,
    size = 14,
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    name = "Stage",
    values = stage_colors
  ) +
  scale_color_manual(
    name = "Stage",
    values = stage_colors
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
    ) +
  labs(
    title = "Pseudotime by stage - UMAP\n(removed species effect, chosen stages)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p1)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime1_umap_no_stage.png", p1, width = 5, height = 5, dpi = 300)

#pseudotime vs stage+species
p2 <- ggplot(umap_df, aes(x = Stage2, y = Pseudotime, fill = Species)) +
  #geom_boxplot(outlier.shape = NA, alpha = 0.8, position = position_dodge(0.8)) +
  geom_point(
    position = position_jitterdodge(
      jitter.width = 0.3,
      dodge.width = 0.8
    ),
    aes(fill = Species),
    shape = 21,
    color = "black",
    stroke = 0.3,
    size = 2.5,
    alpha = 0.8
  ) +
  stat_summary(
    aes(color = Species, group = Species),
    fun = median,
    geom = "point",
    shape = 95,
    size = 12,
    position = position_dodge(width = 0.8),
    alpha = 0.8
  ) +
  theme_minimal() +
  scale_fill_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime by stage and species - UMAP\n(removed species effect, chosen stages)",
    y = "Pseudotime (normalized)",
    x = "Developmental stage"
  )
print(p2)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime2_umap_no_stage.png", p2, width = 7, height = 5, dpi = 300)

#pseudotime summary line plot
pseudotime_summary <- umap_df %>%
  group_by(Species, Stage2) %>%
  summarise(
    median_pseudo = median(Pseudotime),
    sd_pseudo = sd(Pseudotime),
    .groups = "drop"
  )
pseudotime_summary$Stage2 <- factor(pseudotime_summary$Stage2, levels = stages_order)
pseudotime_summary$Stage2 <- droplevels(pseudotime_summary$Stage2)

p3 <- ggplot(pseudotime_summary, aes(x = Stage2, y = median_pseudo, group = Species, color = Species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_manual(
    values = RColorBrewer::brewer.pal(5, "Set1"),
    breaks = names(species_labels),
    labels = species_labels
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12, face = "italic"),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "   Median pseudotime trajectory (removed species effect, chosen stages) - UMAP",
    y = "Median pseudotime",
    x = "Developmental stage"
  )
print(p3)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime3_umap_no_stage.png", p3, width = 11, height = 8, dpi = 300)

#trajectory curve
curve_line <- as.data.frame(umap_curve_all$s)
colnames(curve_line) <- c("UMAP1", "UMAP2")
curve_line <- curve_line[order(umap_df$Pseudotime), ]

#umap plot
p_umap_curve <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Pseudotime, shape = Stage2, label = Sample2)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000, show.legend = FALSE) +
  geom_path(
    data = curve_line,
    aes(x = UMAP1, y = UMAP2),
    inherit.aes = FALSE,
    color = "black",
    size = 1.2,
    alpha = 0.7
  ) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = "Pseudotime on gene expression UMAP (removed species effect, chosen stages)",
    x = "UMAP1",
    y = "UMAP2",
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_umap_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_no_stage.png",
       p_umap_curve, width = 11, height = 8, dpi = 300)

#p1: pseudotime vs stage
p1_clean <- p1 +
  labs(title = "Stages")
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#p2: stage+species
p2_clean <- p2 +
  labs(title = "Stages × species") + 
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title.y = element_blank()
  )

#combined p1 and p2
combined <- p1_clean + p2_clean +
  plot_layout(ncol = 2, widths = c(1, 1.5)) +
  plot_annotation(
    title = "Pseudotime distribution - UMAP (removed species effect, chosen stages)",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

print(combined)
ggsave("/home/cluster_user/Magisterka/plots/pseudotime_umap_no_combined.png",
       combined, width = 11, height = 5, dpi = 300)
```
