---
title: "3. analysis of separate datasets"
format:
  html:
    embed-resources: true
    toc: true
    toc_float: true
    toc-depth: 6
editor: visual
author: "Weronika Oprzędek"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Differential gene expression analysis with PCA, UMAP and pseudotime - inspection of separate datasets.

Removal of batch effect between two Drosophila datasets.

# libraries

```{r}
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(ggrepel)
library(princurve)
library(sva)
library(forcats)
```

# locusta

## differential expression analysis

```{r}
#load count matrix and metadata
locusta_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/locusta_counts.rds")
locusta_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/locusta_metadata.rds")

###DESeq
locusta_dds <- DESeqDataSetFromMatrix(countData = locusta_counts,
                              colData = locusta_metadata,
                              design= ~Stage)
#Run DESeq2 to estimate size factors, dispersions, and fit the model
locusta_dds <- DESeq(locusta_dds)

head(results(locusta_dds, tidy=TRUE))
summary(results(locusta_dds))
```

```{r}
#dont compare 1 with 1, 2 with 2, etc.
stages <- levels(locusta_dds$Stage)

#pairwise differential expression analysis between all unique stage pairs
locusta_comparisons <- list()
for (i in 1:(length(stages)-1)) {
  for (j in (i+1):length(stages)) {
    name <- paste0(stages[j], "_vs_", stages[i])
    locusta_comparisons[[name]] <- results(locusta_dds, contrast = c("Stage", stages[j], stages[i]))
  }
}
```

```{r}
#significantly differentially expressed genes (DEGs) for each comparison
locusta_deg <- lapply(locusta_comparisons, function(results) subset(results, padj < 0.05))

#gene IDs of DEGs for each comparison
locusta_genes <- lapply(locusta_deg, rownames)

#vector of all unique DEGs across all comparisons
locusta_genes_uniq <- unique(unlist(locusta_genes))
```

## pca

```{r}
#variance stabilizing transformation for downstream visualization
locusta_vst <- vst(locusta_dds, blind = FALSE)

#pca with grouping by sample name
pcaData <- plotPCA(locusta_vst, intgroup = "Sample.name", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

pcaData$Stage <- colData(locusta_vst)$Stage

#plot pca, color by stage
p <- ggplot(pcaData, aes(PC1, PC2, color = Stage, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle(expression(bold("PCA of gene expression in ") * bolditalic("L. migratoria"))) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_locusta.png",
       p, width = 8, height = 6, dpi = 300)
```

## heatmap

```{r}
#redo stages (in case i run only this chunk and before other species)
stages <- levels(locusta_dds$Stage)

#creating matrix, filled with zeroes
locusta_matrix <- matrix(0, nrow = length(stages), ncol = length(stages))
rownames(locusta_matrix) <- stages
colnames(locusta_matrix) <- stages

#fill matrix with deg counts
for (name in names(locusta_deg)) {
  #for each pair of stages
  degs <- locusta_deg[[name]]
  comps <- unlist(strsplit(name, "_vs_"))
  group1 <- comps[1]
  group2 <- comps[2]
  
  n_deg <- sum(degs$padj < 0.05, na.rm = TRUE)
  
  #writing in matrix, in both orders
  locusta_matrix[group1, group2] <- n_deg
  locusta_matrix[group2, group1] <- n_deg
}

deg_colors <- colorRampPalette(RColorBrewer::brewer.pal(n = 9, name = "Greens"))(100)

#heatmap of deg counts between stages
png("/home/cluster_user/Magisterka/plots/pheatmap_locusta.png",
    width = 2400, height = 1800, res = 300)

pheatmap(locusta_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "Number of DEGs between stages in L. migratoria",
         color = deg_colors)

dev.off()

pheatmap(locusta_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "Number of DEGs between stages in L. migratoria",
         color = deg_colors)
```

### pseudotime

```{r}
print(percentVar)

#from pca matrix, 2 main components
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:2)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
#standardise
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

#check if pseudotime is according to the stages order
stages_order <- c("Nymph 1", "Nymph 2", "Nymph 3", "Nymph 4", "Nymph 5")

pcaData$Stage <- factor(pcaData$Stage, levels = stages_order)
pcaData$Stage <- droplevels(pcaData$Stage)

#check correlation with the stages order and flip if needed
cor_test <- cor(as.numeric(factor(pcaData$Stage, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")
print(cor_test)
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:2)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(PC1, PC2, shape = Stage, color = Pseudotime, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 8)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = expression("Pseudotime on "* italic("L. migratoria") * " gene expression PCA"),
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_locusta.png",
        p_curve, width = 8, height = 6, dpi = 300)
```

# harmonia

## differential expression analysis

```{r}
harmonia_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/harmonia_counts.rds")
harmonia_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/harmonia_metadata.rds")

###DESeq

harmonia_dds <- DESeqDataSetFromMatrix(countData = harmonia_counts,
                              colData = harmonia_metadata,
                              design= ~Stage)
#estimate factors
harmonia_dds <- DESeq(harmonia_dds)

head(results(harmonia_dds, tidy=TRUE))
summary(results(harmonia_dds))
```

```{r}
stages <- levels(harmonia_dds$Stage)

harmonia_comparisons <- list()
for (i in 1:(length(stages)-1)) {
  for (j in (i+1):length(stages)) {
    name <- paste0(stages[j], "_vs_", stages[i])
    harmonia_comparisons[[name]] <- results(harmonia_dds, contrast = c("Stage", stages[j], stages[i]))
  }
}

#diferential genes
harmonia_deg <- lapply(harmonia_comparisons, function(results) subset(results, padj < 0.05))

#names of the deg
harmonia_genes <- lapply(harmonia_deg, rownames)


#all unique DEGs
harmonia_genes_uniq <- unique(unlist(harmonia_genes))
```

## pca

```{r}
harmonia_vst <- vst(harmonia_dds, blind = FALSE)

pcaData <- plotPCA(harmonia_vst, intgroup = "Sample.name", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

pcaData$Stage <- colData(harmonia_vst)$Stage


stages_order <- c(
  "Larva 1", "Larva 2", "Larva 3", "Larva 4",
  "Prepupa 4h", "Prepupa 8h", "Prepupa 12h",
  "Prepupa 16h", "Prepupa 20h",
  "Pupa d1", "Pupa d2", "Pupa d3", "Pupa d4"
)

pcaData$Stage <- factor(pcaData$Stage, levels = stages_order)


p <- ggplot(pcaData, aes(PC1, PC2, color = Stage, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle(expression(bold("PCA of gene expression in ") * bolditalic("H. axyridis"))) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_harmonia.png",
       p, width = 8, height = 6, dpi = 300)
```

## heatmap

```{r}
stages <- levels(harmonia_dds$Stage)

#creating matrix, filled with zeroes
harmonia_matrix <- matrix(0, nrow = length(stages), ncol = length(stages))
rownames(harmonia_matrix) <- stages
colnames(harmonia_matrix) <- stages

for (name in names(harmonia_deg)) {
  #for each pair of stages
  degs <- harmonia_deg[[name]]
  comps <- unlist(strsplit(name, "_vs_"))
  group1 <- comps[1]
  group2 <- comps[2]
  
  n_deg <- sum(degs$padj < 0.05, na.rm = TRUE)
  
  #writing in matrix
  harmonia_matrix[group1, group2] <- n_deg
  harmonia_matrix[group2, group1] <- n_deg
}

png("/home/cluster_user/Magisterka/plots/pheatmap_harmonia.png",
    width = 2400, height = 1800, res = 300)

pheatmap(harmonia_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "Number of DEGs between stages in H. axyridis",
         color = deg_colors)

dev.off()

pheatmap(harmonia_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "Number of DEGs between stages in H. axyridis",
         color = deg_colors)
```

### pseudotime

```{r}
print(percentVar)

#from pca matrix, 2 main components
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:2)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
#standardise
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

#check if pseudotime is according to the stages order
pcaData$Stage <- factor(pcaData$Stage, levels = stages_order)
pcaData$Stage <- droplevels(pcaData$Stage)

#check correlation with the stages order and flip if needed
cor_test <- cor(as.numeric(factor(pcaData$Stage, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")
print(cor_test)
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:2)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(PC1, PC2, shape = Stage, color = Pseudotime, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0:13)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = expression("Pseudotime on "* italic("H. axyridis") * " gene expression PCA"),
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_harmonia.png",
        p_curve, width = 8, height = 6, dpi = 300)
```

# plutella

## differential expression analysis

```{r}
plutella_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/plutella_counts.rds")
plutella_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/plutella_metadata.rds")

###DESeq
plutella_dds <- DESeqDataSetFromMatrix(countData = plutella_counts,
                              colData = plutella_metadata,
                              design= ~Stage)
#estimate factors
plutella_dds <- DESeq(plutella_dds)

```

```{r}
stages <- levels(plutella_dds$Stage)

plutella_comparisons <- list()
for (i in 1:(length(stages)-1)) {
  for (j in (i+1):length(stages)) {
    name <- paste0(stages[j], "_vs_", stages[i])
    plutella_comparisons[[name]] <- results(plutella_dds, contrast = c("Stage", stages[j], stages[i]))
  }
}

#diferential genes
plutella_deg <- lapply(plutella_comparisons, function(results) subset(results, padj < 0.05))

#names of the deg
plutella_genes <- lapply(plutella_deg, rownames)

#all unique DEGs
plutella_genes_uniq <- unique(unlist(plutella_genes))

```

## pca

```{r}
plutella_vst <- vst(plutella_dds, blind = FALSE)

pcaData <- plotPCA(plutella_vst, intgroup = "Sample.name", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

pcaData$Stage <- colData(plutella_vst)$Stage


stages_order <- c(
  "Larva 1", "Larva 2", "Larva 3", "Larva 4", "Prepupa 4h", "Prepupa 8h","Prepupa 12h","Prepupa 16h", "Pupa d1", "Pupa d2", "Pupa d3"
)

pcaData$Stage <- factor(pcaData$Stage, levels = stages_order)


p <- ggplot(pcaData, aes(PC1, PC2, color = Stage, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle(expression(bold("PCA of gene expression in ") * bolditalic("P. xylostella"))) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )


print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_plutella.png",
       p, width = 8, height = 6, dpi = 300)
```

## heatmap

```{r}
stages <- levels(plutella_dds$Stage)

#creating matrix, filled with zeroes
plutella_matrix <- matrix(0, nrow = length(stages), ncol = length(stages))
rownames(plutella_matrix) <- stages
colnames(plutella_matrix) <- stages

for (name in names(plutella_deg)) {
  #for each pair of stages
  degs <- plutella_deg[[name]]
  comps <- unlist(strsplit(name, "_vs_"))
  group1 <- comps[1]
  group2 <- comps[2]
  
  n_deg <- sum(degs$padj < 0.05, na.rm = TRUE)
  
  #writing in matrix
  plutella_matrix[group1, group2] <- n_deg
  plutella_matrix[group2, group1] <- n_deg
}

png("/home/cluster_user/Magisterka/plots/pheatmap_plutella.png",
    width = 2400, height = 1800, res = 300)

pheatmap(plutella_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "Number of DEGs between stages in P. xylostella",
         color = deg_colors)

dev.off()

pheatmap(plutella_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "Number of DEGs between stages in P. xylostella",
         color = deg_colors)
```

### pseudotime

```{r}
print(percentVar)

#from pca matrix, 2 main components
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:2)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
#standardise
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

#check if pseudotime is according to the stages order
pcaData$Stage <- factor(pcaData$Stage, levels = stages_order)
pcaData$Stage <- droplevels(pcaData$Stage)

#check correlation with the stages order and flip if needed
cor_test <- cor(as.numeric(factor(pcaData$Stage, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")
print(cor_test)
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:2)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(PC1, PC2, shape = Stage, color = Pseudotime, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0:13)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = expression("Pseudotime on "* italic("P. xylostella") * " gene expression PCA"),
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_plutella.png",
        p_curve, width = 8, height = 6, dpi = 300)
```

# drosophila (set I)

## differential expression analysis

```{r}
drosophila_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila_counts.rds")
drosophila_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila_metadata.rds")

###DESeq
drosophila_dds <- DESeqDataSetFromMatrix(countData = drosophila_counts,
                              colData = drosophila_metadata,
                              design= ~Stage)
#estimate factors
drosophila_dds <- DESeq(drosophila_dds)
```

```{r}
stages <- levels(drosophila_dds$Stage)

drosophila_comparisons <- list()
for (i in 1:(length(stages)-1)) {
  for (j in (i+1):length(stages)) {
    name <- paste0(stages[j], "_vs_", stages[i])
    drosophila_comparisons[[name]] <- results(drosophila_dds, contrast = c("Stage", stages[j], stages[i]))
  }
}

#diferential genes
drosophila_deg <- lapply(drosophila_comparisons, function(results) subset(results, padj < 0.05))

#names of the deg
drosophila_genes <- lapply(drosophila_deg, rownames)

#all unique DEGs
drosophila_genes_uniq <- unique(unlist(drosophila_genes))

```

## pca

```{r}
drosophila_vst <- vst(drosophila_dds, blind = FALSE)

pcaData <- plotPCA(drosophila_vst, intgroup = "Sample.name", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

pcaData$Stage <- colData(drosophila_vst)$Stage


stages_order <- c(
  "Larva 1", "Larva 2", "Larva 3","Larva 3 clear guts","White prepupa","White prepupa +12h","White prepupa +24h","Pupa WPP+2d","Pupa WPP+3d","Pupa WPP+4d"
)

pcaData$Stage <- factor(pcaData$Stage, levels = stages_order)

p <- ggplot(pcaData, aes(PC1, PC2, color = Stage, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle(expression(bold("    PCA of gene expression in ") * bolditalic("D. melanogaster")* bold(" (set 1)"))) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_drosophila.png",
       p, width = 8, height = 6, dpi = 300)
```

## heatmap

```{r}
stages <- levels(drosophila_dds$Stage)

#creating matrix, filled with zeroes
drosophila_matrix <- matrix(0, nrow = length(stages), ncol = length(stages))
rownames(drosophila_matrix) <- stages
colnames(drosophila_matrix) <- stages

for (name in names(drosophila_deg)) {
  #for each pair of stages
  degs <- drosophila_deg[[name]]
  comps <- unlist(strsplit(name, "_vs_"))
  group1 <- comps[1]
  group2 <- comps[2]
  
  n_deg <- sum(degs$padj < 0.05, na.rm = TRUE)
  
  #writing in matrix
  drosophila_matrix[group1, group2] <- n_deg
  drosophila_matrix[group2, group1] <- n_deg
}

png("/home/cluster_user/Magisterka/plots/pheatmap_drosophila.png",
    width = 2400, height = 1800, res = 300)

pheatmap(drosophila_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "     Number of DEGs between stages in D. melanogaster (set 1)",
         color = deg_colors)

dev.off()

pheatmap(drosophila_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "     Number of DEGs between stages in D. melanogaster (set 1)",
         color = deg_colors)
```

### pseudotime

```{r}
print(percentVar)

#from pca matrix, 2 main components
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:2)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
#standardise
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

#check if pseudotime is according to the stages order
pcaData$Stage <- factor(pcaData$Stage, levels = stages_order)
pcaData$Stage <- droplevels(pcaData$Stage)

#check correlation with the stages order and flip if needed
cor_test <- cor(as.numeric(factor(pcaData$Stage, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")
print(cor_test)
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:2)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(PC1, PC2, shape = Stage, color = Pseudotime, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0:13)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = expression("Pseudotime on "* italic("D. melanogaster") * " gene expression PCA"),
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_drosophila.png",
        p_curve, width = 8, height = 6, dpi = 300)
```

# blattella

## differential expression analysis

```{r}
blattella_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/blattella_counts.rds")
blattella_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/blattella_metadata.rds")

###DESeq
blattella_dds <- DESeqDataSetFromMatrix(countData = blattella_counts,
                              colData = blattella_metadata,
                              design= ~Stage)
#estimate factors
blattella_dds <- DESeq(blattella_dds)
```

```{r}
stages <- levels(blattella_dds$Stage)

blattella_comparisons <- list()
for (i in 1:(length(stages)-1)) {
  for (j in (i+1):length(stages)) {
    name <- paste0(stages[j], "_vs_", stages[i])
    blattella_comparisons[[name]] <- results(blattella_dds, contrast = c("Stage", stages[j], stages[i]))
  }
}

#diferential genes
blattella_deg <- lapply(blattella_comparisons, function(results) subset(results, padj < 0.05))

#names of the deg
blattella_genes <- lapply(blattella_deg, rownames)

#all unique DEGs
blattella_genes_uniq <- unique(unlist(blattella_genes))

print(stages)
```

## pca

```{r}
blattella_vst <- vst(blattella_dds, blind = FALSE)

pcaData <- plotPCA(blattella_vst, intgroup = "Sample.name", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

pcaData$Stage <- colData(blattella_vst)$Stage


stages_order <- c(
  "Non-fec. egg", "Embryo d0", "Embryo d1","Embryo d2","Embryo d6","Embryo d13", "Nymph 1", "Nymph 3", "Nymph 5", "Nymph 6","Adult female"
)

pcaData$Stage <- factor(pcaData$Stage, levels = stages_order)

p <- ggplot(pcaData, aes(PC1, PC2, color = Stage, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle(expression(bold("PCA of gene expression in ") * bolditalic("B. germanica"))) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_blattella.png",
       p, width = 8, height = 6, dpi = 300)
```

## heatmap

```{r}
stages <- levels(blattella_dds$Stage)

#creating matrix, filled with zeroes
blattella_matrix <- matrix(0, nrow = length(stages), ncol = length(stages))
rownames(blattella_matrix) <- stages
colnames(blattella_matrix) <- stages

for (name in names(blattella_deg)) {
  #for each pair of stages
  degs <- blattella_deg[[name]]
  comps <- unlist(strsplit(name, "_vs_"))
  group1 <- comps[1]
  group2 <- comps[2]
  
  n_deg <- sum(degs$padj < 0.05, na.rm = TRUE)
  
  #writing in matrix
  blattella_matrix[group1, group2] <- n_deg
  blattella_matrix[group2, group1] <- n_deg
}

png("/home/cluster_user/Magisterka/plots/pheatmap_blattella.png",
    width = 2400, height = 1800, res = 300)

pheatmap(blattella_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "Number of DEGs between stages in B. germanica",
         color = deg_colors)

dev.off()

pheatmap(blattella_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "Number of DEGs between stages in B. germanica",
         color = deg_colors)
```

### pseudotime

```{r}
print(percentVar)

#from pca matrix, 2 main components
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:2)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
#standardise
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

#check if pseudotime is according to the stages order
pcaData$Stage <- factor(pcaData$Stage, levels = stages_order)
pcaData$Stage <- droplevels(pcaData$Stage)

#check correlation with the stages order and flip if needed
cor_test <- cor(as.numeric(factor(pcaData$Stage, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")
print(cor_test)
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:2)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(PC1, PC2, shape = Stage, color = Pseudotime, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0:13)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = expression("Pseudotime on "* italic("B. germanica") * " gene expression PCA"),
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_blattella.png",
        p_curve, width = 8, height = 6, dpi = 300)
```

# drosophila (set II)

## differential expression analysis

```{r}
drosophila2_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila2_counts.rds")
drosophila2_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila2_metadata.rds")

###DESeq

drosophila2_dds <- DESeqDataSetFromMatrix(countData = drosophila2_counts,
                              colData = drosophila2_metadata,
                              design= ~Stage)
#estimate factors
drosophila2_dds <- DESeq(drosophila2_dds)
```

```{r}
stages <- levels(drosophila2_dds$Stage)

drosophila2_comparisons <- list()
for (i in 1:(length(stages)-1)) {
  for (j in (i+1):length(stages)) {
    name <- paste0(stages[j], "_vs_", stages[i])
    drosophila2_comparisons[[name]] <- results(drosophila2_dds, contrast = c("Stage", stages[j], stages[i]))
  }
}

#diferential genes
drosophila2_deg <- lapply(drosophila2_comparisons, function(results) subset(results, padj < 0.05))

#names of the deg
drosophila2_genes <- lapply(drosophila2_deg, rownames)

#all unique DEGs
drosophila2_genes_uniq <- unique(unlist(drosophila2_genes))

print(stages)
```

## pca

```{r}
drosophila2_vst <- vst(drosophila2_dds, blind = FALSE)

pcaData <- plotPCA(drosophila2_vst, intgroup = "Sample.name", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

pcaData$Stage <- colData(drosophila2_vst)$Stage


stages_order <- c(
  "Embryo 0-4h","Embryo 4-8h","Embryo 8-12h","Embryo 12-16h","Embryo 16-20h","Embryo 20-24h","Larva 1","Larva 2","Larva 3","Pupa","Adult Female","Adult Male"
)

pcaData$Stage <- factor(pcaData$Stage, levels = stages_order)

p <- ggplot(pcaData, aes(PC1, PC2, color = Stage, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle(expression(bold("PCA of gene expression in ") * bolditalic("D. melanogaster") * bold(" (set 2)"))) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_drosophila2.png",
       p, width = 8, height = 6, dpi = 300)
```

## heatmap

```{r}
deg_colors <- colorRampPalette(RColorBrewer::brewer.pal(n = 9, name = "Greens"))(100)
stages <- levels(drosophila2_dds$Stage)

#creating matrix, filled with zeroes
drosophila2_matrix <- matrix(0, nrow = length(stages), ncol = length(stages))
rownames(drosophila2_matrix) <- stages
colnames(drosophila2_matrix) <- stages

for (name in names(drosophila2_deg)) {
  #for each pair of stages
  degs <- drosophila2_deg[[name]]
  comps <- unlist(strsplit(name, "_vs_"))
  group1 <- comps[1]
  group2 <- comps[2]
  
  n_deg <- sum(degs$padj < 0.05, na.rm = TRUE)
  
  #writing in matrix
  drosophila2_matrix[group1, group2] <- n_deg
  drosophila2_matrix[group2, group1] <- n_deg
}

png("/home/cluster_user/Magisterka/plots/pheatmap_drosophila2.png",
    width = 2400, height = 1800, res = 300)

pheatmap(drosophila2_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "       Number of DEGs between stages in D. melanogaster (set 2)",
         color = deg_colors)

dev.off()

pheatmap(drosophila2_matrix,
         fontsize_row = 12,
         fontsize_col = 12,
         angle_col = 45,       
         cellwidth = 30,       
         cellheight = 20,
         main = "       Number of DEGs between stages in D. melanogaster (set 2)",
         color = deg_colors)
```

### pseudotime

```{r}
print(percentVar)

#from pca matrix, 2 main components
pc_mat_all <- as.matrix(pcaData[, paste0("PC", 1:2)])
rownames(pc_mat_all) <- pcaData$Sample2

#developmental curve
pc_curve_all <- principal_curve(pc_mat_all)

#pseudotime
lambda_all <- pc_curve_all$lambda
#standardise
lambda_norm_all <- (lambda_all - min(lambda_all)) / (max(lambda_all) - min(lambda_all))

pcaData$Pseudotime <- lambda_norm_all

#check if pseudotime is according to the stages order
pcaData$Stage <- factor(pcaData$Stage, levels = stages_order)
pcaData$Stage <- droplevels(pcaData$Stage)

#check correlation with the stages order and flip if needed
cor_test <- cor(as.numeric(factor(pcaData$Stage, levels = stages_order)), pcaData$Pseudotime, use = "complete.obs")
print(cor_test)
if (cor_test < 0) {
  pcaData$Pseudotime <- 1 - pcaData$Pseudotime
}

curve_line <- as.data.frame(pc_curve_all$s)
colnames(curve_line) <- paste0("PC", 1:2)
#sort point
curve_line <- curve_line[order(pcaData$Pseudotime), ]

#pca plot with trajectory curve
p_curve <- ggplot(pcaData, aes(PC1, PC2, shape = Stage, color = Pseudotime, label = Sample.name)) +
  geom_point(size = 2.5) +
  geom_text_repel(size = 3.5, max.overlaps = 1000) +
  geom_path(data = curve_line, aes(x = PC1, y = PC2), inherit.aes = FALSE, 
            color = "black", size = 1.2, alpha = 0.7) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0:13)) +
  scale_color_viridis_c() +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  ) +
  labs(
    title = expression("            Pseudotime on "* italic("D. melanogaster") * " (dataset 2) gene expression PCA"),
    color = "Pseudotime",
    shape = "Stage"
  )

print(p_curve)

ggsave("/home/cluster_user/Magisterka/plots/pseudotime_pca_drosophila2.png",
        p_curve, width = 8, height = 6, dpi = 300)
```

# joined drosophila

```{r}
# vst
vst1 <- vst(drosophila_dds, blind = FALSE)
vst2 <- vst(drosophila2_dds, blind = FALSE)

mat1 <- assay(vst1)
mat2 <- assay(vst2)

# common genes set
common_genes <- intersect(rownames(mat1), rownames(mat2))

mat1 <- mat1[common_genes, ]
mat2 <- mat2[common_genes, ]

# combine columns
mat_combined <- cbind(mat1, mat2)

# Metadata
meta1 <- as.data.frame(colData(vst1))
meta2 <- as.data.frame(colData(vst2))

meta1$Dataset <- "Set 1"
meta2$Dataset <- "Set 2"

meta <- rbind(meta1, meta2)

# PCA
pca <- prcomp(t(mat_combined), scale. = TRUE)
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)))

pca_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  Stage = meta$Stage,
  Dataset = meta$Dataset,
  Sample = meta$Sample.name
)

pca_df$Stage <- as.character(pca_df$Stage) 
pca_df$Stage <- fct_relevel(pca_df$Stage, sort(unique(pca_df$Stage))) 

# colors in the same order
stage_colors <- setNames(scales::hue_pal()(length(levels(pca_df$Stage))),
                         levels(pca_df$Stage))

p <- ggplot(pca_df, aes(PC1, PC2, color = Stage, shape = Dataset, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3.3, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle(expression(bold("Joint PCA in ") * bolditalic("D. melanogaster"))) +
  theme_minimal() +
  scale_color_manual(values = stage_colors) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  guides(
    color = guide_legend(order = 1),
    shape = guide_legend(order = 2)
  )

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_drosophila_joint.png",
       p, width = 8, height = 6, dpi = 300)

```

## remove batch effect

```{r}
# corrected counts only for PCA, UMAP, clustering (after VST on the counts).
# For DE testing and counting DEGs dont use ComBat-Seq counts -> (raw counts in DESeq2)

#intersect of common genes
common_genes <- intersect(rownames(drosophila_counts), rownames(drosophila2_counts))
drosophila_counts <- drosophila_counts[common_genes, ]
drosophila2_counts <- drosophila2_counts[common_genes, ]

combined_counts <- cbind(drosophila_counts, drosophila2_counts)
combined_meta <- rbind(drosophila_metadata, drosophila2_metadata)

combined_meta$dataset <- factor(c(rep("Set1", ncol(drosophila_counts)),
                                rep("Set2", ncol(drosophila2_counts))))
#batch reduction
combat_counts <- ComBat_seq(
  as.matrix(combined_counts),
  batch = combined_meta$dataset,
  group = combined_meta$Stage
)

#save
saveRDS(combat_counts,
        "/home/cluster_user/link_to_datadisk/copy/counts/drosophila_joined_counts.rds")
saveRDS(combined_meta,
        "/home/cluster_user/link_to_datadisk/copy/counts/drosophila_joined_metadata.rds")

#check the effects of batch removal
dds <- DESeqDataSetFromMatrix(
  countData = combat_counts,
  colData = combined_meta,
  design = ~ Stage
)

vst_mat <- vst(dds, blind = TRUE) |> assay()

pca <- prcomp(t(vst_mat), scale. = TRUE)
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)))

pca_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  Stage = combined_meta$Stage,
  Dataset = combined_meta$dataset,
  Sample = combined_meta$Sample.name
)

p_corr <- ggplot(pca_df, 
                 aes(PC1, PC2, color = Stage, shape = Dataset, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3.3, max.overlaps = 1000) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle(expression(bold("Batch-corrected PCA in ") * bolditalic("D. melanogaster"))) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13, face = "bold")
  )+
  guides(
    color = guide_legend(order = 1),
    shape = guide_legend(order = 2)
  )

print(p_corr)

ggsave("/home/cluster_user/Magisterka/plots/PCA_drosophila_joint_batch_corrected.png",
       p_corr, width = 8, height = 6, dpi = 300)
```

# triatoma

## differential expression analysis

```{r}
#| eval: false
#| include: false
triatoma_counts <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/triatoma_counts.rds")
triatoma_metadata <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/triatoma_metadata.rds")

###DESeq

triatoma_dds <- DESeqDataSetFromMatrix(countData = triatoma_counts,
                              colData = triatoma_metadata,
                              design= ~Stage)
#estimate factors
triatoma_dds <- DESeq(triatoma_dds)

head(results(triatoma_dds, tidy=TRUE))
summary(results(triatoma_dds))
```

```{r}
#| eval: false
#| include: false
stages <- levels(triatoma_dds$Stage)

triatoma_comparisons <- list()
for (i in 1:(length(stages)-1)) {
  for (j in (i+1):length(stages)) {
    name <- paste0(stages[j], "_vs_", stages[i])
    triatoma_comparisons[[name]] <- results(triatoma_dds, contrast = c("Stage", stages[j], stages[i]))
  }
}
```

```{r}
#| eval: false
#| include: false
#diferential genes
triatoma_deg <- lapply(triatoma_comparisons, function(results) subset(results, padj < 0.05))

#names of the deg
triatoma_genes <- lapply(triatoma_deg, rownames)

#all unique DEGs
triatoma_genes_uniq <- unique(unlist(triatoma_genes))
```

## pca

```{r}
#| eval: false
#| include: false
triatoma_vst <- vst(triatoma_dds, blind = FALSE)

pcaData <- plotPCA(triatoma_vst, intgroup = "Sample.name", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

pcaData$Stage <- colData(triatoma_vst)$Stage

p <- ggplot(pcaData, aes(PC1, PC2, color = Stage, label = Sample.name)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 20) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of gene expression in Triatoma") +
  theme_minimal() +
  theme(legend.position = "right")

print(p)

ggsave("/home/cluster_user/Magisterka/plots/PCA_triatoma.png",
       p, width = 8, height = 6, dpi = 300)
```

# save deg

```{r}
saveRDS(locusta_deg, file = "/home/cluster_user/link_to_datadisk/copy/counts/locusta_deg.rds")
saveRDS(harmonia_deg, file = "/home/cluster_user/link_to_datadisk/copy/counts/harmonia_deg.rds")
saveRDS(drosophila_deg, file = "/home/cluster_user/link_to_datadisk/copy/counts/drosophila_deg.rds")
saveRDS(plutella_deg, file = "/home/cluster_user/link_to_datadisk/copy/counts/plutella_deg.rds")
saveRDS(drosophila2_deg, file = "/home/cluster_user/link_to_datadisk/copy/counts/drosophila2_deg.rds")
saveRDS(blattella_deg, file = "/home/cluster_user/link_to_datadisk/copy/counts/blattella_deg.rds")
```
