---
title: "7. DEGs analysis"
format: html
editor: visual
---

this notebook includes tested approaches to DEGs analysis, none of them ended up providing notable results

# libraries

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(Biostrings)
library(tibble)
library(DESeq2)
library(UpSetR)
library(pheatmap)
```

# unique genes comparison

## load data

load DEGs data

```{r}
#read degs
locusta_deg <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/locusta_deg.rds")
harmonia_deg <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/harmonia_deg.rds")
plutella_deg <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/plutella_deg.rds")
drosophila_deg <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila_deg.rds")
drosophila2_deg <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/drosophila2_deg.rds")
blattella_deg <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/blattella_deg.rds")
```

load orthogroups and create matrix

```{r}
#read orthogroups and 1:1 orthogroups IDs
orthogroups <- read_tsv("/home/cluster_user/Magisterka/proteins/OrthoFinder/Results_Jun25/Orthogroups/Orthogroups.tsv")

single_copy <- read_lines("/home/cluster_user/Magisterka/proteins/OrthoFinder/Results_Jun25/Orthogroups/Orthogroups_SingleCopyOrthologues.txt")

#filter orthogroups to 1:1 only
og_sc <- orthogroups %>%
  filter(Orthogroup %in% single_copy)

head(og_sc)

#read snap orthogroups
snap_files <- list.files(
  "/home/cluster_user/Magisterka/proteins/OrthoFinder/Results_Jun25/output_snap_ogs/",
  pattern = "\\.fa$",
  full.names = TRUE
)

#extract IDs and species
snap_orthologs <- lapply(snap_files, function(file) {
  fasta <- readAAStringSet(file)
  genes <- names(fasta)
  
  tibble(
    orthogroup = str_remove(basename(file), "\\.fa.orthosnap.0.fa$"),
    gene_id = str_replace(genes, "^.*?_", "") %>% str_replace("\\..*", "")
  ) %>%
    mutate(
      Species = case_when(
        str_detect(gene_id, "^Haxy") ~ "harmonia",
        str_detect(gene_id, "^Bger") ~ "blattella",
        str_detect(gene_id, "^Pxyl") ~ "plutella",
        str_detect(gene_id, "^Dmen") ~ "drosophila",
        str_detect(gene_id, "^g") ~ "locusta"
      )
    )
}) %>% bind_rows()

#orthogroup genes in each species
snap_wide <- snap_orthologs %>%
  group_by(orthogroup, Species) %>%
  summarise(Gene = paste(unique(gene_id), collapse = ","), .groups="drop") %>%
  pivot_wider(names_from = Species, values_from = Gene)

head(snap_wide)

names(snap_wide) <- str_to_lower(names(snap_wide))
names(og_sc) <- str_to_lower(names(orthogroups))

#merge 1:1 standard with SNAP
merged <- bind_rows(og_sc, snap_wide)
head(merged)

#orthogroup species gene
og_long <- merged %>%
  pivot_longer(-orthogroup, names_to="Species", values_to="Gene") %>%
  separate_rows(Gene, sep=",") %>%
  filter(Gene != "")

head(og_long)

saveRDS(og_long, file = "/home/cluster_user/link_to_datadisk/copy/counts/gene2og.rds")
```

```{r}
#list of lists of DEGs
deg_lists_per_species <- list(
  plutella = plutella_deg,
  locusta = locusta_deg,
  blattella = blattella_deg,
  drosophila_1 = drosophila_deg,
  drosophila_2 = drosophila2_deg,
  harmonia = harmonia_deg
)

#function that maps DEGs to matching orthogroups
map_deg_to_og <- function(deg_list, species_name, og_long) {
  #"drosophila_1", "drosophila_2" mapped to "drosophila"
  species_lookup <- case_when(
    species_name %in% c("drosophila_1", "drosophila_2") ~ "drosophila",
    TRUE ~ species_name
  )
  
  gene2og <- og_long %>%
    filter(Species == species_lookup) %>%
    select(Gene, orthogroup)
  
  deg_og <- lapply(deg_list, function(res) {
    genes <- rownames(res)
    inner_join(tibble(Gene = genes), gene2og, by = "Gene") %>%
      pull(orthogroup) %>% unique()
  })
  
  names(deg_og) <- names(deg_list)
  return(deg_og)
}

#list of mapped DEG orthogroups per stage per species
deg_og_per_species <- lapply(names(deg_lists_per_species), function(sp) {
  map_deg_to_og(deg_lists_per_species[[sp]], sp, og_long)
})
names(deg_og_per_species) <- names(deg_lists_per_species)
```

```{r}
### change specific stage names to the common stage names (Stage2)
#based on metadata
metadata <- read.csv(file="/home/cluster_user/Magisterka/metadata2.csv", row.names = 1, header = TRUE, sep = ";") 

species_alias <- c(
  "drosophila_1" = "drosophila",
  "drosophila_2" = "drosophila"
)

#rename comparisons and remove duplicates
rename_comparison_from_metadata <- function(comp, sp, metadata, species_alias) {
  phases <- unlist(strsplit(comp, "_vs_"))
  
  #normalize names using aliases
  sp_clean <- ifelse(sp %in% names(species_alias), species_alias[sp], sp)
  
  #mapping from Stage to Stage2
  stage_map <- setNames(metadata$Stage2[metadata$Species == sp_clean],
                        metadata$Stage[metadata$Species == sp_clean])
  
  new_phases <- unname(stage_map[phases])
  if (any(is.na(new_phases))) return(NA)
  if (length(unique(new_phases)) == 1) return(NA) #delete self-comparison
  
  paste(new_phases, collapse = "_vs_")
}

deg_og_stage2 <- lapply(names(deg_og_per_species), function(sp) {
  comps <- names(deg_og_per_species[[sp]])
  
  renamed <- sapply(comps, function(c) rename_comparison_from_metadata(c, sp, metadata, species_alias),
                    USE.NAMES = FALSE)
  
  valid_idx <- !is.na(renamed) & renamed != "" & !duplicated(renamed)
  
  out <- deg_og_per_species[[sp]][valid_idx]
  names(out) <- renamed[valid_idx]
  out
})
names(deg_og_stage2) <- names(deg_og_per_species)

```

## matrix

Compare differentially expressed orthogroups across species

```{r}
# Exclude species with no significant orthogroups
deg_og_stage2 <- deg_og_stage2[
  sapply(deg_og_stage2, length) > 0
]

### Generate binary presence–absence matrices
# Create a list of all unique orthogroups
all_ogs <- sort(
  unique(
    unlist(
      lapply(deg_og_stage2, unlist)
    )
  )
)

# Create a binary matrix for each species indicating
# the presence or absence of each orthogroup
bin_mat_list <- lapply(deg_og_stage2, function(deg_og) {

  comparisons <- names(deg_og)

  mat <- sapply(comparisons, function(cmp) {

    ogs <- deg_og[[cmp]]

    as.integer(all_ogs %in% ogs)
  })

  rownames(mat) <- all_ogs
  colnames(mat) <- comparisons

  mat
})

### Standardise comparison names

# Convert reciprocal comparisons to a common name
normalize_comparison_name <- function(name) {

  stages <- strsplit(name, "_vs_")[[1]]

  paste(sort(stages), collapse = "_vs_")
}

bin_mat_list_stage2 <- lapply(bin_mat_list, function(mat) {

  colnames(mat) <- sapply(
    colnames(mat),
    normalize_comparison_name
  )

  mat[, !duplicated(colnames(mat)), drop = FALSE]
})

### Combine species-specific matrices

all_mat <- do.call(
  cbind,
  lapply(names(bin_mat_list_stage2), function(sp) {

    mat <- bin_mat_list_stage2[[sp]]

    colnames(mat) <- paste(
      sp,
      colnames(mat),
      sep = "::"
    )

    mat
  })
)
```

## heatmap

```{r}
### Plot binary presence–absence heatmap

# Retain orthogroups present in at least two comparisons
og_filter <- rowSums(all_mat) > 1
mat_plot <- all_mat[og_filter, , drop = FALSE]

png(
  "/home/cluster_user/Magisterka/plots/heatmap_DEG_OGs.png",
  width = 1500,
  height = 2000,
  res = 300
)

pheatmap(
  mat_plot,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  fontsize_col = 6,
  color = c("white", "red"),
  legend = FALSE
)

dev.off()

# Display the heatmap in the R session
pheatmap(
  mat_plot,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  fontsize_col = 6,
  color = c("white", "red"),
  legend = FALSE
)
```

```{r}
## Visualise orthogroup overlap using an UpSet plot

### Convert the binary matrix to a list of orthogroup sets

# Retain orthogroups present in at least two comparisons
og_filter <- rowSums(all_mat) > 1

# Create a list of orthogroup sets for each comparison
sets_list_top <- lapply(colnames(all_mat), function(cmp) {
  rownames(all_mat)[
    all_mat[, cmp] == 1 &
      rownames(all_mat) %in% rownames(all_mat[og_filter, ])
  ]
})

names(sets_list_top) <- colnames(all_mat)

### Save the UpSet plot

png(
  "/home/cluster_user/Magisterka/plots/upset_plot_topOG.png",
  width = 2000,
  height = 2000,
  res = 200
)

upset(
  fromList(sets_list_top),
  nsets = 20,
  nintersects = 30,
  order.by = "freq",
  sets.bar.color = "#4C72B0",
  main.bar.color = "#DD8452",
  text.scale = c(1.5, 1.2, 1, 1, 1, 1),
  mainbar.y.label = "Number of DEG Orthogroups"
)

dev.off()

### Display the UpSet plot

upset(
  fromList(sets_list_top),
  nsets = 20,
  nintersects = 30,
  order.by = "freq",
  sets.bar.color = "#4C72B0",
  main.bar.color = "#DD8452",
  text.scale = c(1.5, 1.2, 1, 1, 1, 1),
  mainbar.y.label = "Number of DEG Orthogroups"
)
```

# trying some other analysis

## comparison to nymphs

Compare nymph-specific orthogroups with holometabolous comparisons

```{r}
### Identify comparisons

# Extract species names from comparison labels
comp_species <- sapply(strsplit(colnames(all_mat), "::"), `[`, 1)

# Separate hemimetabolous and holometabolous comparisons
hemi_comps <- colnames(all_mat)[comp_species %in% hemi_species]
holo_comps <- colnames(all_mat)[comp_species %in% holo_species]

# Select comparisons involving the nymphal stage
nymph_comps <- grep("Nymph", hemi_comps, value = TRUE)

### Calculate orthogroup overlap
# Collect all orthogroups identified in nymph comparisons
nymph_ogs <- unique(unlist(lapply(nymph_comps, function(cmp) {
  rownames(all_mat)[all_mat[, cmp] == 1]
})))

```

```{r}
# Count orthogroups shared with each holometabolous comparison
holo_overlap <- sapply(holo_comps, function(cmp) {
  ogs <- rownames(all_mat)[all_mat[, cmp] == 1]
  length(intersect(nymph_ogs, ogs))
})

sort(holo_overlap, decreasing = TRUE)
```

```{r}
### Visualise orthogroup overlap

barplot(
  sort(holo_overlap, decreasing = TRUE),
  las = 2,
  cex.names = 0.7,
  col = "#4C72B0",
  ylab = "Shared orthogroups with hemimetabolous nymphs",
  main = "Overlap between nymph and holometabolous comparisons"
)
```

## trends of gene expression

```{r}
#load DEGs data
dds <- readRDS("/home/cluster_user/link_to_datadisk/copy/counts/dds.rds")
```

```{r}
## Identify stage-specific and gradual gene expression changes

### Larva vs nymph differential expression

res_larva <- results(dds, contrast = c("Stage2", "Larva", "Nymph"))

# Larva-upregulated genes (relative to nymph)
larva_genes <- rownames(res_larva)[
  res_larva$padj < 0.05 &
    res_larva$log2FoldChange > 1
]

### Prepare ordinal stage variable for trend analysis

# Convert developmental stages into numeric order
metadata$Stage_num <- as.numeric(factor(
  metadata$Stage2,
  levels = c(
    "Eggs",
    "Embryo",
    "Larva",
    "Nymph",
    "Prepupa",
    "Pupa",
    "Adult Male",
    "Adult Female"
  )
))

# Ensure metadata is aligned with count matrix
metadata <- metadata[colnames(combined_counts), ]

### Fit DESeq2 model for developmental trend analysis

dds_trend <- DESeqDataSetFromMatrix(
  countData = combined_counts,
  colData   = metadata,
  design    = ~ Species + Stage_num
)

dds_trend <- DESeq(dds_trend)
res_trend <- results(dds_trend, name = "Stage_num")

### Identify genes with gradual expression changes across development

gradual_genes <- rownames(res_trend)[
  !is.na(res_trend$padj) &
    res_trend$padj < 0.05 &
    abs(res_trend$log2FoldChange) > 0
]

# Split gradual changes into increasing and decreasing trends
gradual_up <- rownames(res_trend)[
  res_trend$padj < 0.05 &
    res_trend$log2FoldChange > 0
]

gradual_down <- rownames(res_trend)[
  res_trend$padj < 0.05 &
    res_trend$log2FoldChange < 0
]

### Additional pairwise stage comparisons

res_nymph <- results(dds, contrast = c("Stage2", "Nymph", "Larva"))

res_pupa <- results(dds, contrast = c("Stage2", "Prepupa", "Nymph"))

# Genes enriched in nymph relative to larva
nymph_genes <- rownames(res_nymph)[
  res_nymph$padj < 0.05 &
    res_nymph$log2FoldChange > 1
]

```
